### 2024-11-24 other/EDA.qmd
added many exploratory graphs

---
title: "EDA"
author: 
  - Robert Ford
format: pdf
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)
library(dplyr)
library(lubridate)
library(scales)
library(knitr)
library(kableExtra)

df <- read_csv(here::here("data/analysis_data/augment_data.csv"))

df <- df %>% select(-link)

```

```{r fig-template}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: template
#| fig-align: center

```

```{r fig-head}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: template
#| fig-align: center

kable(head(df, 5), caption = "Sample of the Data")
```

This graph shows how much was awarded by month
```{r fig-award-amount-month}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: template
#| fig-align: center

df <- df %>%
  mutate(year_month = floor_date(award_date, "month"))

# Aggregate the amount by 'year_month'
monthly_spending <- df %>%
  group_by(year_month) %>%
  summarise(total_amount = sum(amount, na.rm = TRUE))

# Plot the line graph with formatted y-axis labels
ggplot(monthly_spending, aes(x = year_month, y = total_amount)) +
  geom_line(color = "blue") +
  labs(title = "Amount Spent Each Month",
       x = "Time",
       y = "Amount Awarded") +
  scale_x_date(date_labels = "%Y-%m", date_breaks = "1 month") +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r fig-template4}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: template
#| fig-align: center

df$amount <- as.numeric(df$amount)

# Ensure 'award_date' is in Date format (if not already)
# df$award_date <- as.Date(df$award_date, format="%Y-%m-%d")  # Commented out since it's already a Date

# Identify the top 5 buyers by total amount awarded
top_buyers <- df %>%
  group_by(buyer) %>%
  summarise(total_amount = sum(amount, na.rm = TRUE)) %>%
  arrange(desc(total_amount)) %>%
  slice_head(n = 3) %>%
  pull(buyer)

# Filter the data for only the top 5 buyers
df_top_buyers <- df %>%
  filter(buyer %in% top_buyers)

# Create a 'year_month' column using lubridate
df_top_buyers <- df_top_buyers %>%
  mutate(year_month = floor_date(award_date, "month"))

# Aggregate the amount by 'year_month' and 'buyer'
monthly_spending <- df_top_buyers %>%
  group_by(year_month, buyer) %>%
  summarise(total_amount = sum(amount, na.rm = TRUE), .groups = 'drop')

# Plot the line graph
ggplot(monthly_spending, aes(x = year_month, y = total_amount, color = buyer, group = buyer)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(title = "Amount Spent Each Month by Top 5 Buyers",
       x = "Time",
       y = "Amount Awarded",
       color = "Buyer") +
  scale_x_date(date_labels = "%Y-%m", date_breaks = "1 month") +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r fig-template3}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: template
#| fig-align: center

df$amount <- as.numeric(df$amount)

# Summarize total amount spent by each buyer
buyer_totals <- df %>%
  group_by(buyer) %>%
  summarise(total_amount = sum(amount, na.rm = TRUE)) %>%
  arrange(desc(total_amount))

# Select the top ten buyers
top_ten_buyers <- buyer_totals %>%
  slice_head(n = 10)

# Plot the bar chart
ggplot(top_ten_buyers, aes(x = reorder(buyer, total_amount), y = total_amount)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +
  labs(title = "Top 10 Buyers by Total Amount Spent",
       x = "Buyer",
       y = "Total Amount Spent") +
  scale_y_continuous(labels = comma) +
  theme_minimal()

```

```{r fig-template2}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: template
#| fig-align: center

supplier_totals <- df %>%
  group_by(supplier) %>%
  summarise(total_amount = sum(amount, na.rm = TRUE)) %>%
  arrange(desc(total_amount))

# Select the top ten suppliers
top_suppliers <- supplier_totals %>%
  slice_head(n = 10)

# Wrap supplier names for better legibility
top_suppliers$supplier <- str_wrap(top_suppliers$supplier, width = 15)

# Plot the bar chart
ggplot(top_suppliers, aes(x = reorder(supplier, -total_amount), y = total_amount)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Amount Awarded to Top 10 Suppliers",
       x = "Supplier",
       y = "Total Amount Awarded") +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.margin = unit(c(1, 1, 1, 2), "cm")  # Adjust margins if necessary
  )

```

```{r fig-template1}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: template
#| fig-align: center

df <- df %>%
  mutate(contract_length_days = as.numeric(difftime(end_date, start_date, units = "days")))

# Group by buyer and calculate average length of contract and average contract size
buyer_stats <- df %>%
  group_by(buyer) %>%
  summarise(
    `Average Length of Contract (Days)` = mean(contract_length_days, na.rm = TRUE),
    `Average Contract Size` = mean(amount, na.rm = TRUE)
  )

# Reshape the data to have buyers as columns and metrics as rows
buyer_stats_long <- buyer_stats %>%
  pivot_longer(cols = -buyer, 
               names_to = "Metric", 
               values_to = "Value")

# Now pivot wider to have buyers as columns
buyer_stats_wide <- buyer_stats_long %>%
  pivot_wider(names_from = buyer, values_from = Value)

# Optionally, format the numbers
buyer_stats_wide <- buyer_stats_wide %>%
  mutate(across(-Metric, ~ round(.x, 2)))

# Display the table using knitr::kable
kable(buyer_stats_wide, caption = "Average Length of Contract and Average Contract Size by Buyer")

```

### 2024-11-24 scripts/04-test_analysis_data.R
cleaned up

@@ -6,7 +6,7 @@
 # License: MIT
 
 
-# TODO Check that all dates have no NA's, any missing means that data was lost in cleaning
+
 
 #### Workspace setup ####
 library(tidyverse)
@@ -49,23 +49,23 @@ test_that("No missing values in dataset", {
 
 # Test that there are no duplicate rows
 test_that("No duplicate rows", {
-  expect_equal(nrow(analysis_data), nrow(unique(analysis_data)))
+  expect_equal(nrow(data), nrow(unique(data)))
 })
 
 # Test that 'buyer' and 'supplier' are not empty strings
 test_that("'buyer' and 'supplier' are not empty", {
-  expect_true(all(nzchar(analysis_data$buyer)))
-  expect_true(all(nzchar(analysis_data$supplier)))
+  expect_true(all(nzchar(data$buyer)))
+  expect_true(all(nzchar(data$supplier)))
 })
 
 # Test that 'per_day' is greater than 0
 test_that("per_day > 0", {
-  expect_true(all(analysis_data$per_day > 0))
+  expect_true(all(data$per_day > 0))
 })
 
 # Test that date columns are valid dates
 test_that("Date columns are valid dates", {
-  expect_true(all(!is.na(as.Date(analysis_data$award_date))))
-  expect_true(all(!is.na(as.Date(analysis_data$start_date))))
-  expect_true(all(!is.na(as.Date(analysis_data$end_date))))
+  expect_true(all(!is.na(as.Date(data$award_date))))
+  expect_true(all(!is.na(as.Date(data$start_date))))
+  expect_true(all(!is.na(as.Date(data$end_date))))
 })

### 2024-11-24 scripts/04-test_analysis_data.R
added reasonable Date tests

@@ -69,3 +69,14 @@ test_that("Date columns are valid dates", {
   expect_true(all(!is.na(as.Date(data$start_date))))
   expect_true(all(!is.na(as.Date(data$end_date))))
 })
+
+# Test that 'start_date' and 'award_date' are not before 2015
+test_that("No 'start_date' or 'award_date' before 2015", {
+  expect_true(all(analysis_data$start_date >= as.Date("2015-01-01")))
+  expect_true(all(analysis_data$award_date >= as.Date("2015-01-01")))
+})
+
+# Test that 'end_date' is not beyond 2099
+test_that("No 'end_date' beyond 2099", {
+  expect_true(all(analysis_data$end_date <= as.Date("2099-12-31")))
+})

### 2024-11-24 scripts/03-data_cleaning.R
filtering out unreasonable Dates

@@ -61,6 +61,14 @@ cleaned <- cleaned[cleaned$duration_days > 0, ]
 
 cleaned <- cleaned[cleaned$amount > 0, ]
 
+# Filter out contracts with 'start_date' or 'award_date' before 2015
+cleaned <- cleaned[
+  cleaned$start_date >= as.Date("2015-01-01") &
+    cleaned$award_date >= as.Date("2015-01-01"), ]
+
+# Filter out contracts with 'end_date' beyond 2099
+cleaned <- cleaned[cleaned$end_date <= as.Date("2099-12-31"), ]
+
 # Remove rows with any NA values
 cleaned <- na.omit(cleaned)
 

### 2024-11-24 scripts/01-download_data.R
added TODO

@@ -6,6 +6,9 @@
 # License: MIT
 # Pre-requisites: Docker, Firefox Web driver
 # See appendix _ for more details
+# Data was collected on: 21 November 2024
+
+#TODO RE-RUN!
 
 #### Workspace setup ####
 library(RSelenium)
@@ -14,14 +17,14 @@ library(dplyr)
 library(readr)
 library(readr)
 library(dplyr)
-#docker run -d -p 4444:4444 -p 5900:5900 -v /Users/NAME/canadian_gov_procurement/data/raw_data:/home/seluser/Downloads selenium/standalone-firefox:3.141.59
+#docker run -d -p 4444:4444 -p 5900:5900 -v /Users/robert_ford/canadian_gov_procurement/data/raw_data:/home/seluser/Downloads selenium/standalone-firefox:3.141.59
 
 # Paths
 # Path inside the container
 download_dir <- "/home/seluser/Downloads"
 
 # Path on the host
-host_download_dir <- "/Users/NAME/canadian_gov_procurement/data/raw_data"
+host_download_dir <- "/Users/robert_ford/canadian_gov_procurement/data/raw_data"
 
 # Set up Firefox profile with download preferences
 fprof <- makeFirefoxProfile(list(

### 2024-11-24 other/EDA.qmd
added per day graph

@@
 library(kableExtra)
+library(zoo)
 
-df <- read_csv(here::here("data/analysis_data/augment_data.csv"))
+df <- read_csv(here::here("data/analysis_data/cleaned_data.csv"))
 
@@
 
-df$amount <- as.numeric(df$amount)
+start_date_min <- min(df$start_date)
+end_date_max <- as.Date("2030-12-31")
 
-# Ensure 'award_date' is in Date format (if not already)
-# df$award_date <- as.Date(df$award_date, format="%Y-%m-%d")  # Commented out since it's already a Date
+# Create events at start dates (add per_day amount)
+events_start <- data.frame(date = df$start_date, amount = df$per_day)
 
-# Identify the top 5 buyers by total amount awarded
-top_buyers <- df %>%
-  group_by(buyer) %>%
-  summarise(total_amount = sum(amount, na.rm = TRUE)) %>%
-  arrange(desc(total_amount)) %>%
-  slice_head(n = 3) %>%
-  pull(buyer)
+# Create events at end dates + 1 day (subtract per_day amount)
+events_end <- data.frame(date = df$end_date + 1, amount = -df$per_day)
 
-# Filter the data for only the top 5 buyers
-df_top_buyers <- df %>%
-  filter(buyer %in% top_buyers)
+# Combine start and end events
+events <- rbind(events_start, events_end)
 
-# Create a 'year_month' column using lubridate
-df_top_buyers <- df_top_buyers %>%
-  mutate(year_month = floor_date(award_date, "month"))
+# Aggregate amounts by date
+events <- aggregate(amount ~ date, data = events, sum)
 
-# Aggregate the amount by 'year_month' and 'buyer'
-monthly_spending <- df_top_buyers %>%
-  group_by(year_month, buyer) %>%
-  summarise(total_amount = sum(amount, na.rm = TRUE), .groups = 'drop')
+# Sort events by date
+events <- events[order(events$date), ]
 
-# Plot the line graph
-ggplot(monthly_spending, aes(x = year_month, y = total_amount, color = buyer, group = buyer)) +
-  geom_line(size = 1) +
-  geom_point(size = 2) +
-  labs(title = "Amount Spent Each Month by Top 5 Buyers",
-       x = "Time",
-       y = "Amount Awarded",
-       color = "Buyer") +
-  scale_x_date(date_labels = "%Y-%m", date_breaks = "1 month") +
+# Compute cumulative sum of amounts to get the total per_day amount active on each date
+events$cumulative_amount <- cumsum(events$amount)
+
+# Create a full sequence of dates from the earliest start date to the latest end date
+date_seq <- seq(start_date_min, end_date_max, by = "day")
+
+# Merge cumulative amounts with the full date sequence
+amounts <- data.frame(date = date_seq)
+amounts <- merge(amounts, events[, c("date", "cumulative_amount")], by = "date", all.x = TRUE)
+
+# Carry forward the last observed cumulative_amount to fill in missing dates
+amounts$cumulative_amount <- na.locf(amounts$cumulative_amount, na.rm = FALSE)
+amounts$cumulative_amount[is.na(amounts$cumulative_amount)] <- 0
+
+# Plot the line chart using ggplot2
+library(ggplot2)
+ggplot(amounts, aes(x = date, y = cumulative_amount)) +
+  geom_line(color = "blue") +
+  labs(title = "Total Amount Spent Over Time",
+       x = "Date",
+       y = "Total Amount Spent per Day") +
   scale_y_continuous(labels = comma) +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+  theme_minimal()
 

### 2024-11-24 other/EDA.qmd
added intro table

@@
 
-kable(head(df, 5), caption = "Sample of the Data")
+df_table <- df %>% select(-contract, -start_date, -end_date)
+
+kable(head(df_table, 5), caption = "Sample of the Data") %>%
+  kable_styling(font_size = 10, latex_options = c("scale_down"))
+
 ```
@@
 
-df <- df %>%
+df_fig <- df %>%
   mutate(year_month = floor_date(award_date, "month"))
@@
 # Aggregate the amount by 'year_month'
-monthly_spending <- df %>%
+monthly_spending <- df_fig %>%
   group_by(year_month) %>%
@@
 ```
+```{r}
+
+```
 
@@
 #| fig-align: center
-
-df$amount <- as.numeric(df$amount)
 
@@
 
-```{r fig-template1}
+```{r}
 #| message: false
@@
 #| warning: false
-#| fig-cap: template
+#| fig-cap: "Average Contract Length (Days) by Government Department (Top 10)"
 #| fig-align: center
 
-df <- df %>%
-  mutate(contract_length_days = as.numeric(difftime(end_date, start_date, units = "days")))
+# Load necessary libraries
+library(dplyr)
+library(ggplot2)
 
-# Group by buyer and calculate average length of contract and average contract size
-buyer_stats <- df %>%
+# Calculate average contract length per buyer (government department)
+avg_contract_length <- df %>%
+  mutate(contract_length_days = as.numeric(difftime(end_date, start_date, units = "days"))) %>%
   group_by(buyer) %>%
-  summarise(
-    `Average Length of Contract (Days)` = mean(contract_length_days, na.rm = TRUE),
-    `Average Contract Size` = mean(amount, na.rm = TRUE)
-  )
+  summarise(`Average Length of Contract (Days)` = mean(contract_length_days, na.rm = TRUE)) %>%
+  arrange(desc(`Average Length of Contract (Days)`)) %>%
+  slice_head(n = 10)  # Select top 10 departments
 
-# Reshape the data to have buyers as columns and metrics as rows
-buyer_stats_long <- buyer_stats %>%
-  pivot_longer(cols = -buyer, 
-               names_to = "Metric", 
-               values_to = "Value")
-
-# Now pivot wider to have buyers as columns
-buyer_stats_wide <- buyer_stats_long %>%
-  pivot_wider(names_from = buyer, values_from = Value)
-
-# Optionally, format the numbers
-buyer_stats_wide <- buyer_stats_wide %>%
-  mutate(across(-Metric, ~ round(.x, 2)))
-
-# Display the table using knitr::kable
-kable(buyer_stats_wide, caption = "Average Length of Contract and Average Contract Size by Buyer")
+# Create the bar chart
+ggplot(avg_contract_length, aes(x = reorder(buyer, `Average Length of Contract (Days)`), y = `Average Length of Contract (Days)`)) +
+  geom_bar(stat = "identity", fill = "steelblue") +
+  coord_flip() +
+  labs(x = "Government Department", y = "Average Contract Length (Days)", title = "Average Contract Length (Days) by Government Department (Top 10)") +
+  theme_minimal()
 
 ```
+
+```{r}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-cap: "Average Contract Size by Government Department (Top 10)"
+#| fig-align: center
+
+# Load necessary libraries
+library(dplyr)
+library(ggplot2)
+library(scales)  # For formatting currency
+
+# Calculate average contract size per buyer (government department)
+avg_contract_size <- df %>%
+  group_by(buyer) %>%
+  summarise(`Average Contract Size` = mean(amount, na.rm = TRUE)) %>%
+  arrange(desc(`Average Contract Size`)) %>%
+  slice_head(n = 10)  # Select top 10 departments
+
+# Create the bar chart
+ggplot(avg_contract_size, aes(x = reorder(buyer, `Average Contract Size`), y = `Average Contract Size`)) +
+  geom_bar(stat = "identity", fill = "darkgreen") +
+  coord_flip() +
+  scale_y_continuous(labels = dollar_format(prefix = "$")) +
+  labs(x = "Government Department", y = "Average Contract Size", title = "Average Contract Size by Government Department (Top 10)") +
+  theme_minimal()
+
+```

### 2024-11-24 scripts/01-download_data.R
updated data

@@ -5,8 +5,8 @@
 # Contact: robert.ford@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: Docker, Firefox Web driver
-# See appendix _ for more details
-# Data was collected on: 21 November 2024
+# See appendix TODO for more details
+# Data was collected on: 24 November 2024
 
 #TODO RE-RUN!
 
@@ -47,22 +47,22 @@ remDr <- remoteDriver(
 remDr$open()
 
 # Navigate to the website
-url <- "https://theijf.org/procurement/records?table=awards&region=Federal&status=AWARDED"
+url <- "https://theijf.org/procurement/records?table=awards&region=Federal&status=AWARDED&dir=asc&sort=min_date"
 remDr$navigate(url)
 # Wait for the page to load
 Sys.sleep(5)
 
 # Total number of pages to scrape
-total_pages <- 330
+total_pages <- 329
 
-for (page in 143:total_pages) {
+for (page in 1:total_pages) {
   print(page)
-  Sys.sleep(2)
+  Sys.sleep(3)
   # Click the "Download" button
   tryCatch({
     download_button <- remDr$findElement(using = "xpath", "//button[@aria-label='Download data']")
     download_button$clickElement()
-    Sys.sleep(3)  # Wait for the download to complete
+    Sys.sleep(5)  # Wait for the download to complete
   }, error = function(e) {
     message("Error on page ", page, ": ", e$message)
     next  # Skip to the next iteration

### 2024-11-24 scripts/03-data_cleaning.R
remove duplicates

@@ -72,6 +72,7 @@ cleaned <- cleaned[cleaned$end_date <= as.Date("2099-12-31"), ]
 # Remove rows with any NA values
 cleaned <- na.omit(cleaned)
 
+
 duplicate_rows <- cleaned[duplicated(cleaned) | duplicated(cleaned, fromLast = TRUE), ]
 
 View(duplicate_rows)
@@ -79,6 +80,8 @@ View(duplicate_rows)
 # View the dataframe with duplicate rows
 head(duplicate_rows)
 
+
+cleaned <- unique(cleaned)
 
 View(cleaned)
 

### 2024-11-24 scripts/05-exploratory_data_analysis.R
added lines to further explore dataset

@@ -1,5 +1,5 @@
 #### Preamble ####
-# Purpose: EDA, Augementing data by calculating per annum spending
+# Purpose: EDA
 # Author: Robert Ford
 # Date: 23 November 2024
 # Contact: robert.ford@mail.utoronto.ca
@@ -11,14 +11,34 @@ library(tidyverse)
 library(rstanarm)
 
 #### Read data ####
-cleaned <- read_csv("data/analysis_data/cleaned_data.csv")
+df <- read_csv("data/analysis_data/cleaned_data.csv")
 
-u_supps <- data.frame(unique(cleaned$supplier))
+u_supps <- data.frame(unique(df$supplier))
 
-u_buyers <- data.frame(unique(cleaned$buyer))
+u_buyers <- data.frame(unique(df$buyer))
 
-View(u_buyers)
+public_health <- df[df$buyer == "Public Health Agency of Canada", ]
 
+defence <- df[df$buyer == "National Defence", ]
 
-View(cleaned)
+# Ensure 'amount' is numeric and 'end_date' is in Date format
+df <- df %>%
+  mutate(
+    # Remove dollar signs and commas from 'amount' and convert to numeric
+    amount = as.numeric(gsub("[$,]", "", amount)),
+    # Convert 'end_date' to Date format (assuming format like "Aug. 21, 2024")
+    end_date = as.Date(end_date, format = "%b. %d, %Y")
+  )
 
+# Group by 'end_date' and calculate the number of contracts and sum of amounts
+contracts_expiring <- df %>%
+  group_by(end_date) %>%
+  summarise(
+    number_of_contracts_expiring = n(),
+    sum_of_amounts = sum(amount, na.rm = TRUE)
+  ) %>%
+  ungroup()
+
+View(contracts_expiring)
+View(public_health)
+View(defence)

### 2024-11-28 other/EDA.qmd
added more graphs

@@
 ```
+#TODO Color code by buyer
 
@@
   theme_minimal()
+```
+
+```{r}
+# Convert award_date to Date type and extract the year
+df$award_date <- as.Date(df$award_date, format = "%Y-%m-%d")
+df$year <- format(df$award_date, "%Y")
+
+# Filter for military spending (where buyer is "National Defence")
+military_df <- df %>%
+  filter(buyer == "National Defence")
+
+# Summarize total spending per year
+military_spending_per_year <- military_df %>%
+  group_by(year) %>%
+  summarize(total_spending = sum(amount, na.rm = TRUE))
+
+# Create the bar chart
+ggplot(data = military_spending_per_year, aes(x = year, y = total_spending)) +
+  geom_bar(stat = "identity", fill = "steelblue") +
+  labs(title = "Military Spending Per Year", x = "Year", y = "Total Spending") +
+  theme_minimal() +
+  scale_y_continuous(labels = dollar_format(prefix = "$")) +
+  theme(plot.title = element_text(hjust = 0.5))
 
 ```
+
+```{r}
+
+df$year <- format(df$award_date, "%Y")
+
+# Filter for military spending (where buyer is "National Defence")
+public_services_df <- df %>%
+  filter(buyer == "Public Services and Procurement Canada")
+
+# Summarize total spending per year
+public_spending_per_year <- public_services_df %>%
+  group_by(year) %>%
+  summarize(total_spending = sum(amount, na.rm = TRUE))
+
+# Create the bar chart
+ggplot(data = public_spending_per_year, aes(x = year, y = total_spending)) +
+  geom_bar(stat = "identity", fill = "steelblue") +
+  labs(title = "Public Services Spending Per Year", x = "Year", y = "Total Spending") +
+  theme_minimal() +
+  scale_y_continuous(labels = dollar_format(prefix = "$")) +
+  theme(plot.title = element_text(hjust = 0.5))
+
+```
+
+```{r}
+df$year <- format(df$award_date, "%Y")
+
+# Filter data for the specified buyers
+selected_buyers <- c(
+  "National Defence",
+  "Public Services and Procurement Canada",
+  "Shared Services Canada",
+  "Public Health Agency of Canada",
+  "Fisheries and Oceans Canada"
+)
+
+stacked_df <- df %>%
+  filter(buyer %in% selected_buyers) %>%
+  group_by(year, buyer) %>%
+  summarize(total_spending = sum(amount, na.rm = TRUE)) %>%
+  ungroup()
+
+# Ensure the order of the stacks matches the specified order
+stacked_df$buyer <- factor(stacked_df$buyer, levels = selected_buyers)
+
+# Create the stacked bar chart
+ggplot(data = stacked_df, aes(x = year, y = total_spending, fill = buyer)) +
+  geom_bar(stat = "identity") +
+  scale_fill_manual(
+    values = c("National Defence" = "steelblue",
+               "Public Services and Procurement Canada" = "orange",
+               "Shared Services Canada" = "green",
+               "Public Health Agency of Canada" = "purple",
+               "Fisheries and Oceans Canada" = "red")
+  ) +
+  labs(
+    title = "Spending by Selected Buyers Per Year",
+    x = "Year",
+    y = "Total Spending",
+    fill = "Buyer"
+  ) +
+  theme_minimal() +
+  scale_y_continuous(labels = dollar_format(prefix = "$")) +
+  theme(plot.title = element_text(hjust = 0.5))
+
+```
+```{r}
+df$year <- format(df$award_date, "%Y")
+
+# Filter data for the specified buyers
+selected_buyers <- c(
+  "National Defence",
+  "Public Services and Procurement Canada",
+  "Shared Services Canada",
+  "Public Health Agency of Canada",
+  "Fisheries and Oceans Canada"
+)
+
+grouped_df <- df %>%
+  filter(buyer %in% selected_buyers) %>%
+  group_by(year, buyer) %>%
+  summarize(total_spending = sum(amount, na.rm = TRUE)) %>%
+  ungroup()
+
+# Ensure the order of the groups matches the specified order
+grouped_df$buyer <- factor(grouped_df$buyer, levels = selected_buyers)
+
+# Create the grouped bar chart
+ggplot(data = grouped_df, aes(x = year, y = total_spending, fill = buyer)) +
+  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
+  scale_fill_manual(
+    values = c("National Defence" = "steelblue",
+               "Public Services and Procurement Canada" = "orange",
+               "Shared Services Canada" = "green",
+               "Public Health Agency of Canada" = "purple",
+               "Fisheries and Oceans Canada" = "red")
+  ) +
+  labs(
+    title = "Spending by Selected Buyers Per Year",
+    x = "Year",
+    y = "Total Spending",
+    fill = "Buyer"
+  ) +
+  theme_minimal() +
+  theme(plot.title = element_text(hjust = 0.5))
+
+```
+
+```{r}
+
+df$year <- format(df$award_date, "%Y")
+
+# Define the selected buyers
+selected_buyers <- c(
+  "National Defence",
+  "Public Services and Procurement Canada",
+  "Shared Services Canada",
+  "Public Health Agency of Canada",
+  "Fisheries and Oceans Canada"
+)
+
+# Add a new column for buyer categories
+df <- df %>%
+  mutate(buyer_category = ifelse(buyer %in% selected_buyers, buyer, "Other"))
+
+# Summarize total spending per year for each buyer category
+grouped_df <- df %>%
+  group_by(year, buyer_category) %>%
+  summarize(total_spending = sum(amount, na.rm = TRUE)) %>%
+  ungroup()
+
+# Ensure the order of the groups matches the specified order, with "Other" last
+grouped_df$buyer_category <- factor(
+  grouped_df$buyer_category,
+  levels = c(selected_buyers, "Other")
+)
+
+# Create the grouped bar chart
+ggplot(data = grouped_df, aes(x = year, y = total_spending, fill = buyer_category)) +
+  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
+  scale_fill_manual(
+    values = c(
+      "National Defence" = "steelblue",
+      "Public Services and Procurement Canada" = "orange",
+      "Shared Services Canada" = "green",
+      "Public Health Agency of Canada" = "purple",
+      "Fisheries and Oceans Canada" = "red",
+      "Other" = "gray"
+    )
+  ) +
+  labs(
+    title = "Spending by Selected and Other Buyers Per Year",
+    x = "Year",
+    y = "Total Spending",
+    fill = "Buyer"
+  ) +
+  theme_minimal() +
+  theme(plot.title = element_text(hjust = 0.5))
+
+```

### 2024-11-28 scripts/03-data_cleaning.R
parsing supplier names

@@ -12,9 +12,12 @@ library(tidyverse)
 #### Clean data ####
 combined <- read_csv("data/raw_data/combined_data.csv")
 
+
+
 View(combined)
 # Step 1: Create a copy of the original dataframe
 cleaned <- combined
+
 
 # Step 2: Convert 'amount' to numeric in the 'cleaned' dataframe
 cleaned$amount <- as.numeric(gsub("[\\$,]", "", cleaned$amount))
@@ -83,7 +86,148 @@ head(duplicate_rows)
 
 cleaned <- unique(cleaned)
 
+# Load necessary libraries
+library(dplyr)
+library(stringr)
+
+# Create a list of words to remove
+banned_words <- c(
+  "inc", "ltd", "limited", "company", "co", "corporation", "corp", "llc", "the", "and", "of", "canada", "sa",
+  "services", "solutions", "technologies", "consulting", "group", "systems", "construction", "canadian",
+  "international", "technology", "fuels", "de", "scientific", "university", "associates",
+  "ottawa", "air", "northern", "business", "engineering", "global", "oil", "micro", "llp", "communications",
+  "research", "motor", "les", "marine", "helicopters", "industries", "interiors", "aviation", "products",
+  "consultants", "environmental", "ontario", "fuel", "energy", "management", "equipment", "nisha", "advanced",
+  "electric", "service", "lp", "authority", "enterprises", "harbour", "a", "fisher", "resources", "staffing",
+  "office", "quebec", "centre", "mechanical", "north", "security", "supplies", "general", "human", "contracting",
+  "supply", "world", "industrial", "institute", "incorporated", "maritime", "electronics", "translation",
+  "groupe", "professional", "power", "language", "corps", "it", "partnership", "joint", "training",
+  "military", "lighting", "program", "defence", "jpo", "ii", "lightning", "space", "macdonald", "land"
+)
+
+word_pairs_to_concatenate <- list(
+  c("vancouver", "shipyards")
+  # Add more pairs here if needed, e.g., c("another", "pair")
+)
+
+concatenate_specific_pairs <- function(words, word_pairs) {
+  for(pair in word_pairs) {
+    # Check if all words in the pair are present
+    if(all(pair %in% words)) {
+      # Concatenate the pair into a single string
+      concatenated <- paste(pair, collapse = " ")
[…]

### 2024-11-28 scripts/05-exploratory_data_analysis.R
more exploration

@@ -12,12 +12,29 @@ library(rstanarm)
 
 #### Read data ####
 df <- read_csv("data/analysis_data/cleaned_data.csv")
+View(df)
 
-u_supps <- data.frame(unique(df$supplier))
+u_dsupps <- df %>%
+  filter(buyer == "National Defence", amount >= 100000) %>%
+  group_by(supplier) %>%
+  summarize(total_amount = sum(amount, na.rm = TRUE), .groups = "drop")
+
+# Display the unique suppliers
+View(u_dsupps)
+
+u_contracts <- data.frame(unique(df$contract))
+
+max.print(u_contracts)
 
 u_buyers <- data.frame(unique(df$buyer))
 
+u_supps <- data.frame(unique(df$supplier))
+
+View(u_supps)
+
 public_health <- df[df$buyer == "Public Health Agency of Canada", ]
+
+View(public_health)
 
 defence <- df[df$buyer == "National Defence", ]
 

### 2024-11-28 scripts/01-download_data.R
added View

@@ -97,6 +97,7 @@ list_of_data_frames <- lapply(csv_files, read_csv)
 # Combine all data frames into one
 all_data <- bind_rows(list_of_data_frames)
 
+View(all_data)
 
 # Save the combined data to a single CSV file
 combined_csv_path <- file.path(host_download_dir, "combined_data.csv")

### 2024-11-28 other/EDA.qmd
added another graph

@@
 ```{r}
+# Add "year" column to the dataframe
 df$year <- format(df$award_date, "%Y")
 
-# Filter data for the specified buyers
+# Define selected buyers
 selected_buyers <- c(
@@
 
+# Create a new column to categorize buyers
+df$buyer_category <- ifelse(df$buyer %in% selected_buyers, df$buyer, "Other")
+
+# Summarize data
 stacked_df <- df %>%
-  filter(buyer %in% selected_buyers) %>%
-  group_by(year, buyer) %>%
+  group_by(year, buyer_category) %>%
   summarize(total_spending = sum(amount, na.rm = TRUE)) %>%
@@
 # Ensure the order of the stacks matches the specified order
-stacked_df$buyer <- factor(stacked_df$buyer, levels = selected_buyers)
+stacked_df$buyer_category <- factor(stacked_df$buyer_category, levels = c(selected_buyers, "Other"))
 
 # Create the stacked bar chart
-ggplot(data = stacked_df, aes(x = year, y = total_spending, fill = buyer)) +
+ggplot(data = stacked_df, aes(x = year, y = total_spending, fill = buyer_category)) +
   geom_bar(stat = "identity") +
   scale_fill_manual(
-    values = c("National Defence" = "steelblue",
+    values = c(
+      "National Defence" = "steelblue",
       "Public Services and Procurement Canada" = "orange",
@@
       "Public Health Agency of Canada" = "purple",
-               "Fisheries and Oceans Canada" = "red")
+      "Fisheries and Oceans Canada" = "red",
+      "Other" = "gray"
+    )
   ) +
@@
   theme_minimal() +
-  scale_y_continuous(labels = dollar_format(prefix = "$")) +
+  scale_y_continuous(labels = scales::dollar_format(prefix = "$")) +
   theme(plot.title = element_text(hjust = 0.5))
-
 ```
@@
   theme_minimal() +
+  scale_y_continuous(labels = dollar_format(prefix = "$")) +
   theme(plot.title = element_text(hjust = 0.5))
@@
 ```
+
+```{r}
+hist(df$amount,
+     breaks = 30,  # Number of bins; adjust as needed
+     main = "Distribution of Award Amounts",
+     xlab = "Award Amount (CAD)",
+     ylab = "Frequency",
+     col = "skyblue",
+     border = "black")
+
+```

### 2024-11-28 scripts/03-data_cleaning.R
parsing suppliers better

@@ -102,7 +102,8 @@ banned_words <- c(
   "office", "quebec", "centre", "mechanical", "north", "security", "supplies", "general", "human", "contracting",
   "supply", "world", "industrial", "institute", "incorporated", "maritime", "electronics", "translation",
   "groupe", "professional", "power", "language", "corps", "it", "partnership", "joint", "training",
-  "military", "lighting", "program", "defence", "jpo", "ii", "lightning", "space", "macdonald", "land"
+  "military", "lighting", "program", "defence", "jpo", "ii", "lightning", "space", "macdonald", "land", "systemscanada",
+  "in", "aeronautical"
 )
 
 word_pairs_to_concatenate <- list(
@@ -158,11 +159,11 @@ word_counts <- word_counts %>%
 
 #View(word_counts)
 
-View(cleaned)
+#View(cleaned)
 
 df <- cleaned
 
-View(df)
+#View(df)
 
 word_dict <- c()
 

### 2024-11-28 scripts/03-data_cleaning.R
fixed up

@@ -230,6 +230,7 @@ ggplot(top_ten, aes(x = reorder(word, total_amount), y = total_amount)) +
   ) +
   theme_minimal()
 
+cleaned <- df
 
 #### Save data ####
 write_csv(cleaned, "data/analysis_data/cleaned_data.csv")

### 2024-11-28 scripts/03-data_cleaning.R
added script to download top ten suppliers, and fixed cleaned_dataset.csv not storing processed_suppliers correctly

@@ -90,6 +90,8 @@ cleaned <- unique(cleaned)
 library(dplyr)
 library(stringr)
 
+df <- cleaned
+
 # Create a list of words to remove
 banned_words <- c(
   "inc", "ltd", "limited", "company", "co", "corporation", "corp", "llc", "the", "and", "of", "canada", "sa",
@@ -106,11 +108,14 @@ banned_words <- c(
   "in", "aeronautical"
 )
 
+# 2. Define Word Pairs to Concatenate
+# Each element is a vector of words to concatenate
 word_pairs_to_concatenate <- list(
   c("vancouver", "shipyards")
   # Add more pairs here if needed, e.g., c("another", "pair")
 )
 
+# Function to concatenate specific word pairs
 concatenate_specific_pairs <- function(words, word_pairs) {
   for(pair in word_pairs) {
     # Check if all words in the pair are present
@@ -126,8 +131,8 @@ concatenate_specific_pairs <- function(words, word_pairs) {
   return(words)
 }
 
-
-cleaned <- cleaned %>%
+# 3. Process Supplier Names
+df <- df %>%
   mutate(
     # Lowercase all characters and remove punctuation
     processed_supplier = supplier %>%
@@ -146,28 +151,10 @@ cleaned <- cleaned %>%
     })
   )
 
-
-word_counts <- cleaned %>%
-  unnest(processed_supplier) %>%                 # Expand the lists into rows
-  group_by(processed_supplier) %>%               # Group by each word
-  summarize(count = n()) %>%                     # Count occurrences
-  arrange(desc(count))                           # Sort by descending frequency
-
-# Rename columns for clarity
-word_counts <- word_counts %>%
-  rename(word = processed_supplier)
-
-#View(word_counts)
-
-#View(cleaned)
-
-df <- cleaned
-
-#View(df)
-
+# 4. Initialize an Empty Named Vector for the Dictionary
 word_dict <- c()
 
-# 4. Loop Through Each Supplier and Accumulate Amounts
+# 5. Loop Through Each Supplier and Accumulate Amounts
 for (i in 1:nrow(df)) {
   words <- df$processed_supplier[[i]]
   amt <- df$amount[i]
@@ -181,19 +168,19 @@ for (i in 1:nrow(df)) {
   }
 }
 
-# 5. Convert the Dictionary to a Dataframe
+# 6. Convert the Dictionary to a Dataframe
 word_df <- data.frame(
[…]

### 2024-11-28 scripts/06-model.R
added 6 simple linear models, 5 for the top biggest Buyers and 1 for the rest of Buyers

@@ -1,37 +1,70 @@
 #### Preamble ####
-# Purpose: Models... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Models the 5 biggest Buyers and Other(rest of the Buyers)
+# Author: Robert Ford
+# Date: 28 November 2024
+# Contact: robert.ford@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+
 
 
 #### Workspace setup ####
 library(tidyverse)
-library(rstanarm)
+library(dplyr)
 
 #### Read data ####
-analysis_data <- read_csv("data/analysis_data/analysis_data.csv")
+df <- read_csv("data/analysis_data/cleaned_data.csv")
 
 ### Model data ####
-first_model <-
-  stan_glm(
-    formula = flying_time ~ length + width,
-    data = analysis_data,
-    family = gaussian(),
-    prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_aux = exponential(rate = 1, autoscale = TRUE),
-    seed = 853
+# Define the specific buyers
+specific_buyers <- c(
+  "National Defence",
+  "Public Services and Procurement Canada",
+  "Shared Services Canada",
+  "Public Health Agency of Canada",
+  "Fisheries and Oceans Canada"
+)
+
+# Create a new column 'buyer_group' categorizing each buyer
+df <- df %>%
+  mutate(
+    buyer_group = ifelse(buyer %in% specific_buyers, buyer, "Other")
   )
+
+df <- df %>%
+  mutate(
+    award_date_numeric = as.numeric(award_date)
+  )
+
+# Split the dataframe into a list of dataframes by 'buyer_group'
+buyer_groups <- split(df, df$buyer_group)
+
+# Initialize an empty list to store the models
+models <- list()
+
+# Loop through each buyer group and fit the linear model
+for (buyer in names(buyer_groups)) {
+  # Subset the data for the current buyer group
+  data_subset <- buyer_groups[[buyer]]
+  
+  # Check if there are enough data points to fit a model
+  if(nrow(data_subset) > 1) {  # At least two points needed for regression
[…]

### 2024-11-28 other/EDA/EDA.qmd
fixed file structure

--- 2024-11-28 other/EDA.qmd
@@
 
-df <- df %>% select(-link)
+View(df)
 
@@
 ```{r}
-hist(df$amount,
-     breaks = 30,  # Number of bins; adjust as needed
-     main = "Distribution of Award Amounts",
+#View(df)
+
+# Define the bin width
+bin_width <- 100000
+
+# Determine the range of 'amount'
+min_amount <- floor(min(df$amount, na.rm = TRUE) / bin_width) * bin_width
+max_amount <- ceiling(1000000000 / bin_width) * bin_width
+
+# Create a sequence of breaks
+breaks <- seq(min_amount, max_amount, by = bin_width)
+
+# Categorize 'amount' into bins
+df$amount_bin <- cut(df$amount, breaks = breaks, right = FALSE, include.lowest = TRUE)
+
+# Count the number of contracts in each bin
+contract_counts <- table(df$amount_bin)
+
+# Convert to a dataframe for better readability
+contract_counts_df <- as.data.frame(contract_counts)
+colnames(contract_counts_df) <- c("Amount_Range", "Contract_Count")
+
+# View the counts
+print(contract_counts_df)
+
+# Basic Bar Plot
+barplot(contract_counts,
+        main = "Number of Contracts per 1,000,000 CAD Bin",
         xlab = "Award Amount (CAD)",
-     ylab = "Frequency",
+        ylab = "Number of Contracts",
         col = "skyblue",
-     border = "black")
+        las = 2,  # Make label text perpendicular to axis
+        cex.names = 0.7)  # Adjust label size if needed
 

### 2024-11-28 scripts/06-model.R
renamed models storage

@@ -66,5 +66,5 @@ print(model_summaries)
 
 
 #### Save model ####
-saveRDS(models, file = "models/buyer_group_linear_models.rds")
+saveRDS(models, file = "models/buyer_models.rds")
 

### 2024-11-28 paper/paper.qmd
set up outline

---
title: "Procurements of the Canadian Government, Runaway Military Contracts"
author: 
  - Robert Ford
thanks: "Code and data are available at: https://github.com/Ford-Robert"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)
library(dplyr)
library(lubridate)
library(scales)
library(knitr)
library(kableExtra)
library(zoo)

df <- read_csv(here::here("data/analysis_data/cleaned_data.csv"))
aggregated_df <- (here::here("data/analysis_data/top_ten_suppliers.csv"))
models <- read_csv(here::here("models/buyer_models.rds"))
```

# Introduction

TODO Broader context of Canadian spending, talk about huge new military spending package announced this year

TODO What are procurement, how are they awarded, what are Buyers and Suppliers

TODO Gap that needs to be filled; What is the Government buying from private companies and how is their consumption changing over time

TODO What was done

TODO What was found

TODO Why is it important

TODO Paper layout, description of each section with jump links

TODO Cite R and all other packages

Estimand?

# Data {#sec-data}

TODO Description of each variable head of table

```{r fig-head}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: template
#| fig-align: center

df_table <- df %>% select(-contract, -start_date, -end_date)

kable(head(df_table, 5), caption = "Sample of the Data") %>%
  kable_styling(font_size = 10, latex_options = c("scale_down"))

```

TODO Brief talk about how cleaning was carried out

TODO Description of variables that were created, duration_days, processed_supplier (brief, see appendix)

## Shape

TODO Talk about top 5 buyers and other

```{r fig-top-buyers}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: template
#| fig-align: center

#TODO color each buyer with the correct color

# Summarize total amount spent by each buyer
buyer_totals <- df %>%
  group_by(buyer) %>%
  summarise(total_amount = sum(amount, na.rm = TRUE)) %>%
  arrange(desc(total_amount))

# Identify the top 5 buyers
top_five_buyers <- buyer_totals %>%
  slice_head(n = 5) %>%
  pull(buyer)

# Create a new column 'buyer_group' where buyers not in top 5 are labeled as 'Other'
buyer_totals_grouped <- buyer_totals %>%
  mutate(buyer_group = ifelse(buyer %in% top_five_buyers, buyer, "Other")) %>%
  group_by(buyer_group) %>%
  summarise(total_amount = sum(total_amount, na.rm = TRUE)) %>%
  arrange(desc(total_amount))

# Plot the bar chart
ggplot(buyer_totals_grouped, aes(x = reorder(buyer_group, total_amount), y = total_amount)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +
  labs(title = "Top 5 Buyers by Total Amount Spent and Others",
       x = "Buyer",
       y = "Total Amount Spent (CAD)") +
  scale_y_continuous(labels = comma) +
  theme_minimal()

```

TODO Top Ten Suppliers

```{r fig-top-supps}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: template
#| fig-align: center

aggregated_df <- aggregated_df %>%
  left_join(top_ten %>% select(word, supplier_name), by = "word")

# Create the bar chart with legend removed
ggplot(aggregated_df, aes(x = reorder(word, total_amount), y = total_amount, fill = buyer)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Top 10 Suppliers by Aggregated Amount Awarded",
    x = "Supplier",
    y = "Total Amount Awarded"
    # Remove the legend title by not specifying 'fill' in labs
  ) +
  theme_minimal() +
  theme(legend.position = "none")

# 10. Transform `processed_supplier` into a Character Column for CSV
df <- df %>%
  mutate(
    processed_supplier = sapply(processed_supplier, function(x) {
      if(length(x) > 0) {
        paste(x, collapse = " ")
      } else {
        ""  # Represent empty lists as empty strings
      }
    })
  )
```

TODO Military Spending Per year

```{r fig-mil-spending}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: template
#| fig-align: center

```

## Measurement

TODO How IJF collected all this data

# Model

TODO What the is the goal of our models

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

TODO define variables used and structure

TODO Cite R and packages used to run model

### Model justification

TODO Why I chose those variables

# Results

TODO Add model summary tables

# Discussion

## First discussion point {#sec-first-point}

## Second discussion point

## Third discussion point

## Weaknesses and next steps

\newpage

\appendix

# Appendix {-}

# Additional data details

TODO Removal of negative values, adding 1 day to contract lengths of 0, see cleaning script

TODO processed_supplier

# Model details {#sec-model-details}

## Posterior predictive check

## Diagnostics

\newpage

# References

### 2024-11-28 paper/paper.qmd
added mil spending graph

@@
 
+df_filtered <- df %>%
+  filter(format(award_date, "%Y") != "2019")
+
+df_filtered$year <- format(df_filtered$award_date, "%Y")
+
+# Filter for military spending (where buyer is "National Defence")
+public_services_df <- df_filtered %>%
+  filter(buyer == "Public Services and Procurement Canada")
+
+# Summarize total spending per year
+public_spending_per_year <- public_services_df %>%
+  group_by(year) %>%
+  summarize(total_spending = sum(amount, na.rm = TRUE))
+
+# Create the bar chart
+ggplot(data = public_spending_per_year, aes(x = year, y = total_spending)) +
+  geom_bar(stat = "identity", fill = "steelblue") +
+  labs(title = "Public Services Spending Per Year", x = "Year", y = "Total Spending") +
+  theme_minimal() +
+  scale_y_continuous(labels = dollar_format(prefix = "$")) +
+  theme(plot.title = element_text(hjust = 0.5))
+
 ```

### 2024-11-28 paper/paper.qmd
added public spending graph

@@
 TODO Military Spending Per year
-
 ```{r fig-mil-spending}
@@
 #| fig-align: center
+
+#TODO Make right color
+
+df_filtered <- df %>%
+  filter(format(award_date, "%Y") != "2019")
+
+# Convert award_date to Date type and extract the year
+df_filtered$award_date <- as.Date(df_filtered$award_date, format = "%Y-%m-%d")
+df_filtered$year <- format(df_filtered$award_date, "%Y")
+
+# Filter for military spending (where buyer is "National Defence")
+military_df <- df_filtered %>%
+  filter(buyer == "National Defence")
+
+# Summarize total spending per year
+military_spending_per_year <- military_df %>%
+  group_by(year) %>%
+  summarize(total_spending = sum(amount, na.rm = TRUE))
+
+# Create the bar chart
+ggplot(data = military_spending_per_year, aes(x = year, y = total_spending)) +
+  geom_bar(stat = "identity", fill = "steelblue") +
+  labs(title = "Military Spending Per Year", x = "Year", y = "Total Spending") +
+  theme_minimal() +
+  scale_y_continuous(labels = dollar_format(prefix = "$")) +
+  theme(plot.title = element_text(hjust = 0.5))
+
+```
+
+TODO Public Services Spending
+```{r fig-public-spending}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-cap: template
+#| fig-align: center
+
+#TODO Make right color
 

### 2024-11-29 other/EDA/EDA.qmd
filtering for clearer graphs
exploring small contracts (<=200k)

--- 2024-11-28 other/EDA.qmd
@@
 
-# Filter for military spending (where buyer is "National Defence")
+# Filter for public services
 public_services_df <- df %>%
@@
 ```{r}
+df_filtered <- df %>%
+  filter(format(award_date, "%Y") != "2019")
+
 # Add "year" column to the dataframe
-df$year <- format(df$award_date, "%Y")
+df_filtered$year <- format(df$award_date, "%Y")
 
@@
 # Create a new column to categorize buyers
-df$buyer_category <- ifelse(df$buyer %in% selected_buyers, df$buyer, "Other")
+df_filtered$buyer_category <- ifelse(df_filtered$buyer %in% selected_buyers, df_filtered$buyer, "Other")
 
 # Summarize data
-stacked_df <- df %>%
+stacked_df <- df_filtered %>%
   group_by(year, buyer_category) %>%
@@
 ```{r}
+df_filtered <- df %>%
+  filter(format(award_date, "%Y") != "2019")
 
-df$year <- format(df$award_date, "%Y")
+df_filtered$year <- format(df_filtered$award_date, "%Y")
 
@@
 # Add a new column for buyer categories
-df <- df %>%
+df_filtered <- df_filtered %>%
   mutate(buyer_category = ifelse(buyer %in% selected_buyers, buyer, "Other"))
@@
 # Summarize total spending per year for each buyer category
-grouped_df <- df %>%
+grouped_df <- df_filtered %>%
   group_by(year, buyer_category) %>%
@@
 ```{r}
-hist(df$amount,
-     breaks = 30,  # Number of bins; adjust as needed
-     main = "Distribution of Award Amounts",
+#View(df)
+
+df_filtered <- df %>%
+  filter(amount <= 200000)
+
+# Optional: Check how many rows are left
+num_less_than_200k <- nrow(df_filtered)
+total_contracts <- nrow(df)
+
+percent_less_than <- num_less_than_200k / total_contracts
+
+print(percent_less_than)
+
+# Step 3: Create bins spanning every 10,000 CAD up to 200,000
+# Define the bin width
+bin_width <- 10000  # 10,000 CAD
+
+# Create a sequence of break points from 0 to 200,000
+breaks <- seq(0, 200000, by = bin_width)
+
+# To ensure that the upper limit includes 200,000, set 'right = FALSE'
+# This means intervals are [0,10000), [10000,20000), ..., [190000,200000]
+df_filtered <- df_filtered %>%
+  mutate(amount_bin = cut(amount, breaks = breaks, right = FALSE, include.lowest = TRUE))
+
+# Step 4: Count the number of contracts in each bin
+contract_counts_df <- df_filtered %>%
+  group_by(amount_bin) %>%
+  summarize(Contract_Count = n()) %>%
+  ungroup()
+
+# View the binned counts
+print(contract_counts_df)
+
+contract_counts_df$amount_bin <- as.character(contract_counts_df$amount_bin)
+
+# Create the bar plot
+barplot(
+  height = contract_counts_df$Contract_Count,
+  names.arg = contract_counts_df$amount_bin,
+  las = 2,  # Rotate x-axis labels to perpendicular
+  col = "skyblue",
+  border = "black",
+  main = "Number of Contracts per 10,000 CAD Award Amount Range",
   xlab = "Award Amount (CAD)",
-     ylab = "Frequency",
+  ylab = "Number of Contracts",
+  cex.names = 0.8  # Adjust label size for readability
+)
+```
+
+```{r}
+
+df_filtered <- df %>%
+  filter(buyer == "National Defence", amount <= 200000)
+
+# Optional: Check how many rows are left
+num_less_than_200k <- nrow(df_filtered)
+total_contracts <- nrow(df)
+
+percent_less_than <- num_less_than_200k / total_contracts
+
+print(percent_less_than)
+
+# Step 3: Create bins spanning every 10,000 CAD up to 200,000
+# Define the bin width
+bin_width <- 10000  # 10,000 CAD
+
+# Create a sequence of break points from 0 to 200,000
+breaks <- seq(0, 200000, by = bin_width)
+
+# To ensure that the upper limit includes 200,000, set 'right = FALSE'
+# This means intervals are [0,10000), [10000,20000), ..., [190000,200000]
+df_filtered <- df_filtered %>%
+  mutate(amount_bin = cut(amount, breaks = breaks, right = FALSE, include.lowest = TRUE))
+
+# Step 4: Count the number of contracts in each bin
+contract_counts_df <- df_filtered %>%
+  group_by(amount_bin) %>%
+  summarize(Contract_Count = n()) %>%
+  ungroup()
+
+# View the binned counts
+print(contract_counts_df)
+
+contract_counts_df$amount_bin <- as.character(contract_counts_df$amount_bin)
+
+# Create the bar plot
+barplot(
+  height = contract_counts_df$Contract_Count,
+  names.arg = contract_counts_df$amount_bin,
+  las = 2,  # Rotate x-axis labels to perpendicular
   col = "skyblue",
-     border = "black")
+  border = "black",
+  main = "Number of Contracts per 10,000 CAD Award Amount Range",
+  xlab = "Award Amount (CAD)",
+  ylab = "Number of Contracts",
+  cex.names = 0.8  # Adjust label size for readability
+)
 

### 2024-11-29 paper/paper.qmd
added more TODO

@@
 df <- read_csv(here::here("data/analysis_data/cleaned_data.csv"))
-aggregated_df <- (here::here("data/analysis_data/top_ten_suppliers.csv"))
+aggregated_df <- read_csv(here::here("data/analysis_data/top_ten_suppliers.csv"))
 models <- read_csv(here::here("models/buyer_models.rds"))
+
+View(df)
 ```
@@
 
-aggregated_df <- aggregated_df %>%
-  left_join(top_ten %>% select(word, supplier_name), by = "word")
+#TODO Fix colors and supplier names
 
@@
 
-# 10. Transform `processed_supplier` into a Character Column for CSV
-df <- df %>%
-  mutate(
-    processed_supplier = sapply(processed_supplier, function(x) {
-      if(length(x) > 0) {
-        paste(x, collapse = " ")
-      } else {
-        ""  # Represent empty lists as empty strings
-      }
-    })
-  )
 ```
@@
 
+TODO add spending by selected buyers graph
+
 ## Measurement

### 2024-11-29 scripts/03-data_cleaning.R
cleaned up file

@@ -255,7 +255,10 @@ df <- df %>%
 
 cleaned <- df
 
-#View(cleaned)
+View(cleaned)
+
+# -------------------------------------------------------------------------
+
 
 #### Save data ####
 write_csv(cleaned, "data/analysis_data/cleaned_data.csv")

### 2024-11-29 scripts/03-data_cleaning.R
cleaned up

@@ -255,10 +255,7 @@ df <- df %>%
 
 cleaned <- df
 
-View(cleaned)
-
-# -------------------------------------------------------------------------
-
+#View(cleaned)
 
 #### Save data ####
 write_csv(cleaned, "data/analysis_data/cleaned_data.csv")

### 2024-11-29 scripts/05-exploratory_data_analysis.R
added percentage calculations for small contracts

@@ -56,6 +56,16 @@ contracts_expiring <- df %>%
   ) %>%
   ungroup()
 
+space_total <- df %>%
+  filter(buyer == "Canadian Space Agency") %>%
+  summarize(total_amount = sum(amount, na.rm = TRUE)) %>%
+  pull(total_amount)
+
+print(space_total)
+
+print(2199059594/ space_total)
+
+
 View(contracts_expiring)
 View(public_health)
 View(defence)

### 2024-11-29 scripts/01-download_data.R
added prereqs and Other info to headers

@@ -1,30 +1,36 @@
 #### Preamble ####
-# Purpose: A webscraper to grab ijf procurement awards data, and collate into one combined dataset
+# Purpose: A webscraper to grab IJF procurement awards data, and collate into one combined dataset
 # Author: Robert Ford
 # Date: 21 November 2024
 # Contact: robert.ford@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: Docker, Firefox Web driver
-# See appendix TODO for more details
-# Data was collected on: 24 November 2024
+# Pre-requisites: 
+# - The application Docker, and Firefox Web driver must be installed
+# - The `RSelenium` package must be installed and loaded
+# - The `rvest` package must be installed and loaded
+# - The `dplyr` package must be installed and loaded
+# - The `readr` package must be installed and loaded
 
-#TODO RE-RUN!
+# See appendix of paper for detailed step by step instructions
+# Be sure to fill in all instances of USER_PATH with where ever you have stored this rproj on your machine
 
+# -! IMPORTANT !- You must FIRST run this code in your terminal which is cd in this file:
+# docker run -d -p 4444:4444 -p 5900:5900 -v /USER_PATH/canadian_gov_procurement/data/raw_data:/home/seluser/Downloads selenium/standalone-firefox:3.141.59
+
+
+# Data was last collected on: 24 November 2024
 #### Workspace setup ####
 library(RSelenium)
 library(rvest)
 library(dplyr)
 library(readr)
-library(readr)
-library(dplyr)
-#docker run -d -p 4444:4444 -p 5900:5900 -v /Users/robert_ford/canadian_gov_procurement/data/raw_data:/home/seluser/Downloads selenium/standalone-firefox:3.141.59
 
 # Paths
 # Path inside the container
 download_dir <- "/home/seluser/Downloads"
 
 # Path on the host
-host_download_dir <- "/Users/robert_ford/canadian_gov_procurement/data/raw_data"
+host_download_dir <- "/USER_PATH/canadian_gov_procurement/data/raw_data"
 
 # Set up Firefox profile with download preferences
 fprof <- makeFirefoxProfile(list(
@@ -80,7 +86,6 @@ for (page in 1:total_pages) {
     })
   }
 }
-
 
 host_download_dir <- "data/raw_data"
 

### 2024-11-29 scripts/03-data_cleaning.R

@@ -4,10 +4,17 @@
 # Date: 22 November 2024
 # Contact: robert.ford@mail.utoronto.ca
 # License: MIT
+# Pre-requisites: 
+# - The `tidyverse` package must be installed and loaded
+# - The `dplyr` package must be installed and loaded
+# - The `stringr` package must be installed and loaded
+# Other information? Make sure you are in the `canadian_gov_procurement` rproj
 
 
 #### Workspace setup ####
 library(tidyverse)
+library(dplyr)
+library(stringr)
 
 #### Clean data ####
 combined <- read_csv("data/raw_data/combined_data.csv")
@@ -85,10 +92,6 @@ head(duplicate_rows)
 
 
 cleaned <- unique(cleaned)
-
-# Load necessary libraries
-library(dplyr)
-library(stringr)
 
 df <- cleaned
 

### 2024-11-29 scripts/04-test_analysis_data.R

@@ -4,6 +4,10 @@
 # Date: 22 November 2024
 # Contact: robert.ford@mail.utoronto.ca
 # License: MIT
+# Pre-requisites: 
+# - The `tidyverse` package must be installed and loaded
+# - The `testthat` package must be installed and loaded
+# Any other information needed? Make sure you are in the `canadian_gov_procurement` rproj
 
 
 

### 2024-11-29 scripts/05-exploratory_data_analysis.R

@@ -4,11 +4,13 @@
 # Date: 23 November 2024
 # Contact: robert.ford@mail.utoronto.ca
 # License: MIT
+# Pre-requisites: 
+# - The `tidyverse` package must be installed and loaded
+# Other information? Make sure you are in the `canadian_gov_procurement` rproj
 
 
 #### Workspace setup ####
 library(tidyverse)
-library(rstanarm)
 
 #### Read data ####
 df <- read_csv("data/analysis_data/cleaned_data.csv")

### 2024-11-29 scripts/06-model.R

@@ -4,6 +4,11 @@
 # Date: 28 November 2024
 # Contact: robert.ford@mail.utoronto.ca
 # License: MIT
+# Pre-requisites: 
+# - The `tidyverse` package must be installed and loaded
+# - The `dplyr` package must be installed and loaded
+# - 
+# Other information? Make sure you are in the `canadian_gov_procurement` rproj
 
 
 
@@ -13,6 +18,8 @@ library(dplyr)
 
 #### Read data ####
 df <- read_csv("data/analysis_data/cleaned_data.csv")
+
+#TODO Simplify Down to one model
 
 ### Model data ####
 # Define the specific buyers

### 2024-11-29 paper/paper.qmd
added webscraping appendix

@@
 TODO Description of each variable head of table
-
 ```{r fig-head}
@@
 TODO Talk about top 5 buyers and other
-
 ```{r fig-top-buyers}
@@
   theme_minimal()
-
-```
-
-TODO Top Ten Suppliers
-
-```{r fig-top-supps}
-#| message: false
-#| echo: false
-#| warning: false
-#| fig-cap: template
-#| fig-align: center
-
-#TODO Fix colors and supplier names
-
-# Create the bar chart with legend removed
-ggplot(aggregated_df, aes(x = reorder(word, total_amount), y = total_amount, fill = buyer)) +
-  geom_bar(stat = "identity") +
-  coord_flip() +
-  labs(
-    title = "Top 10 Suppliers by Aggregated Amount Awarded",
-    x = "Supplier",
-    y = "Total Amount Awarded"
-    # Remove the legend title by not specifying 'fill' in labs
-  ) +
-  theme_minimal() +
-  theme(legend.position = "none")
 
@@
 
+TODO Top Ten Suppliers
+```{r fig-top-supps}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-cap: template
+#| fig-align: center
+
+#TODO Fix colors and supplier names
+
+# Create the bar chart with legend removed
+ggplot(aggregated_df, aes(x = reorder(word, total_amount), y = total_amount, fill = buyer)) +
+  geom_bar(stat = "identity") +
+  coord_flip() +
+  labs(
+    title = "Top 10 Suppliers by Aggregated Amount Awarded",
+    x = "Supplier",
+    y = "Total Amount Awarded"
+    # Remove the legend title by not specifying 'fill' in labs
+  ) +
+  theme_minimal() +
+  theme(legend.position = "none")
+
+```
+
 ## Measurement
@@
 TODO How IJF collected all this data
+
+T
 
@@
 
+TODO Model Diagnostics
+
 # Results
@@
 
+TODO What was done
+
 ## Second discussion point
+
+TODO What did we learn about Canada
+    Military spending has exploded in the last couple of years, and other departments have fallen significantly
+    The vast majority of contracts are smaller than 200,000, and the most common size of contract is between 10-20k
 
@@
 
-## Weaknesses and next steps
+## Weaknesses
+
+TODO Model was weak, need more data to improve model
+    Analysis on the Suppliers Names needs to be more rigorous
+
+## Next Steps
+
+TODO Next Steps: Need to collect more data:
+      From Past spending beyond 2020
+      Other types of Government Spending; Seeing how much of spending is done through private contracts and how much through in house spending on salaries and other things like that
+      Do analysis on the what the contracts descriptions say
+      Why does public health Canada spending go to zero
+      Why are contracts so small
 
@@
 # Appendix {-}
+
+# Web Scraping
+These are step by step instructions on how to Web Scrape the IFJ procurement data page, because as of November 2024 there is no easy/automated way of downloading the 330 pages of 1000 rows worth of data.
+
+Note: It is important that docker, java, python and firefox are installed on your machine as RSelenium requires these packages.
+To download Selenium and create a docker image, open the docker application and search for the Selenium package.
+
+1.
+Run a compatible Selenium docker image.
+You must make sure that the version of Selenium you are using is compatible with Firefox or whichever browser you choose.
+You must launch docker from your terminal whose directory matches that of the r file’s path to 01-download_data.R .
+You can check if the docker image has successfully been launched using the docker application's nice GUI.
+It is also important in this step to define the path of the local directory that the data should be saved into, this is shown as USER_PATH in the script.
+
+Terminal Command:
+
+user:canadian_gov_procurement USER_PATH$ docker run -d -p 4444:4444 -p 5900:5900 -v /USER_PATH/canadian_gov_procurement/data/raw_data:/home/seluser/Downloads selenium/standalone-firefox:3.141.59
+
+2.
+It is possible that the tags used to find the buttons on the web-site may change over time.
+So visit the IJF procurement data website and right click and inspect the download and next page buttons.
+This code uses the aria labels to locate these buttons, but there are many ways to identify them using html and css elements.
+
+3. Now the 01-down-load-data.R file can be executed. Be sure to modify USER_PATH in both to download_dir and host_down_load_dir to match the path of you docker and personal computer path.
+
+4.
+The for-loop simply runs "click download -> click next page" process 329 times, however IJF may add more pages so modify total_pages to scrape the number of pages IJF has.
+In the event of an error mid-way through scraping take note of which page failed to download (page that you are on is printed to the console) and simply set the start of the for-loop.
+After restarting the loop be sure to verify that the next downloaded page is the correct one, otherwise you will have to restart the process from step 1.
+
+5. Done! Enjoy!
 
@@
 
-## Posterior predictive check
+## Diagnostics
 
-## Diagnostics
+TODO Add Diagnostics
 

### 2024-11-29 scripts/00-simulate_data.R
simulation and simulation testing

@@ -1,19 +1,270 @@
 #### Preamble ####
-# Purpose: Simulates... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Simulates Government Contract Data
+# Author: Robert Ford
+# Date: 25 November 2024
+# Contact: robert.ford@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: 
+# - The `tidyverse` package must be installed and loaded
+# - The `VGAM` package must be installed and loaded
+# - The `dplyr` package must be installed and loaded
+# - The `lubridate` package must be installed and loaded
+# Any other information? Make sure you are in the `canadian_gov_procurement` rproj
 
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
+library(VGAM)      # For Pareto distribution
+library(dplyr)     # For data manipulation
+library(lubridate)
 
 #### Simulate data ####
-# [...ADD CODE HERE...]
+# Set seed for reproducibility
+set.seed(123)
 
+# 1. Define the buyer list excluding "National Defence"
+buyer_list <- c(
+  "Statistics Canada",
+  "Canadian Institutes of Health Research",
+  "The National Battlefields Commission",
+  "Department of Justice Canada",
+  # "National Defence", # Excluded
+  "Public Services and Procurement Canada",
+  "Canada Revenue Agency",
+  "Canada Border Services Agency",
+  "Natural Resources Canada",
+  "Agriculture and Agri-Food Canada",
+  "Correctional Service of Canada",
+  "Treasury Board of Canada Secretariat",
+  "Environment and Climate Change Canada",
+  "Impact Assessment Agency of Canada",
+  "Canadian Nuclear Safety Commission",
+  "Canadian Food Inspection Agency",
+  "Innovation, Science and Economic Development Canada",
+  "National Research Council Canada",
+  "Employment and Social Development Canada",
+  "Health Canada",
+  "Crown-Indigenous Relations and Northern Affairs Canada",
+  "Indigenous Services Canada",
+  "Parks Canada",
+  "Canadian Heritage",
+  "Financial Consumer Agency of Canada",
[…]

### 2024-11-29 scripts/02-test_simulated_data.R

@@ -1,89 +1,83 @@
 #### Preamble ####
-# Purpose: Tests the structure and validity of the simulated Australian 
-  #electoral divisions dataset.
-# Author: Rohan Alexander
-# Date: 26 September 2024
-# Contact: rohan.alexander@utoronto.ca
+# Purpose: Tests the structure and validity of the simulated Procurment data
+# Author: Robert Ford
+# Date: 25 November 2024
+# Contact: robert.ford@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: 
   # - The `tidyverse` package must be installed and loaded
   # - 00-simulate_data.R must have been run
-# Any other information needed? Make sure you are in the `starter_folder` rproj
+# Any other information? Make sure you are in the `canadian_gov_procurement` rproj
 
 
 #### Workspace setup ####
 library(tidyverse)
 
-analysis_data <- read_csv("data/00-simulated_data/simulated_data.csv")
-
-# Test if the data was successfully loaded
-if (exists("analysis_data")) {
-  message("Test Passed: The dataset was successfully loaded.")
-} else {
-  stop("Test Failed: The dataset could not be loaded.")
-}
+data <- read_csv("data/simulated_data/simulated_data.csv")
 
 
 #### Test data ####
 
-# Check if the dataset has 151 rows
-if (nrow(analysis_data) == 151) {
-  message("Test Passed: The dataset has 151 rows.")
-} else {
-  stop("Test Failed: The dataset does not have 151 rows.")
-}
+# Test that each column is the right data type
+test_that("Columns have the correct data types", {
+  expect_type(data$region, "character")
+  expect_type(data$contract, "character")
+  expect_type(data$buyer, "character")
+  expect_type(data$supplier, "character")
+  expect_type(data$amount, "double")  # or "integer" if amount is integer
+  expect_s3_class(data$award_date, "Date")
+  expect_s3_class(data$start_date, "Date")
+  expect_s3_class(data$end_date, "Date")
+  expect_type(data$link, "character")
+  expect_type(data$duration_days, "double")  # or "integer" if it's integer
+  expect_type(data$per_day, "double")
+})
 
-# Check if the dataset has 3 columns
-if (ncol(analysis_data) == 3) {
-  message("Test Passed: The dataset has 3 columns.")
-} else {
-  stop("Test Failed: The dataset does not have 3 columns.")
-}
+# Test that 'duration_days' is greater than 0
+test_that("duration_days > 0", {
+  expect_true(all(data$duration_days > 0))
+})
[…]

### 2024-11-29 paper/paper.qmd
More TODO's! And TODO Refs

@@
 ```
+TODO mention that detwiler makes up 88+% spending of the Space Agency
+
+TODO add pie chart showing top suppliers and other of National defence spending
+
+TODO add small contracts distribution chart
 
@@
 # Web Scraping
-These are step by step instructions on how to Web Scrape the IFJ procurement data page, because as of November 2024 there is no easy/automated way of downloading the 330 pages of 1000 rows worth of data.
 
-Note: It is important that docker, java, python and firefox are installed on your machine as RSelenium requires these packages.
-To download Selenium and create a docker image, open the docker application and search for the Selenium package.
+These are step by step instructions on how to Web Scrape the IFJ procurement data page, because as of November 2024 there is no easy/automated way of downloading the 330 pages of data.
+
+Note: It is important that docker, java, python and Firefox are installed on your machine as RSelenium requires these packages.
+To download Selenium and create a docker image, open the docker application and search for the Selenium package for the browser you wish to use, in this case Firefox.
 
@@
 4.
-The for-loop simply runs "click download -> click next page" process 329 times, however IJF may add more pages so modify total_pages to scrape the number of pages IJF has.
+The for-loop simply runs "click download -> click next page" process 329 times, however IJF may add more pages so modify total_pages to scrape the number of pages IJF is offering.
 In the event of an error mid-way through scraping take note of which page failed to download (page that you are on is printed to the console) and simply set the start of the for-loop.
@@
 
-5. Done! Enjoy!
+5. Done!
 
@@
 # References
+
+TODO Ref RSelenuim

### 2024-11-29 paper/paper.qmd
TODO Refs
added top-buyers chart

@@
 TODO Talk about top 5 buyers and other
-```{r fig-top-buyers}
+
+```{r fig-pie-buyers}
 #| message: false
@@
 #| warning: false
-#| fig-cap: template
+#| fig-cap: Top 5 Buyers by Total Amount Spent and Others (Pie Chart)
 #| fig-align: center
-
-#TODO color each buyer with the correct color
 
@@
   summarise(total_amount = sum(total_amount, na.rm = TRUE)) %>%
-  arrange(desc(total_amount))
+  arrange(desc(total_amount)) %>%
+  mutate(percentage = total_amount / sum(total_amount) * 100)
 
-# Plot the bar chart
-ggplot(buyer_totals_grouped, aes(x = reorder(buyer_group, total_amount), y = total_amount)) +
-  geom_bar(stat = "identity", fill = "skyblue") +
-  coord_flip() +
-  labs(title = "Top 5 Buyers by Total Amount Spent and Others",
-       x = "Buyer",
-       y = "Total Amount Spent (CAD)") +
-  scale_y_continuous(labels = comma) +
-  theme_minimal()
+# TODO: Assign specific colors to each buyer group
+# Example color assignments (replace with desired colors)
+buyer_colors <- c(
+  "National Defence" = "#1f77b4",
+  "Public Services and Procurement Canada" = "#ff7f0e",
+  "Shared Services Canada" = "#2ca02c",
+  "Public Health Agency of Canada" = "#d62728",
+  "Fisheries and Oceans Canada" = "#9467bd",
+  "Other"  = "#8c564b"
+)
 
+# Plot the pie chart
+ggplot(buyer_totals_grouped, aes(x = "", y = total_amount, fill = buyer_group)) +
+  geom_bar(stat = "identity", width = 1, color = "white") +
+  coord_polar(theta = "y") +
+  labs(
+    title = "Top 5 Buyers by Total Amount Spent and Others",
+    fill = "Buyer"
+  ) +
+  scale_fill_manual(values = buyer_colors) +
+  theme_void() + # Removes background, grid, and numeric labels
+  theme(
+    legend.position = "right",
+    plot.title = element_text(hjust = 0.5, size = 16, face = "bold")
+  ) +
+  geom_text(aes(label = paste0(round(percentage, 1), "%")),
+            position = position_stack(vjust = 0.5),
+            color = "white",
+            size = 4)
 ```
@@
 TODO add spending by selected buyers graph
+```{r fig-top-buyers}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-cap: template
+#| fig-align: center
+
+# Filter out records from 2019
+df_filtered <- df %>%
+  filter(format(award_date, "%Y") != "2019")
+
+# Extract the year from award_date
+df_filtered$year <- format(df_filtered$award_date, "%Y")
+
+# Define the selected buyers
+selected_buyers <- c(
+  "National Defence",
+  "Public Services and Procurement Canada",
+  "Shared Services Canada",
+  "Public Health Agency of Canada",
+  "Fisheries and Oceans Canada"
+)
+
+# Add a new column for buyer categories
+df_filtered <- df_filtered %>%
+  mutate(buyer_category = ifelse(buyer %in% selected_buyers, buyer, "Other"))
+
+# Summarize total spending per year for each buyer category
+grouped_df <- df_filtered %>%
+  group_by(year, buyer_category) %>%
+  summarize(total_spending = sum(amount, na.rm = TRUE)) %>%
+  ungroup()
+
+# Ensure the order of the groups matches the specified order, with "Other" last
+grouped_df$buyer_category <- factor(
+  grouped_df$buyer_category,
+  levels = c(selected_buyers, "Other")
+)
+
+# Scale the total_spending to billions
+grouped_df <- grouped_df %>%
+  mutate(total_spending_billion = total_spending / 1e9)
+
+# Create the grouped bar chart
+ggplot(data = grouped_df, aes(x = year, y = total_spending_billion, fill = buyer_category)) +
+  geom_bar(
+    stat = "identity",
+    position = position_dodge(width = 0.9), # Adjust dodge width
+    width = 0.8 # Adjust bar width
+  ) +
+  scale_fill_manual(
+    values = c(
+      "National Defence" = "steelblue",
+      "Public Services and Procurement Canada" = "orange",
+      "Shared Services Canada" = "green",
+      "Public Health Agency of Canada" = "purple",
+      "Fisheries and Oceans Canada" = "red",
+      "Other" = "gray"
+    )
+  ) +
+  scale_y_continuous(
+    labels = comma_format(accuracy = 1), # Formats y-axis labels without decimal places
+    breaks = pretty_breaks(n = 5), # Ensures approximately 5 breaks like 5, 10, 15
+    expand = expansion(mult = c(0, 0.05)) # Adds a small space above the highest bar
+  ) +
+  labs(
+    title = "Spending by Selected and Other Buyers Per Year",
+    subtitle = "Amounts are in Billions of CAD",
+    x = "Year",
+    y = "Total Spending (Billions)",
+    fill = "Buyer",
+    caption = "."
+  ) +
+  theme_minimal() +
+  theme(
+    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
+    plot.subtitle = element_text(hjust = 0.5, size = 12),
+    axis.text.x = element_text(angle = 45, hjust = 1), # Rotate x-axis labels for readability
+    axis.text = element_text(size = 12),
+    axis.title = element_text(size = 14),
+    legend.title = element_text(size = 14),
+    legend.text = element_text(size = 12),
+    plot.caption = element_text(hjust = 0.5, size = 10, face = "italic") # Centered caption
+  )
+
+```
 
@@
 
-TODO Ref RSelenuim
+TODO Ref:
+
+RSelenuim
+VGAM
+stringr
+dplyr
+tidyverse
+zoo
+kableExtra
+knitr
+scales
+lubridate
+ggplot2
+rvest
+readr
+testthat

### 2024-11-29 other/EDA/EDA.qmd
fixed up top-buyers chart

@@
 ```{r}
+library(ggplot2)
+library(dplyr)
+library(scales) # For formatting y-axis labels
+
+library(ggplot2)
+library(dplyr)
+library(scales) # For formatting y-axis labels
+
+# Filter out records from 2019
 df_filtered <- df %>%
@@
 
+# Extract the year from award_date
 df_filtered$year <- format(df_filtered$award_date, "%Y")
@@
 
+# Scale the total_spending to billions
+grouped_df <- grouped_df %>%
+  mutate(total_spending_billion = total_spending / 1e9)
+
 # Create the grouped bar chart
-ggplot(data = grouped_df, aes(x = year, y = total_spending, fill = buyer_category)) +
-  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
+ggplot(data = grouped_df, aes(x = year, y = total_spending_billion, fill = buyer_category)) +
+  geom_bar(
+    stat = "identity",
+    position = position_dodge(width = 0.9), # Adjust dodge width
+    width = 0.8 # Adjust bar width
+  ) +
   scale_fill_manual(
@@
   ) +
+  scale_y_continuous(
+    labels = comma_format(accuracy = 1), # Formats y-axis labels without decimal places
+    breaks = pretty_breaks(n = 5), # Ensures approximately 5 breaks like 5, 10, 15
+    expand = expansion(mult = c(0, 0.05)) # Adds a small space above the highest bar
+  ) +
   labs(
     title = "Spending by Selected and Other Buyers Per Year",
+    subtitle = "Amounts are in Billions of CAD",
     x = "Year",
-    y = "Total Spending",
-    fill = "Buyer"
+    y = "Total Spending (Billions)",
+    fill = "Buyer",
+    caption = "."
   ) +
   theme_minimal() +
-  theme(plot.title = element_text(hjust = 0.5))
+  theme(
+    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
+    plot.subtitle = element_text(hjust = 0.5, size = 12),
+    axis.text.x = element_text(angle = 45, hjust = 1), # Rotate x-axis labels for readability
+    axis.text = element_text(size = 12),
+    axis.title = element_text(size = 14),
+    legend.title = element_text(size = 14),
+    legend.text = element_text(size = 12),
+    plot.caption = element_text(hjust = 0.5, size = 10, face = "italic") # Centered caption
+  )
 

### 2024-11-29 paper/paper.qmd
added distribution chart

@@
 
-## Shape
+## Measurement
+
+TODO How IJF collected all this data
+
+## Analysis
 
@@
 
+## Distribution
+
 TODO add small contracts distribution chart
 
-## Measurement
+```{r fig-distibution}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-cap: template
+#| fig-align: center
 
-TODO How IJF collected all this data
+df_filtered <- df %>%
+  filter(amount <= 100000)
 
-T
+# Optional: Check how many rows are left
+num_less_than_100k <- nrow(df_filtered)
+total_contracts <- nrow(df)
+
+percent_less_than <- num_less_than_100k / total_contracts
+
+print(paste0("Percentage of contracts <= 100,000 CAD: ", round(percent_less_than * 100, 2), "%"))
+
+# Step 2: Create bins spanning every 10,000 CAD up to 100,000
+bin_width <- 10000  # 10,000 CAD
+
+# Create a sequence of break points from 0 to 100,000
+breaks <- seq(0, 100000, by = bin_width)
+
+# Assign each amount to a bin
+df_filtered <- df_filtered %>%
+  mutate(amount_bin = cut(amount, breaks = breaks, right = FALSE, include.lowest = TRUE))
+
+# Step 3: Count the number of contracts in each bin
+contract_counts_df <- df_filtered %>%
+  group_by(amount_bin) %>%
+  summarize(Contract_Count = n()) %>%
+  ungroup()
+
+# Ensure that all bins are represented, even if count is zero
+all_bins <- levels(df_filtered$amount_bin)
+contract_counts_df <- contract_counts_df %>%
+  right_join(data.frame(amount_bin = all_bins), by = "amount_bin") %>%
+  replace_na(list(Contract_Count = 0)) %>%
+  arrange(amount_bin)
+
+# Step 4: Create readable labels for x-axis in thousands
+contract_counts_df <- contract_counts_df %>%
+  mutate(
+    # Extract lower and upper bounds from amount_bin
+    lower = as.numeric(gsub("\\[|\\)", "", sapply(strsplit(as.character(amount_bin), ","), `[`, 1))),
+    upper = as.numeric(gsub("\\[|\\)|\\]", "", sapply(strsplit(as.character(amount_bin), ","), `[`, 2))),
+    amount_label = paste0(lower / 1000, "-", upper / 1000, "")
+  )
+
+# Step 5: Scale the y-axis counts to thousands if applicable
+max_count <- max(contract_counts_df$Contract_Count)
+if (max_count >= 1000) {
+  contract_counts_df <- contract_counts_df %>%
+    mutate(Contract_Count_k = Contract_Count / 1000)
+  y_label_unit <- "Thousands"
+} else {
+  contract_counts_df <- contract_counts_df %>%
+    mutate(Contract_Count_k = Contract_Count)  # No scaling
+  y_label_unit <- "Number of Contracts"
+}
+
+# Step 6: Create the ggplot2 bar chart
+ggplot(contract_counts_df, aes(x = amount_label, y = Contract_Count_k)) +
+  geom_bar(stat = "identity", fill = "skyblue", color = "black") +
+  scale_y_continuous(
+    labels = if (max_count >= 1000) {
+      label_number(scale = 1, accuracy = 1)
+    } else {
+      label_number(accuracy = 1)
+    },
+    expand = expansion(mult = c(0, 0.05))
+  ) +
+  labs(
+    title = "Number of Contracts per 10,000 CAD Award Amount Range",
+    x = "Award Amount (Thousands)",
+    y = paste0("Number of Contracts ", if (max_count >= 1000) "(Thousands)" else ""),
+    caption = "Data includes contracts <= 100,000 CAD."
+  ) +
+  theme_minimal() +
+  theme(
+    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
+    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
+    axis.text.y = element_text(size = 12),
+    axis.title = element_text(size = 14),
+    plot.caption = element_text(hjust = 0.5, size = 10, face = "italic")
+  )
+
+```
+
+TODO Bar chart that shows how much money is awarded in small contracts (<=200k) versus how much money is awarded big contracts
 

### 2024-11-29 scripts/05-exploratory_data_analysis.R
fixed distribution chart

@@ -32,7 +32,7 @@ u_buyers <- data.frame(unique(df$buyer))
 
 u_supps <- data.frame(unique(df$supplier))
 
-View(u_supps)
+View(u_buyers)
 
 public_health <- df[df$buyer == "Public Health Agency of Canada", ]
 
@@ -58,15 +58,30 @@ contracts_expiring <- df %>%
   ) %>%
   ungroup()
 
-space_total <- df %>%
+dept_total <- df %>%
   filter(buyer == "Canadian Space Agency") %>%
   summarize(total_amount = sum(amount, na.rm = TRUE)) %>%
   pull(total_amount)
 
-print(space_total)
+supp_total <- df %>%
+  filter(processed_supplier == "dettwiler") %>%
+  summarize(total_amount = sum(amount, na.rm = TRUE)) %>%
+  pull(total_amount)
 
-print(2199059594/ space_total)
+total_spend <- df %>%
+  summarize(total_amount = sum(amount, na.rm = TRUE)) %>%
+  pull(total_amount)
 
+print(supp_total / dept_total)
+
+print(dept_total)
+
+print(dept_total/ total_spend)
+
+
+war_planes <- 1792979385 + 11219370646 + 3718137594 + 4482200974 + 4322729867
+
+print(war_planes)
 
 View(contracts_expiring)
 View(public_health)

### 2024-11-29 other/EDA/EDA.qmd
fixed distribution chart

@@
 df_filtered <- df %>%
-  filter(amount <= 200000)
+  filter(amount <= 100000)
 
@@
 )
+```
+```{r}
+# Step 1: Filter the data for contracts <= 100,000 CAD
+df_filtered <- df %>%
+  filter(amount <= 100000)
+
+# Optional: Check how many rows are left
+num_less_than_100k <- nrow(df_filtered)
+total_contracts <- nrow(df)
+
+percent_less_than <- num_less_than_100k / total_contracts
+
+print(paste0("Percentage of contracts <= 100,000 CAD: ", round(percent_less_than * 100, 2), "%"))
+
+# Step 2: Create bins spanning every 10,000 CAD up to 100,000
+bin_width <- 10000  # 10,000 CAD
+
+# Create a sequence of break points from 0 to 100,000
+breaks <- seq(0, 100000, by = bin_width)
+
+# Assign each amount to a bin
+df_filtered <- df_filtered %>%
+  mutate(amount_bin = cut(amount, breaks = breaks, right = FALSE, include.lowest = TRUE))
+
+# Step 3: Count the number of contracts in each bin
+contract_counts_df <- df_filtered %>%
+  group_by(amount_bin) %>%
+  summarize(Contract_Count = n()) %>%
+  ungroup()
+
+# Ensure that all bins are represented, even if count is zero
+all_bins <- levels(df_filtered$amount_bin)
+contract_counts_df <- contract_counts_df %>%
+  right_join(data.frame(amount_bin = all_bins), by = "amount_bin") %>%
+  replace_na(list(Contract_Count = 0)) %>%
+  arrange(amount_bin)
+
+# Step 4: Create readable labels for x-axis in thousands
+contract_counts_df <- contract_counts_df %>%
+  mutate(
+    # Extract lower and upper bounds from amount_bin
+    lower = as.numeric(gsub("\\[|\\)", "", sapply(strsplit(as.character(amount_bin), ","), `[`, 1))),
+    upper = as.numeric(gsub("\\[|\\)|\\]", "", sapply(strsplit(as.character(amount_bin), ","), `[`, 2))),
+    amount_label = paste0(lower / 1000, "-", upper / 1000, "")
+  )
+
+# Step 5: Scale the y-axis counts to thousands if applicable
+max_count <- max(contract_counts_df$Contract_Count)
+if (max_count >= 1000) {
+  contract_counts_df <- contract_counts_df %>%
+    mutate(Contract_Count_k = Contract_Count / 1000)
+  y_label_unit <- "Thousands"
+} else {
+  contract_counts_df <- contract_counts_df %>%
+    mutate(Contract_Count_k = Contract_Count)  # No scaling
+  y_label_unit <- "Number of Contracts"
+}
+
+# Step 6: Create the ggplot2 bar chart
+ggplot(contract_counts_df, aes(x = amount_label, y = Contract_Count_k)) +
+  geom_bar(stat = "identity", fill = "skyblue", color = "black") +
+  scale_y_continuous(
+    labels = if (max_count >= 1000) {
+      label_number(scale = 1, accuracy = 1)
+    } else {
+      label_number(accuracy = 1)
+    },
+    expand = expansion(mult = c(0, 0.05))
+  ) +
+  labs(
+    title = "Number of Contracts per 10,000 CAD Award Amount Range",
+    x = "Award Amount (Thousands)",
+    y = paste0("Number of Contracts ", if (max_count >= 1000) "(Thousands)" else ""),
+    caption = "Data includes contracts <= 100,000 CAD."
+  ) +
+  theme_minimal() +
+  theme(
+    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
+    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
+    axis.text.y = element_text(size = 12),
+    axis.title = element_text(size = 14),
+    plot.caption = element_text(hjust = 0.5, size = 10, face = "italic")
+  )
+
 ```

### 2024-11-29 scripts/06-model.R
added model diagnostics

@@ -7,7 +7,8 @@
 # Pre-requisites: 
 # - The `tidyverse` package must be installed and loaded
 # - The `dplyr` package must be installed and loaded
-# - 
+# - The `car` package must be installed and loaded
+# - The `caret` package must be installed and loaded
 # Other information? Make sure you are in the `canadian_gov_procurement` rproj
 
 
@@ -15,63 +16,125 @@
 #### Workspace setup ####
 library(tidyverse)
 library(dplyr)
+library(car)
+library(caret)
 
 #### Read data ####
 df <- read_csv("data/analysis_data/cleaned_data.csv")
 
-#TODO Simplify Down to one model
 
 ### Model data ####
-# Define the specific buyers
-specific_buyers <- c(
-  "National Defence",
-  "Public Services and Procurement Canada",
-  "Shared Services Canada",
-  "Public Health Agency of Canada",
-  "Fisheries and Oceans Canada"
-)
-
-# Create a new column 'buyer_group' categorizing each buyer
-df <- df %>%
-  mutate(
-    buyer_group = ifelse(buyer %in% specific_buyers, buyer, "Other")
-  )
-
 df <- df %>%
   mutate(
     award_date_numeric = as.numeric(award_date)
   )
 
-# Split the dataframe into a list of dataframes by 'buyer_group'
-buyer_groups <- split(df, df$buyer_group)
-
-# Initialize an empty list to store the models
-models <- list()
-
-# Loop through each buyer group and fit the linear model
-for (buyer in names(buyer_groups)) {
-  # Subset the data for the current buyer group
-  data_subset <- buyer_groups[[buyer]]
-  
-  # Check if there are enough data points to fit a model
-  if(nrow(data_subset) > 1) {  # At least two points needed for regression
-    # Fit the linear model: amount ~ award_date_numeric + duration_days
-    model <- lm(amount ~ award_date_numeric + duration_days, data = data_subset)
-    
-    # Store the model in the list with the buyer group name as the key
-    models[[buyer]] <- model
-  } else {
-    warning(paste("Not enough data to fit a model for:", buyer))
-  }
-}
 
 
-model_summaries <- lapply(models, summary)
-print(model_summaries)
+model <- lm(amount ~ award_date_numeric + duration_days, data = df)
 
+mil_model <- lm(amount ~ award_date_numeric + duration_days, data = df %>% filter(buyer == "National Defence"))
+
+model_summary <- summary(model)
+mil_model_summ <- summary(mil_model)
+
[…]

### 2024-11-30 scripts/01-download_data.R
added correct appendix

@@ -11,7 +11,7 @@
 # - The `dplyr` package must be installed and loaded
 # - The `readr` package must be installed and loaded
 
-# See appendix of paper for detailed step by step instructions
+# See appendix A of paper for detailed step by step instructions
 # Be sure to fill in all instances of USER_PATH with where ever you have stored this rproj on your machine
 
 # -! IMPORTANT !- You must FIRST run this code in your terminal which is cd in this file:

### 2024-11-30 paper/paper.qmd
updated appendix A

@@
   )
-
 ```
@@
 After restarting the loop be sure to verify that the next downloaded page is the correct one, otherwise you will have to restart the process from step 1.
+While it may be tempting to reduce the Sys.sleep times to download faster (11 seconds delay per loop and 329 loops that's about 1 hour!), I found reducing the time cause errors as the pages were slow to load and thus ruining the process.
 
@@
 TODO Removal of negative values, adding 1 day to contract lengths of 0, see cleaning script
+
+# Proccessing Supplier
 

### 2024-11-30 other/EDA/EDA.qmd
fixed small and big contracts bar chart

@@
 
+print(num_less_than_100k)
+
 percent_less_than <- num_less_than_100k / total_contracts
@@
 ```{r}
+# Step 1: Create a new column 'contract_size' categorizing contracts
+df <- df %>%
+  mutate(
+    contract_size = ifelse(amount <= 100000, "Small Contracts (≤ 100k)", "Big Contracts (> 100k)")
+  )
 
-df_filtered <- df %>%
-  filter(buyer == "National Defence", amount <= 200000)
-
-# Optional: Check how many rows are left
-num_less_than_200k <- nrow(df_filtered)
-total_contracts <- nrow(df)
-
-percent_less_than <- num_less_than_200k / total_contracts
-
-print(percent_less_than)
-
-# Step 3: Create bins spanning every 10,000 CAD up to 200,000
-# Define the bin width
-bin_width <- 10000  # 10,000 CAD
-
-# Create a sequence of break points from 0 to 200,000
-breaks <- seq(0, 200000, by = bin_width)
-
-# To ensure that the upper limit includes 200,000, set 'right = FALSE'
-# This means intervals are [0,10000), [10000,20000), ..., [190000,200000]
-df_filtered <- df_filtered %>%
-  mutate(amount_bin = cut(amount, breaks = breaks, right = FALSE, include.lowest = TRUE))
-
-# Step 4: Count the number of contracts in each bin
-contract_counts_df <- df_filtered %>%
-  group_by(amount_bin) %>%
-  summarize(Contract_Count = n()) %>%
+# Summarize total amount spent per contract size category
+total_spent_df <- df %>%
+  group_by(contract_size) %>%
+  summarize(
+    total_amount_spent = sum(amount, na.rm = TRUE)
+  ) %>%
   ungroup()
 
-# View the binned counts
-print(contract_counts_df)
+# Convert the total amount to billions of CAD
+total_spent_df <- total_spent_df %>%
+  mutate(total_amount_billion = total_amount_spent / 1e9)
 
-contract_counts_df$amount_bin <- as.character(contract_counts_df$amount_bin)
+# Optional: Display the summarized data
+print(total_spent_df)
 
-# Create the bar plot
-barplot(
-  height = contract_counts_df$Contract_Count,
-  names.arg = contract_counts_df$amount_bin,
-  las = 2,  # Rotate x-axis labels to perpendicular
-  col = "skyblue",
-  border = "black",
-  main = "Number of Contracts per 10,000 CAD Award Amount Range",
-  xlab = "Award Amount (CAD)",
-  ylab = "Number of Contracts",
-  cex.names = 0.8  # Adjust label size for readability
+# Create the bar chart
+ggplot(total_spent_df, aes(x = contract_size, y = total_amount_billion, fill = contract_size)) +
+  geom_bar(stat = "identity", width = 0.6) +
+  geom_text(aes(label = paste0(round(total_amount_billion, 2), "B")), 
+            vjust = -0.5, 
+            size = 5) +  # Adds labels above the bars
+  scale_fill_manual(
+    values = c("Small Contracts (≤ 100k)" = "steelblue", "Big Contracts (> 100k)" = "tomato")
+  ) +
+  scale_y_continuous(
+    labels = scales::number_format(scale = 1, suffix = "B"),  # Displays y-axis in billions
+    expand = expansion(mult = c(0, 0.1))  # Adds space above the bars for labels
+  ) +
+  labs(
+    title = "Total Amount Spent on Contracts",
+    subtitle = "Comparison between Small (≤ 100k CAD) and Big (> 100k CAD) Contracts",
+    x = "Contract Size",
+    y = "Total Amount Spent (Billions of CAD)"
+  ) +
+  theme_minimal() +
+  theme(
+    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
+    plot.subtitle = element_text(hjust = 0.5, size = 14),
+    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 14),
+    axis.text.y = element_text(size = 12),
+    axis.title = element_text(size = 16),
+    legend.position = "none",  # Removes the legend as it's redundant
+    plot.caption = element_text(hjust = 0.5, size = 10, face = "italic")
   )
@@
 ```
+
+```{r}
+# Load necessary libraries
+library(ggplot2)
+library(dplyr)
+library(scales)
+
+# Assuming 'df' is your data frame and has a column 'amount'
+
+# Step 1: Categorize contracts based on amount
+df <- df %>%
+  mutate(
+    contract_size = ifelse(amount <= 100000, "Small Contracts (≤ 100k)", "Big Contracts (> 100k)")
+  )
+
+# Step 2: Ensure the order of contract_size (Small on left, Big on right)
+df$contract_size <- factor(df$contract_size, levels = c("Small Contracts (≤ 100k)", "Big Contracts (> 100k)"))
+
+# Step 3: Summarize total amount spent per contract size category
+total_spent_df <- df %>%
+  group_by(contract_size) %>%
+  summarize(
+    total_amount_spent = sum(amount, na.rm = TRUE)
+  ) %>%
+  ungroup()
+
+# Step 4: Convert total amount to billions of CAD
+total_spent_df <- total_spent_df %>%
+  mutate(total_amount_billion = total_amount_spent / 1e9)
+
+# Optional: Display the summarized data
+print(total_spent_df)
+
+# Step 5: Create the ggplot2 bar chart
+ggplot(total_spent_df, aes(x = contract_size, y = total_amount_billion, fill = contract_size)) +
+  geom_bar(stat = "identity", width = 0.6) +
+  geom_text(aes(label = paste0(round(total_amount_billion, 2), "B")), 
+            vjust = -0.5, 
+            size = 5) +
+  scale_fill_manual(
+    values = c("Small Contracts (≤ 100k)" = "steelblue", "Big Contracts (> 100k)" = "tomato")
+  ) +
+  scale_y_continuous(
+    labels = scales::number_format(scale = 1, suffix = "B"),  # Y-axis in billions
+    expand = expansion(mult = c(0, 0.1))  # Adds space above the bars for labels
+  ) +
+  labs(
+    title = "Total Amount Spent on Contracts",
+    subtitle = "Comparison between Small (≤ 100k CAD) and Big (> 100k CAD) Contracts",
+    x = "Contract Size",
+    y = "Total Amount Spent (Billions of CAD)"
+  ) +
+  theme_minimal() +
+  theme(
+    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
+    plot.subtitle = element_text(hjust = 0.5, size = 14),
+    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 14),
+    axis.text.y = element_text(size = 12),
+    axis.title = element_text(size = 16),
+    legend.position = "none",  # Removes the legend as it's redundant
+    plot.caption = element_text(hjust = 0.5, size = 10, face = "italic"),
+    panel.grid.major.y = element_line(color = "grey80", linetype = "dashed")  # Adds dashed grid lines
+  )
+
+```

### 2024-11-30 paper/paper.qmd
added small and big contracts graph

@@
 
-df_table <- df %>% select(-contract, -start_date, -end_date)
+df_table <- df %>% select(-contract, -start_date, -end_date, -link)
 
@@
 
+```{r fig-small-big}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-cap: template
+#| fig-align: center
+
+# Step 1: Create a new column 'contract_size' categorizing contracts
+df <- df %>%
+  mutate(
+    contract_size = ifelse(amount <= 100000, "Small Contracts (≤ 100k)", "Big Contracts (> 100k)")
+  )
+
+# Summarize total amount spent per contract size category
+total_spent_df <- df %>%
+  group_by(contract_size) %>%
+  summarize(
+    total_amount_spent = sum(amount, na.rm = TRUE)
+  ) %>%
+  ungroup()
+
+# Convert the total amount to billions of CAD
+total_spent_df <- total_spent_df %>%
+  mutate(total_amount_billion = total_amount_spent / 1e9)
+
+# Optional: Display the summarized data
+print(total_spent_df)
+
+# Create the bar chart
+ggplot(total_spent_df, aes(x = contract_size, y = total_amount_billion, fill = contract_size)) +
+  geom_bar(stat = "identity", width = 0.6) +
+  geom_text(aes(label = paste0(round(total_amount_billion, 2), "B")), 
+            vjust = -0.5, 
+            size = 5) +  # Adds labels above the bars
+  scale_fill_manual(
+    values = c("Small Contracts (≤ 100k)" = "steelblue", "Big Contracts (> 100k)" = "tomato")
+  ) +
+  scale_y_continuous(
+    labels = scales::number_format(scale = 1, suffix = "B"),  # Displays y-axis in billions
+    expand = expansion(mult = c(0, 0.1))  # Adds space above the bars for labels
+  ) +
+  labs(
+    title = "Total Amount Spent on Contracts",
+    subtitle = "Comparison between Small (≤ 100k CAD) and Big (> 100k CAD) Contracts",
+    x = "Contract Size",
+    y = "Total Amount Spent (Billions of CAD)"
+  ) +
+  theme_minimal() +
+  theme(
+    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
+    plot.subtitle = element_text(hjust = 0.5, size = 14),
+    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 14),
+    axis.text.y = element_text(size = 12),
+    axis.title = element_text(size = 16),
+    legend.position = "none",  # Removes the legend as it's redundant
+    plot.caption = element_text(hjust = 0.5, size = 10, face = "italic")
+  )
+
+```
+
 # Model

### 2024-11-30 scripts/05-exploratory_data_analysis.R
finding how many one doller contracts there are

@@ -24,6 +24,12 @@ u_dsupps <- df %>%
 # Display the unique suppliers
 View(u_dsupps)
 
+one_dollar <- df %>%
+  filter(amount == 1) %>%
+  summarize(number_of_one_dollar_contracts = n())
+
+print(one_dollar)
+
 u_contracts <- data.frame(unique(df$contract))
 
 max.print(u_contracts)

### 2024-12-02 paper/paper.qmd
added model results table + TODO proccessed supplier + Introduction section

@@
 library(zoo)
+library(broom)
 
@@
 aggregated_df <- read_csv(here::here("data/analysis_data/top_ten_suppliers.csv"))
-models <- read_csv(here::here("models/buyer_models.rds"))
+model <- read_rds(here::here("models/gen_model.rds"))
+mil_model <- read_rds(here::here("models/mil_model.rds"))
 
-View(df)
+aView(df)
 ```
@@
 
-TODO Broader context of Canadian spending, talk about huge new military spending package announced this year
+In recent years, Canada's federal spending landscape has undergone significant transformations, particularly with a marked increase in military expenditures.
+The government has announced substantial new military spending packages, [REF]
 
-TODO What are procurement, how are they awarded, what are Buyers and Suppliers
+Federal procurements involve the acquisition of goods and services by government departments, Buyers, from companies, Suppliers.
+These procurements are essential for a wide range of government functions from; acquiring defense equipment, building infrastructure, procuring healthcare supplies, technological services and engineering consultants for example.
+This paper will analyse federal procurement data to dissect government procurements over the last four years.
 
-TODO Gap that needs to be filled; What is the Government buying from private companies and how is their consumption changing over time
+The primary questions this paper aims to address are: What is the government buying? Who are the primary suppliers? How are spending patterns shifting across different departments and over time?
 
-TODO What was done
+I collected and processed detailed data on over 301,000 contracts awarded since January 2020.
+The dataset includes variables such as contract amounts, durations, award dates, Buyers, and Suppliers.
+We derived additional metrics to facilitate our analysis, such as Baseline Days (the number of days since January 1, 2020) and Duration Days (the length of each contract), and standardized supplier names to ensure consistency.
 
-TODO What was found
+Some significant findings presented in this paper include: military procurement spending has surged in the past few years.
+Where the Department of National Defence now dominates federal procurement spending, with expenditures exceeding $56 billion since 2020.
+This increase in spending is in part thanks to large contracts, including a $11.2 billion contract to SkyAlyne for fighter pilot training—the largest in our dataset—and multi-billion-dollar contracts for fighter jets and naval vessels.
 
-TODO Why is it important
+While  other government departments have experienced a notable decrease in procurement spending.
+For instance, the Public Health Agency of Canada's spending, which was substantial in 2020 due to the COVID-19 pandemic, has declined significantly in subsequent years.
 
-TODO Paper layout, description of each section with jump links
+These findings are important for several reasons.
+They provide critical insights into how government spending priorities are shifting, which has implications for economic policy, industry stakeholders, and public transparency.
+Understanding the concentration of spending in large military contracts can inform discussions on national defense strategies, budget allocations, and the impact on domestic industries.
+
+The paper is structured as follows: [Section 2] details the data collection and processing methodologies, including the sources of procurement data and the steps taken to clean and standardize the dataset.
+Then I present a series of graphs to demonstrate government procurements from various angles.
+[Section 3] describes the linear regression models used to assess trends in contract amounts over time and contract duration.
+[Section 4] presents the results of the analysis, highlighting key trends in military and non-military procurement spending.
+[Section 5] discusses the limitations of the study, including data constraints and methodological considerations, and suggests areas for future research.
+Finally, Section 6 concludes with a summary of the findings and their implications for policymakers, industry stakeholders, and future studies.
+
+By providing a comprehensive analysis of federal procurement spending, this paper aims to fill the existing gap in understanding how government consumption of private sector goods and services is changing over time.
+The insights gained from this study can inform policy decisions, contribute to academic discourse, and enhance transparency regarding government spending practices.
 
@@
 
-Estimand?
+TODO Estimand?
 
@@
 
-df_table <- df %>% select(-contract, -start_date, -end_date, -link)
+df_table <- df %>% select(-contract, -start_date, -end_date, -link, -duration_days, -per_day, -processed_supplier)
 
@@
 
+```{r fig-results}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-cap: template
+#| fig-align: center
+
+# Tidy up the model results
+model_tidy <- model %>% tidy() %>% mutate(model = "Excluding Military Contracts")
+mil_model_tidy <- mil_model %>% tidy() %>% mutate(model = "Military Contracts Only")
+
+# Combine the results into one table
+combined_results <- bind_rows(model_tidy, mil_model_tidy)
+
+combined_results <- combined_results %>%
+  mutate(
+    term = case_when(
+      term == "(Intercept)" ~ "Intercept",
+      term == "baseline_days" ~ "Baseline Days",
+      term == "duration_days" ~ "Duration Days",
+      TRUE ~ term # Keep any other terms as-is
+    )
+  )
+
+# Format the table for Quarto using kableExtra
+combined_results %>%
+  select(model, term, estimate, std.error, statistic, p.value) %>%
+  rename(
+    Model = model,
+    Term = term,
+    Estimate = estimate,
+    `Std. Error` = std.error,
+    `t-Statistic` = statistic,
+    `p-Value` = p.value
+  ) %>%
+  kable(format = "html", digits = 3, align = "c", caption = "Model Results Summary") %>%
+  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
+
+```
+
 # Discussion
@@
 TODO processed_supplier
+Talk about what I did thourougly
+
+Talk about next steps.
+11th top supplier, and how there are still tons of useless words in there.
+Even how to approach I used may never work.
+Because at some point we will ban too many word and some suppliers with super generic names will be eliminated entirely.
+
+Propose alternate methods
 

### 2024-12-02 paper/paper.qmd
added text for Data Section

@@
 
-TODO Description of each variable head of table
+The data was collected using R and other packages [REF].
+A web scraping script was used to collect all Federal Award Data from the IJF [REF].
+The variables this paper focuses collected from the IJF are; Contract, Buyer, Supplier, Amount, Award Date, Start Date, and End Date.
+
+Contract: A few words detailing  a brief description of what the contract signed was for
+
+Buyer: The name of government department who awarded the contract
+
+Supplier: The name of the company that fulfilled the contract by agreeing to provide the good or service to the Buyer
+
+Amount: The value of the contract in Canadian Dollars
+
+Award Date: The day the Buyer awarded the contract to the supplier
+
+Start Date: When the contract is due to start
+
+End Date: When the contract is due to end
+
+I derived three additional variables using the IJF variables:
+
+Baseline Days: Number of days since 1st January 2020
+
+Duration Days: End Date subtracted from the Start Date to calculate the duration of the contract
+
+Processed Supplier: The name of suppliers regularized to allow for aggregation and comparison
+
+Baseline Days and Duration Days are used as predictors in the linear model, as Dates are not valid entries into a linear model.
+
+Processed Supplier is required as many companies have slightly different spellings, such as [“General Dynamics”, “GENERAL DYNAMICS”, “General Dynamics Mission Systems”, “GENERAL DYNAMICS MISSION SYSTEMS INC”].
+See Appendix TODO for how I processed and regularized the suppliers.
+
+Here is the head of the data that was used in this analysis for a sense of what these variables represent. 
+TODO Fix table
 ```{r fig-head}
@@
 
-TODO Brief talk about how cleaning was carried out
-
-TODO Description of variables that were created, duration_days, processed_supplier (brief, see appendix)
+The data was then cleaned.
+First I began by setting each variable to the correct data type.
+Then I removed any data point where Amount was negative or zero, because I am not sure what a negative amount means in this context and contracts that are not worth anything are not worth looking at.
+There were a number of contracts which started and ended on the same day, to simplify later calculation I simply added a day to their End Date, figuring that same day contracts are equivalent to one day contracts.
+Further data cleaning details are available in Appendix TODO.
+After cleaning I was left with 301,094 contracts spanning TODO to TODO.
 
@@
 
-TODO How IJF collected all this data
+The data is recorded and published through various government agencies, who use different accounting standards and procedures.
+The three sources IFJ used to collect the federal procurement data are; Canada Buys, Buy and Sell, and Proactive Disclosures (which makes up the vast majority of the procurement data).
+However, because the data used in this paper is pulled from the IFJ’s Federal Awards page this data comes exclusively from the Proactive Disclosures.
+The Treasury Board of Canada Secretariat is responsible for publishing these Proactive Disclosures under the Access to Information Act [REF].
 
@@
 
-TODO Talk about top 5 buyers and other
+This is a break down of total amount awarded by the top 5 biggest spending Buyers.
+The Other category consists the of the total awards given by the remaining 88 smallest Government Buyers.
+It is clear that National Defence dwarfs the procurement budget, spending over /$56 Billion on procurements since 2020.
 
@@
 
-TODO Military Spending Per year
+FIG show how National Defence spending has taken off in recent years, where is 2021 it had not even crested /$5 Billion, but in the last 3 years spending has exploded to /$15 Billion a year.
+
 ```{r fig-mil-spending}
@@
 
-TODO Public Services Spending
+Public Services and Procurements Canada is the clearinghouse for the entire federal government.
+This means they are responsible for managing the procurement process for buyers across the government [REF].
+It is important to note that the PSPC sometimes does not publish small contracts, less than /$40,000.
+PSPC spending over time shows while military spending increases the rest of the government is taking a hit.
 ```{r fig-public-spending}
@@
 
-TODO add spending by selected buyers graph
+This FIG shows the top five Buyers spending per year since 2020.
+In the first 2 years military spending held steady and did not dominate the budget by any means.
+But since then Defence spending exploded not just nominally but compared to the rest of the budget as well.
+Notably the Public Health Agency of Canada’s spending while strong in 2020, decreased to nothing over the following four years.
+The spending in 2020 is linked to purchases of medical equipment, so it is possible they have not needed to invest in new equipment since then.
+But this illustrates a boarder trend among all departments except National Defence, where their spending has steadily declined in the last 4 years.
+
 ```{r fig-top-buyers}
@@
 
-TODO Top Ten Suppliers
+To find out the firms that are benefiting the most through this procurement process we can look to FIG.
+Unsurprisingly, the biggest suppliers are primarily are awarded contracts by the military.
+Interestingly the top supplier, Skyalyne, was only awarded one huge 11.2 Billion contract for fighter pilot training services.
+This is the largest contract in the dataset.
+Then considering the largest contracts awarded in the last 3 years to; the F-35 program (Fighter Planes, 4.5 Billion), General Atomics (Airplanes, 1.7 Billion), CAE (Aviation Engineering and Training, 4.4 Billion), Airbus (Airplanes, 3.7 Billion), for a grand total of 25.5 Billion.
+These large single contracts devoted to advanced fighter jets and training pilots to fly them is part of the reason that military spending has exploded in the last 3 years.
+The military also paid Vancouver Shipyards 3.4 Billion in 2024 for Ships and Boat parts.
+
 ```{r fig-top-supps}
@@
 
-#TODO Fix colors and supplier names
+#TODO Fix colors and supplier names and order of supps (highest to lowest)
 
@@
 ```
-TODO mention that detwiler makes up 88+% spending of the Space Agency
 
-TODO add pie chart showing top suppliers and other of National defence spending
+Dettwiler is an engineering firm that provides the Canadian Space Agency with engineering and scientific consulting. Surprisingly 86.3% of the Space Agency’s procurement budget goes to Dettwiler.
 
@@
 
-TODO add small contracts distribution chart
+This graph represents 81.75% of all contracts, indicating that the vast majority of contracts awarded by the government are quite small.
+Furthermore, it seems that the number of contracts follows a Pareto distribution.
+There are 102 contracts that are worth $1, this maybe some legal formality.
 
@@
 
-TODO Bar chart that shows how much money is awarded in small contracts (<=200k) versus how much money is awarded big contracts
+Though most contracts are small contracts (less than $100,000), the sum of these 246,000 contracts is only worth a measly 7.34 Billion. The Skyalyne contract itself outweighs all the small contracts.
 

### 2024-12-02 paper/paper.qmd
Added model, results and discussion sections

@@
 
-TODO What the is the goal of our models
-
-Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].
+This paper will use a simple linear regression to better understand the how contract award amounts are affected over time compared to their lengths.
+To do this the response variable will be Amount, and the two dependent variables will be the number of day since January 1st 2020 (start of the data set) and the length of the contract.
+The coefficients of these variables will indicate the relative importance of when the contract was issues compared to the length of contracts on determining the size of contracts.
+I will create two models, one to measure the coefficients for all the contracts excluding military contracts and a model for solely military contracts.
+This way we can compare to see if military contracts are getting larger because they are relatively longer or because recently they have purchased more expensive equipment.
 
@@
 
-TODO define variables used and structure
+All model construction and diagnostics were conduction using R [REF], and the following packages:
 
-TODO Cite R and packages used to run model
+Tidyverse: For data manipulation and visualization [REF]
 
-### Model justification
+Dplyr: [REF]
 
-TODO Why I chose those variables
+Car: [REF]
 
-TODO Model Diagnostics
+Caret:[REF]
+
+Model Formula:
+
+TODO
+
+Variables:
+
+Amount:
+
+Baseline Days:
+
+Duration Days:
+
+## Limitations
+
+One significant limitation is the assumption of a linear relationship between the predictors and the contract amount.
+In reality, the relationship may be more complex or nonlinear, especially over an extended period where economic factors and policy changes could introduce fluctuations not captured by a simple linear trend.
+
+Another limitation is the potential omission of relevant variables that influence contract amounts.
+Factors such as inflation, changes in government procurement policies, economic cycles, or industry-specific trends can significantly affect contract values.
+The exclusion of these variables may lead to omitted variable bias, where the estimated effects of the included predictors are distorted because they are capturing the influence of missing factors.
+
+Additionally, the presence of outliers or extreme values, particularly in financial data like contract amounts, can disproportionately influence the model estimates and lead to misleading conclusions.
+
+This model treats all contracts as a  homogeneous group.
+But considering that contracts both for National Defence and general government awards vary widely in scope and cost that would suggest a linear model is not appropriate.
+A more segmented model may be appropriate.
+This may be difficult as dividing the data in to groups based on contract size would not be sufficient as there is still a wide variety of contracts that are of similar size.
+Possibly using the Contract descriptions to divide the data, but this would require more advanced Natural Language processing techniques and in many cases the descriptions are hardly descriptive.
+
+Moreover, the model assumes that the effect of each predictor is constant over time.
+If the influence of contract duration or time on contract amounts changes due to evolving market conditions or government strategies, the model's fixed coefficients would not reflect these dynamic relationships.
+
+## Diagnostics
+
+TODO
 
@@
 
-TODO Add model summary tables
+The model general contracts indicates that the contract amount tends to decrease over time.
+Because Baseline days is slightly negative.
+This value suggests that every day that passes on average the value of non-military contracts is going down by $61.62 a day.
+This decline could reflect shifts in government spending priorities away from non-military departments.
+Additionally, as expect Duration Days is positive of the contract amount .
+This means that longer contracts are more expensive for the government.
+
+In contrast, the model for military contracts reveals that the contract amount increases over time.
+The positive coefficient for Baseline days suggests an upward trend in military contract values since 2020.
+In this case the model suggests that the cost of military contracts is increasing by $852.50 every day.
+This is trend is due to the huge increases in military procurement we observed in FIG.
+
+The impact of contract duration on the contract amount is substantially larger in military contracts compared to non-military ones.
+Where military contracts cost $38,146 more per day of the length of the contract.
+This is again due largely by the 11 Billon Skyalyne contract, and the other large long contracts for training and fighter jets.
 
@@
 
-TODO What was done
+This study conducted an in-depth analysis of Canadian federal procurement data, with a focus on military contracts.
+The data was collected and processed to ensure accuracy and relevance.
+Using web scraping techniques, we gathered information from the Institute for Government's (IJF) Federal Award Data.
+The primary variables included contract descriptions, buyer and supplier names, contract amounts, and key dates such as award, start, and end dates.
+
+To facilitate the analysis, we derived additional variables: Baseline Days, representing the number of days since January 1st, 2020, and Duration Days, calculated by subtracting the start date from the end date of each contract.
+These transformations were necessary because date variables are not directly suitable for linear modeling.
+We also standardized supplier names through a Processed Supplier variable to ensure consistency, given the variations in how supplier names were recorded.
+
+Data cleaning was a critical step.
+Contracts with negative or zero amounts were excluded due to their ambiguous nature.
+Contracts that started and ended on the same day were adjusted by extending the end date by one day to simplify duration calculations.
+After these refinements, the dataset comprised 301,094 contracts spanning from [TODO: start date] to [TODO: end date].
+
+For the modeling phase, we employed a simple linear regression to explore how contract amounts are influenced by time and contract duration.
+Two separate models were constructed: one encompassing all contracts excluding military contracts, and another focusing solely on military contracts.
+This bifurcation allowed us to compare trends between military and non-military spending.
 
@@
 
-TODO What did we learn about Canada
-    Military spending has exploded in the last couple of years, and other departments have fallen significantly
-    The vast majority of contracts are smaller than 200,000, and the most common size of contract is between 10-20k
+The analysis revealed significant trends in Canada's federal procurement spending:
+
+Surge in Military Spending: Military spending has dramatically increased over the past couple of years.
+National Defence emerged as the dominant spender, allocating over $56 billion since 2020.
+While military spending was under $5 billion in 2021, it escalated to approximately $15 billion annually in the following years.
+This spike is attributed to substantial contracts for advanced military equipment and services, such as the $11.2 billion contract awarded to SkyAlyne for fighter pilot training—the largest in the dataset.
+Other notable contracts include multi-billion-dollar deals with Lockheed Martin, General Atomics, CAE, and Airbus for fighter planes and aviation services.
+
+Decline in Other Departments' Spending: In contrast, other government departments have experienced significant reductions in procurement spending.
+For instance, the Public Health Agency of Canada's spending, which was robust in 2020 due to pandemic-related expenditures, diminished to negligible levels in the subsequent years.
+
+Prevalence of Small Contracts: The majority of government contracts are relatively small.
+Approximately 81.75% of all contracts are valued under \$100,000, with the most common contract size ranging between \$10,000 and \$20,000.
+This pattern indicates a governmental preference for smaller contracts, possibly for regulatory reasons.
+There may be a rule that if a contract is below 20000 then it requires less scrutiny to be passed, in effect encouraging department to award more smaller contracts.
+Though it does not seem to be out of control as only TODO% of all spending in last 4 years is concentrated in contracts less than 100000.
 
@@
 
-TODO Model was weak, need more data to improve model
-    Analysis on the Suppliers Names needs to be more rigorous
+While the study provides valuable insights, several limitations must be acknowledged:
+
+Supplier Name Analysis: The methodology for standardizing supplier names, requires more rigorous techniques.
+The method used ensured that the top 10 suppliers are accurately aggregated.
+However, some contracts that these suppliers may have been missed.
+But beyond the top 10, the names of suppliers not well parsed out, with problems like variations of “incorporated” and that sort of thing, dominating the rest of the suppliers list.
+
+Model Limitations: The linear regression model demonstrated weaknesses, primarily due to its simplicity and the limitations of the dataset.
+The assumption of a linear relationship between the predictors (Baseline Days and Duration Days) and contract amounts may not capture the complexities of procurement spending, which can be influenced by many things such as; policy changes, economic conditions, and sector-specific dynamics.
+
+Methodological Constraints: The model did not account for external variables such as inflation, exchange rates, or changes in government procurement policies, which could significantly influence contract amounts.
+Additionally, treating all contracts as a homogeneous group overlooks the diversity in contract types and scopes.
 
@@
 
-TODO Next Steps: Need to collect more data:
-      From Past spending beyond 2020
-      Other types of Government Spending; Seeing how much of spending is done through private contracts and how much through in house spending on salaries and other things like that
-      Do analysis on the what the contracts descriptions say
-      Why does public health Canada spending go to zero
-      Why are contracts so small
+To address these weaknesses and build upon the findings, the following steps are recommended:
+
+Enhance Data Analysis Techniques: 
+Supplier Name Processing: Implement advanced text processing techniques, such as fuzzy matching, to accurately standardize supplier names and improve data reliability. ADD MORE
+
+Contract Description Analysis: Utilize text mining and content analysis methods on contract descriptions to uncover patterns in procurement objectives and priorities. ADD MORE
+
+Expand Data Collection:
+Comprehensive Spending Data: Include other forms of government spending, such as in-house expenditures on salaries and operational costs, as this would contextualize the procurement budget in the larger scheme of federal spending overall.
+How much of the total federal budget goes into procurements?
+Are procurements more cost effective than in house spending?
+
+Investigate Specific Trends:
+Public Health Agency Spending: Examine the factors contributing to the decline in the Public Health Agency of Canada's procurement spending.
+
+Proliferation of Small Contracts: Analyze the rationale behind the high frequency of small contracts, exploring whether this is a strategic decision to support smaller businesses or a reflection of procurement policy regulations.
+
+Refine Modeling Approaches:
+Advanced Statistical Models: Explore more complex modeling techniques, such as nonlinear regression, time series analysis, or machine learning models, to capture the intricate relationships between variables.
+
+Incorporate Additional Variables: Include economic indicators, policy changes, and other relevant factors to improve the explanatory power of the models.
+
+By undertaking these steps, future research can provide a more comprehensive and nuanced understanding of Canadian federal procurement spending.
+This further analysis could inform policy decisions, promote transparency, and ensure that government resources are allocated effectively and efficiently.
 
@@
 TODO processed_supplier
-Talk about what I did thourougly
+Talk about what I did thourougly, see data_cleaning script
 

### 2024-12-02 paper/paper.qmd
fixed

@@
 mil_model <- read_rds(here::here("models/mil_model.rds"))
-
-aView(df)
 ```

### 2024-12-02 paper/paper.qmd
updated paper

@@
 #| warning: false
-#| fig-cap: template
+#| fig-cap: Model Results Summary
 #| fig-align: center
@@
 
-# Format the table for Quarto using kableExtra
+# Determine the output format
+output_format <- knitr::opts_knit$get("rmarkdown.pandoc.to")
+
+# Choose styling based on output format
+if (output_format == "latex") {
+  combined_results %>%
+    select(model, term, estimate, std.error, statistic, p.value) %>%
+    rename(
+      Model = model,
+      Term = term,
+      Estimate = estimate,
+      `Std. Error` = std.error,
+      `t-Statistic` = statistic,
+      `p-Value` = p.value
+    ) %>%
+    kable(booktabs = TRUE, caption = "Model Results Summary") %>%
+    kable_styling(latex_options = c("striped", "hold_position"))
+} else {
   combined_results %>%
@@
     kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
-
+}
 ```

### 2024-12-02 paper/paper.qmd
fixed table and cleaned up intro

@@
 
-Contract: A few words detailing  a brief description of what the contract signed was for
+- **Contract**: A few words detailing a brief description of what the contract signed was for.
+- **Buyer**: The name of the government department who awarded the contract.
+- **Supplier**: The name of the company that fulfilled the contract by agreeing to provide the good or service to the Buyer.
+- **Amount**: The value of the contract in Canadian Dollars.
+- **Award Date**: The day the Buyer awarded the contract to the supplier.
+- **Start Date**: When the contract is due to start.
+- **End Date**: When the contract is due to end.
 
-Buyer: The name of government department who awarded the contract
+The head of the IFJ Database:
 
-Supplier: The name of the company that fulfilled the contract by agreeing to provide the good or service to the Buyer
-
-Amount: The value of the contract in Canadian Dollars
-
-Award Date: The day the Buyer awarded the contract to the supplier
-
-Start Date: When the contract is due to start
-
-End Date: When the contract is due to end
-
-I derived three additional variables using the IJF variables:
-
-Baseline Days: Number of days since 1st January 2020
-
-Duration Days: End Date subtracted from the Start Date to calculate the duration of the contract
-
-Processed Supplier: The name of suppliers regularized to allow for aggregation and comparison
-
-Baseline Days and Duration Days are used as predictors in the linear model, as Dates are not valid entries into a linear model.
-
-Processed Supplier is required as many companies have slightly different spellings, such as [“General Dynamics”, “GENERAL DYNAMICS”, “General Dynamics Mission Systems”, “GENERAL DYNAMICS MISSION SYSTEMS INC”].
-See Appendix TODO for how I processed and regularized the suppliers.
-
-Here is the head of the data that was used in this analysis for a sense of what these variables represent. 
-TODO Fix table
-```{r fig-head}
+```{r fig-head1}
 #| message: false
@@
 #| fig-align: center
+#| fig-pos: "H"
 
-df_table <- df %>% select(-contract, -start_date, -end_date, -link, -duration_days, -per_day, -processed_supplier)
+df_table1 <- df %>%
+  select(
+    Contract = contract,
+    Buyer = buyer,
+    Supplier = supplier,
+    Amount = amount,
+    `Award Date` = award_date,
+    `Start Date` = start_date,
+    `End Date` = end_date
+  )
 
-kable(head(df_table, 5), caption = "Sample of the Data") %>%
-  kable_styling(font_size = 10, latex_options = c("scale_down"))
+df_table1_truncated <- df_table1 %>%
+  mutate(across(where(is.character), ~ str_trunc(., 15, side = "right", ellipsis = "...")))
+
+# Display the first table
+kable(head(df_table1_truncated, 5), caption = "Sample of the Data - Part 1") %>%
+  kable_styling(
+    font_size = 10,
+    latex_options = c("scale_down")
+  )
 
 ```
+
+Additional Derived Variables:
+
+- **Baseline Days**: Number of days since 1st January 2020.
+- **Duration Days**: End Date subtracted from the Start Date to calculate the duration of the contract.
+- **Processed Supplier**: The name of suppliers regularized to allow for aggregation and comparison.
+
+```{r fig-head2}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-cap: template
+#| fig-align: center
+#| fig-pos: "H"
+
+df_table2 <- df %>%
+  select(
+    `Baseline Days` = baseline_days,
+    `Duration Days` = duration_days,
+    `Processed Supplier` = processed_supplier
+  )
+
+# Truncate character columns if necessary
+df_table2_truncated <- df_table2 %>%
+  mutate(across(where(is.character), ~ str_trunc(., 15, side = "right", ellipsis = "...")))
+
+# Display the second table
+kable(head(df_table2_truncated, 5), caption = "Sample of the Data - Part 2") %>%
+  kable_styling(
+    font_size = 10,
+    latex_options = c("scale_down")
+  )
+
+```
+
+Processed Supplier is required as many companies have slightly different spellings, such as [“General Dynamics”, “GENERAL DYNAMICS”, “General Dynamics Mission Systems”, “GENERAL DYNAMICS MISSION SYSTEMS INC”].
+See Appendix TODO for how I processed and regularized the suppliers.
 

### 2024-12-02 scripts/03-data_cleaning.R
updated cleaned data

@@ -258,7 +258,16 @@ df <- df %>%
 
 cleaned <- df
 
-#View(cleaned)
+cleaned <- cleaned %>%
+  mutate(
+    baseline_days = as.numeric(award_date - as.Date("2020-01-01"))
+  )
+
+cleaned <- cleaned %>%
+  filter(format(award_date, "%Y") != "2019")
+
+
+View(cleaned)
 
 #### Save data ####
 write_csv(cleaned, "data/analysis_data/cleaned_data.csv")

### 2024-12-02 scripts/06-model.R
cleaned model file

@@ -24,16 +24,11 @@ df <- read_csv("data/analysis_data/cleaned_data.csv")
 
 
 ### Model data ####
-df <- df %>%
-  mutate(
-    award_date_numeric = as.numeric(award_date)
-  )
 
+# Fit the models using baseline_days
+model <- lm(amount ~ baseline_days + duration_days, data = df %>% filter(buyer != "National Defence"))
 
-
-model <- lm(amount ~ award_date_numeric + duration_days, data = df)
-
-mil_model <- lm(amount ~ award_date_numeric + duration_days, data = df %>% filter(buyer == "National Defence"))
+mil_model <- lm(amount ~ baseline_days + duration_days, data = df %>% filter(buyer == "National Defence"))
 
 model_summary <- summary(model)
 mil_model_summ <- summary(mil_model)

### 2024-12-03 paper/paper.qmd
added custom colors + fixed first 3 graphs

@@
 ```
+
+\newpage
 
@@
 
+\newpage
+
 # Data {#sec-data}
@@
 The head of the IFJ Database:
-
 ```{r fig-head1}
@@
 #| warning: false
-#| fig-cap: template
+#| fig-cap: Sample of the Data
 #| fig-align: center
@@
     `End Date` = end_date
+  ) %>%
+  mutate(
+    Amount = comma(Amount)  # Format Amount with commas
   )
@@
 # Display the first table
-kable(head(df_table1_truncated, 5), caption = "Sample of the Data - Part 1") %>%
+kable(head(df_table1_truncated, 5)) %>%
   kable_styling(
@@
 #| warning: false
-#| fig-cap: template
+#| fig-cap: Sample of Derived Data
 #| fig-align: center
@@
     `Processed Supplier` = processed_supplier
+  ) %>%
+  mutate(
+    `Baseline Days` = comma(`Baseline Days`),  # Format Baseline Days with commas
+    `Duration Days` = comma(`Duration Days`)   # Format Duration Days with commas
   )
@@
 # Display the second table
-kable(head(df_table2_truncated, 5), caption = "Sample of the Data - Part 2") %>%
+kable(head(df_table2_truncated, 5)) %>%
   kable_styling(
@@
 #| warning: false
-#| fig-cap: Top 5 Buyers by Total Amount Spent and Others (Pie Chart)
+#| fig-cap: Top 5 Buyers by Total Amount Spent and Others
 #| fig-align: center
+#| fig-pos: "H"
+
+colors <- c("#377eb8", "#e41a1c", "#4daf4a", "#ff7f00", "#984ea3", "#808080")
 
@@
 buyer_colors <- c(
-  "National Defence" = "#1f77b4",
-  "Public Services and Procurement Canada" = "#ff7f0e",
-  "Shared Services Canada" = "#2ca02c",
-  "Public Health Agency of Canada" = "#d62728",
-  "Fisheries and Oceans Canada" = "#9467bd",
-  "Other"  = "#8c564b"
+  "National Defence" = "#e41a1c",
+  "Public Services and Procurement Canada" = "#984ea3",
+  "Shared Services Canada" = "#ff7f00",
+  "Public Health Agency of Canada" = "#4daf4a",
+  "Fisheries and Oceans Canada" = "#377eb8",
+  "Other"  = "#808080"
 )
@@
   labs(
-    title = "Top 5 Buyers by Total Amount Spent and Others",
     fill = "Buyer"
@@
             color = "white",
-            size = 4)
+            size = 2.5)
 ```
@@
 #| warning: false
-#| fig-cap: template
+#| fig-cap: Military Spending Per Year in Billions
 #| fig-align: center
+#| fig-pos: "H"
 
@@
 
-# Summarize total spending per year
 military_spending_per_year <- military_df %>%
   group_by(year) %>%
-  summarize(total_spending = sum(amount, na.rm = TRUE))
+  summarize(total_spending = sum(amount, na.rm = TRUE) / 1e9)
 
@@
 ggplot(data = military_spending_per_year, aes(x = year, y = total_spending)) +
-  geom_bar(stat = "identity", fill = "steelblue") +
-  labs(title = "Military Spending Per Year", x = "Year", y = "Total Spending") +
+  geom_bar(stat = "identity", fill = "#e41a1c") +
+  labs(x = "Year", y = "Total Spending (in Billions)") +
   theme_minimal() +
@@
 #| warning: false
-#| fig-cap: template
+#| fig-cap: Public Services Spending Per Year in Billions
 #| fig-align: center
+#| fig-pos: "H"
 
@@
   group_by(year) %>%
-  summarize(total_spending = sum(amount, na.rm = TRUE))
+  summarize(total_spending = sum(amount, na.rm = TRUE)  / 1e9 )
 
@@
 ggplot(data = public_spending_per_year, aes(x = year, y = total_spending)) +
-  geom_bar(stat = "identity", fill = "steelblue") +
-  labs(title = "Public Services Spending Per Year", x = "Year", y = "Total Spending") +
+  geom_bar(stat = "identity", fill = "#984ea3") +
+  labs(x = "Year", y = "Total Spending (in Billions)") +
   theme_minimal() +
@@
 #| fig-align: center
+#| fig-pos: "H"
 
@@
 
-# Optional: Display the summarized data
-print(total_spent_df)
-
 # Create the bar chart
@@
 TODO
+
+\newpage
 
@@
 testthat
+colorbrewerV2

### 2024-12-03 paper/paper.qmd
fixed top suppliers graph

@@
 mil_model <- read_rds(here::here("models/mil_model.rds"))
+
+#colors!
+def_c <- "#e41a1c"
+fish_c <- "#377eb8"
+health_c <- "#4daf4a"
+services_c <- "#ff7f00"
+public_c <- "#984ea3"
+space_c <- "#ffff33"
+
 ```
@@
 
-colors <- c("#377eb8", "#e41a1c", "#4daf4a", "#ff7f00", "#984ea3", "#808080")
-
 # Summarize total amount spent by each buyer
@@
 buyer_colors <- c(
-  "National Defence" = "#e41a1c",
-  "Public Services and Procurement Canada" = "#984ea3",
-  "Shared Services Canada" = "#ff7f00",
-  "Public Health Agency of Canada" = "#4daf4a",
-  "Fisheries and Oceans Canada" = "#377eb8",
-  "Other"  = "#808080"
+  "National Defence" = def_c,
+  "Public Services and Procurement Canada" = public_c,
+  "Shared Services Canada" = services_c,
+  "Public Health Agency of Canada" = health_c,
+  "Fisheries and Oceans Canada" = fish_c,
+  "Other"  = "gray"
 )
@@
 ggplot(data = military_spending_per_year, aes(x = year, y = total_spending)) +
-  geom_bar(stat = "identity", fill = "#e41a1c") +
+  geom_bar(stat = "identity", fill = def_c) +
   labs(x = "Year", y = "Total Spending (in Billions)") +
@@
 ggplot(data = public_spending_per_year, aes(x = year, y = total_spending)) +
-  geom_bar(stat = "identity", fill = "#984ea3") +
+  geom_bar(stat = "identity", fill = public_c) +
   labs(x = "Year", y = "Total Spending (in Billions)") +
@@
 #| warning: false
-#| fig-cap: template
+#| fig-cap: Spending by Top 5 Buyers and Others Per Year
 #| fig-align: center
@@
     values = c(
-      "National Defence" = "steelblue",
-      "Public Services and Procurement Canada" = "orange",
-      "Shared Services Canada" = "green",
-      "Public Health Agency of Canada" = "purple",
-      "Fisheries and Oceans Canada" = "red",
+  "National Defence" = def_c,
+  "Public Services and Procurement Canada" = public_c,
+  "Shared Services Canada" = services_c,
+  "Public Health Agency of Canada" = health_c,
+  "Fisheries and Oceans Canada" = fish_c,
   "Other"  = "gray"
@@
   labs(
-    title = "Spending by Selected and Other Buyers Per Year",
-    subtitle = "Amounts are in Billions of CAD",
     x = "Year",
@@
   theme(
-    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
-    plot.subtitle = element_text(hjust = 0.5, size = 12),
-    axis.text.x = element_text(angle = 45, hjust = 1), # Rotate x-axis labels for readability
-    axis.text = element_text(size = 12),
-    axis.title = element_text(size = 14),
-    legend.title = element_text(size = 14),
-    legend.text = element_text(size = 12),
-    plot.caption = element_text(hjust = 0.5, size = 10, face = "italic") # Centered caption
+    legend.text = element_text(size = 8), # Adjusts the size of legend text
+    legend.title = element_text(size = 9), # Adjusts the size of legend title
+    legend.key.size = unit(0.6, "lines") # Reduces the size of legend symbols
   )
@@
 #| warning: false
-#| fig-cap: template
+#| fig-cap: Top 10 Suppliers by Total Amount Awarded
 #| fig-align: center
 #| fig-pos: "H"
+# Merge total aggregated amounts back into aggregated_df
+aggregated_df <- aggregated_df %>%
+  left_join(aggregated_totals, by = "word", suffix = c("", "_total"))
 
-#TODO Fix colors and supplier names and order of supps (highest to lowest)
+# Get the top 10 suppliers
+top_suppliers <- aggregated_totals %>%
+  top_n(10, total_amount)
 
-# Create the bar chart with legend removed
-ggplot(aggregated_df, aes(x = reorder(word, total_amount), y = total_amount, fill = buyer)) +
+# Filter aggregated_df to include only the top 10 suppliers
+aggregated_df_top10 <- aggregated_df %>%
+  filter(word %in% top_suppliers$word)
+
+# Use this ordering in the plot
+# Use this ordering in the plot (reversed order)
+ggplot(aggregated_df_top10, aes(x = reorder(word, total_amount_total), y = total_amount / 1e9, fill = buyer)) +
   geom_bar(stat = "identity") +
@@
   labs(
-    title = "Top 10 Suppliers by Aggregated Amount Awarded",
     x = "Supplier",
-    y = "Total Amount Awarded"
-    # Remove the legend title by not specifying 'fill' in labs
+    y = "Total Amount Awarded (in Billions)",
+    fill = "Buyer"  # Adds a title to the legend
+  ) +
+  scale_y_continuous(labels = scales::comma) +  # Format y-axis with commas
+  scale_fill_manual(
+    values = c(
+      "National Defence" = def_c, 
+      "Canadian Space Agency" = space_c, 
+      "Shared Services Canada" = services_c, 
+      "Public Health Agency of Canada" = health_c, 
+      "Fisheries and Oceans Canada" = fish_c,
+      "Other" = "gray"  # Assign a default color for other buyers if necessary
+        ), 
+    breaks = selected_buyers  # Restricts legend to selected buyers
   ) +
   theme_minimal() +
-  theme(legend.position = "none")
+  theme(
+    legend.position = "right"  # Positions the legend on the right
+  )
 

### 2024-12-03 paper/paper.qmd
new color scheme

@@
 #colors!
-def_c <- "#e41a1c"
-fish_c <- "#377eb8"
+def_c <- "#fb8072"
+fish_c <- "#80b1d3"
 health_c <- "#4daf4a"
-services_c <- "#ff7f00"
-public_c <- "#984ea3"
-space_c <- "#ffff33"
+services_c <- "#fdb462"
+public_c <- "#bebada"
+space_c <- "#ffffb3"
 
@@
   "Other"  = "gray"
-    )
+    ),
+  guide = guide_legend(nrow = 3)
   ) +
@@
   theme_minimal() +
-  theme(
-    legend.text = element_text(size = 8), # Adjusts the size of legend text
-    legend.title = element_text(size = 9), # Adjusts the size of legend title
-    legend.key.size = unit(0.6, "lines") # Reduces the size of legend symbols
+  theme(legend.position = "bottom"
   )
@@
 #| fig-pos: "H"
+
+# Create a custom mapping for supplier renaming
+supplier_rename <- c(
+  "cae" = "CAE", 
+  "skyalyne" = "Skyalyne", 
+  "vancouver shipyards" = "Vancouver Shipyards", 
+  "f35" = "F-35 Joint Program", 
+  "bell" = "Bell", 
+  "airbus" = "Airbus", 
+  "textron" = "Textron Avaition", 
+  "dynamics" = "General Dynamics",
+  "dettwiler" = "Dettwiler and Associates", 
+  "atomics" = "Atomics Aeronautical Systems"
+)
+
+selected_buyers <- c(
+  "National Defence", 
+  "Canadian Space Agency", 
+  "Shared Services Canada", 
+  "Public Health Agency of Canada", 
+  "Fisheries and Oceans Canada"
+)
+
+# Adjust the dataframe with supplier renaming
+aggregated_df <- aggregated_df %>%
+  mutate(word = ifelse(word %in% names(supplier_rename), supplier_rename[word], word))
+
+aggregated_totals <- aggregated_df %>%
+  group_by(word) %>%
+  summarise(total_amount = sum(total_amount, na.rm = TRUE)) %>%
+  arrange(desc(total_amount))
+
 # Merge total aggregated amounts back into aggregated_df
@@
     y = "Total Amount Awarded (in Billions)",
-    fill = "Buyer"  # Adds a title to the legend
+    fill = "Buyer"
   ) +
@@
         ), 
-    breaks = selected_buyers  # Restricts legend to selected buyers
+    breaks = selected_buyers,
+    guide = guide_legend(nrow = 2)
   ) +
@@
   theme(
-    legend.position = "right"  # Positions the legend on the right
+    legend.position = "bottom"  # Positions the legend on the right
   )

### 2024-12-03 paper/paper.qmd
added diagnostics subsection + diagnostics tables in appendix

@@
 library(broom)
+library(car)
+library(caret)
 
@@
   geom_bar(stat = "identity", fill = def_c) +
-  labs(x = "Year", y = "Total Spending (in Billions)") +
+  labs(x = "Year", y = "Total Spending (Billions)") +
   theme_minimal() +
-  scale_y_continuous(labels = dollar_format(prefix = "$")) +
+  scale_y_continuous() +
   theme(plot.title = element_text(hjust = 0.5))
@@
   geom_bar(stat = "identity", fill = public_c) +
-  labs(x = "Year", y = "Total Spending (in Billions)") +
+  labs(x = "Year", y = "Total Spending (Billions)") +
   theme_minimal() +
-  scale_y_continuous(labels = dollar_format(prefix = "$")) +
+  scale_y_continuous() +
   theme(plot.title = element_text(hjust = 0.5))
@@
   scale_y_continuous(
-    labels = comma_format(accuracy = 1), # Formats y-axis labels without decimal places
-    breaks = pretty_breaks(n = 5), # Ensures approximately 5 breaks like 5, 10, 15
-    expand = expansion(mult = c(0, 0.05)) # Adds a small space above the highest bar
+    labels = comma_format(accuracy = 1),
+    breaks = pretty_breaks(n = 5),
+    expand = expansion(mult = c(0, 0.05))
   ) +
@@
 #| warning: false
-#| fig-cap: template
+#| fig-cap: Distribution of Contracts that are less than $100,000
 #| fig-align: center
@@
 percent_less_than <- num_less_than_100k / total_contracts
-
-print(paste0("Percentage of contracts <= 100,000 CAD: ", round(percent_less_than * 100, 2), "%"))
 
@@
 ggplot(contract_counts_df, aes(x = amount_label, y = Contract_Count_k)) +
-  geom_bar(stat = "identity", fill = "skyblue", color = "black") +
+  geom_bar(stat = "identity", fill = "skyblue") +
   scale_y_continuous(
@@
   labs(
-    title = "Number of Contracts per 10,000 CAD Award Amount Range",
     x = "Award Amount (Thousands)",
-    y = paste0("Number of Contracts ", if (max_count >= 1000) "(Thousands)" else ""),
-    caption = "Data includes contracts <= 100,000 CAD."
-  ) +
-  theme_minimal() +
-  theme(
-    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
-    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
-    axis.text.y = element_text(size = 12),
-    axis.title = element_text(size = 14),
-    plot.caption = element_text(hjust = 0.5, size = 10, face = "italic")
-  )
+    y = paste0("Number of Contracts ", if (max_count >= 1000) "(Thousands)" else "")
+  ) +
+  theme_minimal()
 ```
@@
 #| warning: false
-#| fig-cap: template
+#| fig-cap: Total Amount Spent on Contracts
 #| fig-align: center
@@
   mutate(
-    contract_size = ifelse(amount <= 100000, "Small Contracts (≤ 100k)", "Big Contracts (> 100k)")
+    contract_size = factor(
+      ifelse(amount <= 100000, 
+             "Small Contracts (≤ $100,000)", 
+             "Big Contracts (> $100,000)"),
+      levels = c("Small Contracts (≤ $100,000)", "Big Contracts (> $100,000)")  # Specify order
+    )
   )
@@
   scale_fill_manual(
-    values = c("Small Contracts (≤ 100k)" = "steelblue", "Big Contracts (> 100k)" = "tomato")
+    values = c("Small Contracts (≤ $100,000)" = "skyblue", "Big Contracts (> $100,000)" = "skyblue")
   ) +
@@
   labs(
-    title = "Total Amount Spent on Contracts",
-    subtitle = "Comparison between Small (≤ 100k CAD) and Big (> 100k CAD) Contracts",
     x = "Contract Size",
-    y = "Total Amount Spent (Billions of CAD)"
+    y = "Total Amount Spent (Billions)"
   ) +
@@
   theme(
-    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
-    plot.subtitle = element_text(hjust = 0.5, size = 14),
-    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 14),
-    axis.text.y = element_text(size = 12),
-    axis.title = element_text(size = 16),
-    legend.position = "none",  # Removes the legend as it's redundant
-    plot.caption = element_text(hjust = 0.5, size = 10, face = "italic")
+    legend.position = "none"
   )
@@
 
-TODO
+The model diagnostics highlight the limitations of these two linear models, see Appendix TODO for relevant diagnostic plots.
+The Residuals Vs.
+Fitted plots indicate large clumps in the first quadrant.
+This indicates a significant bias in the predictions for both models.
+The QQ-Normal plots indicate that the majority of the residuals follow a normal distribution.
+However, there is a significant tail off in the end likely due to a large number of outliers which conclusively violates the normality assumption.
+The Scale-Location plot for the General Model indicates a funnel shape near the origin, indicates the assumption of homoscedasticity is violated.
+While the National Defence Model has a clump there is no clear pattern so the assumption of homoscedasticity maybe intact.
+For all predictors across both models their VIF values are very close to 1, indicating no multicollinearity.
+Both Cook's distance plots show there is a significant number of outliers in both models.
 
@@
 
-TODO Add Diagnostics
+```{r fig-resid-model}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-align: center
+#| fig-cap: Residual Vs Fitted for General Model
+
+plot(model, which = 1)
+
+```
+
+```{r fig-resid-mil}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-align: center
+#| fig-cap: Residual Vs Fitted for Military Model
+
+plot(mil_model, which = 1)
+```
+
+```{r fig-qq-model}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-align: center
+#| fig-cap: QQ-Normal General Model
+
+# Q-Q Plot for full model
+plot(model, which = 2)
+
+```
+
+```{r fig-qq-mil}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-align: center
+#| fig-cap: QQ-Normal National Defence Model
+
+# Q-Q Plot for National Defence model
+plot(mil_model, which = 2)
+```
+
+```{r fig-sl-model}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-align: center
+#| fig-cap: Scale-Location Plot for General model
+
+# Scale-Location Plot for full model
+plot(model, which = 3)
+
+```
+
+```{r fig-sl-mil}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-align: center
+#| fig-cap: Scale-Location Plot for National Defence model
+
+# Scale-Location Plot for National Defence model
+plot(mil_model, which = 3)
+```
+
+```{r fig-rl-model}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-align: center
+#| fig-cap: Residuals vs Leverage for Full model
+
+# Residuals vs Leverage for full model
+plot(model, which = 5)
+```
+
+```{r fig-rl-mil}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-align: center
+#| fig-cap: Residuals vs Leverage for National Defence model
+
+# Residuals vs Leverage for National Defence model
+plot(mil_model, which = 5)
+```
+
+```{r fig-cooks-model}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-align: center
+#| fig-cap: Cook's Distance for full model
+
+#Identify influential data points using Cook's Distance.
+# Cook's Distance for full model
+plot(model, which = 4)
+```
+
+```{r fig-cooks-mil}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-align: center
+#| fig-cap: Cook's Distance for National Defence model
+
+#Identify influential data points using Cook's Distance.
+# Cook's Distance for National Defence model
+plot(mil_model, which = 4)
+```
+
+```{r}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-align: center
+#| fig-cap: VIF Table
+
+vif_model <- vif(model)
+vif_mil_model <- vif(mil_model)
+
+# Create a data frame with VIF values
+vif_table <- data.frame(
+  Predictor = names(vif_model),
+  Model = round(as.numeric(vif_model), 6),
+  Mil_Model = round(as.numeric(vif_mil_model), 6)
+)
+
+predictor_names <- c(
+  "baseline_days" = "Baseline Days",
+  "duration_days" = "Duration Days"
+)
+
+vif_table$Predictor <- predictor_names[vif_table$Predictor]
+
+# Display the table using knitr's kable for better formatting
+kable(vif_table, ,
+      col.names = c("Predictor", "General Model", "National Defence Model"),
+      align = 'c')
+
+```
 

### 2024-12-03 scripts/06-model.R
modified diagnostics

@@ -75,12 +75,6 @@ vif(model)
 # VIF for National Defence model
 vif(mil_model)
 
-#Perform the Shapiro-Wilk test to check if residuals are normally distributed.
-# Shapiro-Wilk test for full model residuals
-shapiro.test(residuals(model))
-
-# Shapiro-Wilk test for National Defence model residuals
-shapiro.test(residuals(mil_model))
 
 #Identify influential data points using Cook's Distance.
 # Cook's Distance for full model
@@ -91,7 +85,7 @@ n <- nrow(df)
 k <- length(coef(model)) - 1
 cooks_threshold <- 4 / (n - k - 1)
 influential_full <- which(cooks.distance(model) > cooks_threshold)
-influential_full
+length(influential_full)
 
 # Cook's Distance for National Defence model
 plot(mil_model, which = 4)
@@ -100,31 +94,7 @@ n_mil <- nrow(df %>% filter(buyer == "National Defence"))
 k_mil <- length(coef(mil_model)) - 1
 cooks_threshold_mil <- 4 / (n_mil - k_mil - 1)
 influential_mil <- which(cooks.distance(mil_model) > cooks_threshold_mil)
-influential_mil
-
-
-#Use k-fold cross-validation to assess the model's predictive performance.
-# Install and load the 'caret' package if not already installed
-
-# Define training control with 10-fold cross-validation
-train_control <- trainControl(method = "cv", number = 10)
-
-# Cross-validation for full model
-set.seed(123)  # For reproducibility
-cv_model <- train(amount ~ award_date_numeric + duration_days,
-                  data = df,
-                  method = "lm",
-                  trControl = train_control)
-print(cv_model)
-
-# Cross-validation for National Defence model
-df_mil <- df %>% filter(buyer == "National Defence")
-set.seed(123)  # For reproducibility
-cv_mil_model <- train(amount ~ award_date_numeric + duration_days,
-                      data = df_mil,
-                      method = "lm",
-                      trControl = train_control)
-print(cv_mil_model)
+length(influential_mil)
 
 
 

### 2024-12-03 scripts/03-data_cleaning.R
cleaned data cleaning

@@ -224,6 +224,9 @@ aggregated_df <- aggregated_df %>%
   ) %>%
   mutate(total_amount = ifelse(is.na(total_amount), 0, total_amount))
 
+
+aggregated_df <- aggregated_df[aggregated_df$total_amount > 0, ]
+
 write.csv(aggregated_df, "data/analysis_data/top_ten_suppliers.csv", row.names = FALSE)
 
 # 10. Create a Stacked Bar Chart for the Top Ten Suppliers
@@ -231,18 +234,6 @@ write.csv(aggregated_df, "data/analysis_data/top_ten_suppliers.csv", row.names =
 aggregated_df <- aggregated_df %>%
   left_join(top_ten %>% select(word, supplier_name), by = "word")
 
-# Create the bar chart with legend removed
-ggplot(aggregated_df, aes(x = reorder(word, total_amount), y = total_amount, fill = buyer)) +
-  geom_bar(stat = "identity") +
-  coord_flip() +
-  labs(
-    title = "Top 10 Suppliers by Aggregated Amount Awarded",
-    x = "Supplier",
-    y = "Total Amount Awarded"
-    # Remove the legend title by not specifying 'fill' in labs
-  ) +
-  theme_minimal() +
-  theme(legend.position = "none")
 
 # 10. Transform `processed_supplier` into a Character Column for CSV
 df <- df %>%

### 2024-12-03 other/EDA/EDA.qmd
deleted redundant data

@@
 #| fig-align: center
+
+```
+
+```{r fig-template96}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-cap: template
+#| fig-align: center
+
+#TODO Fix colors and supplier names and order of supps (highest to lowest)
+View(aggregated_df)
+
+# Create the bar chart with legend removed
+ggplot(aggregated_df, aes(x = reorder(word, total_amount), y = total_amount, fill = buyer)) +
+  geom_bar(stat = "identity") +
+  coord_flip() +
+  labs(
+    title = "Top 10 Suppliers by Aggregated Amount Awarded",
+    x = "Supplier",
+    y = "Total Amount Awarded"
+    # Remove the legend title by not specifying 'fill' in labs
+  ) +
+  theme_minimal() +
+  theme(legend.position = "right")
 

### 2024-12-03 paper/paper.qmd
fixed $ signs

@@
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "TODO First sentence. TODO Second sentence. TODO Third sentence. TODO Fourth sentence."
 format: pdf
@@
 Some significant findings presented in this paper include: military procurement spending has surged in the past few years.
-Where the Department of National Defence now dominates federal procurement spending, with expenditures exceeding $56 billion since 2020.
-This increase in spending is in part thanks to large contracts, including a $11.2 billion contract to SkyAlyne for fighter pilot training—the largest in our dataset—and multi-billion-dollar contracts for fighter jets and naval vessels.
+Where the Department of National Defence now dominates federal procurement spending, with expenditures exceeding \$56 billion since 2020.
+This increase in spending is in part thanks to large contracts, including a \$11.2 billion contract to SkyAlyne for fighter pilot training—the largest in our dataset—and multi-billion-dollar contracts for fighter jets and naval vessels.
 
@@
 The Other category consists the of the total awards given by the remaining 88 smallest Government Buyers.
-It is clear that National Defence dwarfs the procurement budget, spending over /$56 Billion on procurements since 2020.
+It is clear that National Defence dwarfs the procurement budget, spending over \$56 Billion on procurements since 2020.
 
@@
 
-FIG show how National Defence spending has taken off in recent years, where is 2021 it had not even crested /$5 Billion, but in the last 3 years spending has exploded to /$15 Billion a year.
+FIG show how National Defence spending has taken off in recent years, where is 2021 it had not even crested \$5 Billion, but in the last 3 years spending has exploded to \$15 Billion a year.
 
@@
 This means they are responsible for managing the procurement process for buyers across the government [REF].
-It is important to note that the PSPC sometimes does not publish small contracts, less than /$40,000.
+It is important to note that the PSPC sometimes does not publish small contracts, less than \$40,000.
 PSPC spending over time shows while military spending increases the rest of the government is taking a hit.
@@
 Furthermore, it seems that the number of contracts follows a Pareto distribution.
-There are 102 contracts that are worth $1, this maybe some legal formality.
+There are 102 contracts that are worth \$1, this maybe some legal formality.
 
@@
 
-Though most contracts are small contracts (less than $100,000), the sum of these 246,000 contracts is only worth a measly 7.34 Billion. The Skyalyne contract itself outweighs all the small contracts.
+Though most contracts are small contracts (less than \$100,000), the sum of these 246,000 contracts is only worth a measly 7.34 Billion. The Skyalyne contract itself outweighs all the small contracts.
 
@@
 Because Baseline days is slightly negative.
-This value suggests that every day that passes on average the value of non-military contracts is going down by $61.62 a day.
+This value suggests that every day that passes on average the value of non-military contracts is going down by \$61.62 a day.
 This decline could reflect shifts in government spending priorities away from non-military departments.
@@
 The positive coefficient for Baseline days suggests an upward trend in military contract values since 2020.
-In this case the model suggests that the cost of military contracts is increasing by $852.50 every day.
+In this case the model suggests that the cost of military contracts is increasing by \$852.50 every day.
 This is trend is due to the huge increases in military procurement we observed in FIG.
@@
 The impact of contract duration on the contract amount is substantially larger in military contracts compared to non-military ones.
-Where military contracts cost $38,146 more per day of the length of the contract.
+Where military contracts cost \$38,146 more per day of the length of the contract.
 This is again due largely by the 11 Billon Skyalyne contract, and the other large long contracts for training and fighter jets.
@@
 Surge in Military Spending: Military spending has dramatically increased over the past couple of years.
-National Defence emerged as the dominant spender, allocating over $56 billion since 2020.
-While military spending was under $5 billion in 2021, it escalated to approximately $15 billion annually in the following years.
-This spike is attributed to substantial contracts for advanced military equipment and services, such as the $11.2 billion contract awarded to SkyAlyne for fighter pilot training—the largest in the dataset.
+National Defence emerged as the dominant spender, allocating over \$56 billion since 2020.
+While military spending was under \$5 billion in 2021, it escalated to approximately \$15 billion annually in the following years.
+This spike is attributed to substantial contracts for advanced military equipment and services, such as the \$11.2 billion contract awarded to SkyAlyne for fighter pilot training—the largest in the dataset.
 Other notable contracts include multi-billion-dollar deals with Lockheed Martin, General Atomics, CAE, and Airbus for fighter planes and aviation services.
@@
 This pattern indicates a governmental preference for smaller contracts, possibly for regulatory reasons.
-There may be a rule that if a contract is below 20000 then it requires less scrutiny to be passed, in effect encouraging department to award more smaller contracts.
-Though it does not seem to be out of control as only TODO% of all spending in last 4 years is concentrated in contracts less than 100000.
+There may be a rule that if a contract is below \$20,000 then it requires less scrutiny to be passed, in effect encouraging department to award more smaller contracts.
+Though this does not seem to be out of control as only TODO% of all spending in last 4 years is concentrated in contracts less than \$100,000.
 

### 2024-12-03 scripts/00-simulate_data.R
udated simulated data to make more realistic

@@ -14,7 +14,7 @@
 
 #### Workspace setup ####
 library(tidyverse)
-library(VGAM)      # For Pareto distribution
+library(actuar)     # For Pareto distribution
 library(dplyr)     # For data manipulation
 library(lubridate)
 
@@ -28,7 +28,7 @@ buyer_list <- c(
   "Canadian Institutes of Health Research",
   "The National Battlefields Commission",
   "Department of Justice Canada",
-  # "National Defence", # Excluded
+  #"National Defence", # Excluded
   "Public Services and Procurement Canada",
   "Canada Revenue Agency",
   "Canada Border Services Agency",
@@ -182,7 +182,6 @@ supplier_list <- c(
   "Ernst & Young Canada"
 )
 
-# 3. Simulate the 'buyer' column
 n <- 10000
 
 # 80% "National Defence"
@@ -205,41 +204,46 @@ contract_column <- sample(contract_list, size = n, replace = TRUE)
 supplier_column <- sample(supplier_list, size = n, replace = TRUE)
 
 # 6. Simulate the 'amount' column using Pareto distribution
-# Parameters for Pareto: x_m = 10,000; alpha = 3.5
-x_m <- 10000
-alpha <- 3.5
+# Parameters for Pareto: scale = 10,000; shape = 3.5
 
-# Using VGAM::rpareto
-amount_column <- rpareto(n, location = x_m, shape = alpha)
+scale <- 10000
+shape <- 3.5
 
-# Alternatively, if VGAM::rpareto is not available, use inverse transform sampling
-# U <- runif(n)
-# amount_column <- x_m / (U)^(1/alpha)
+# Using actuar::rpareto
+amount_column <- actuar::rpareto(n, scale = scale, shape = shape)
 
 # 7. Simulate 'award_date' and 'start_date'
 # Define date range
 start_date_min <- as.Date("2018-01-01")
 start_date_max <- as.Date("2024-12-31")
 
-# Generate random dates
+# Generate random award_start_dates
 award_start_dates <- as.Date(runif(n, min = as.numeric(start_date_min), max = as.numeric(start_date_max)), origin = "1970-01-01")
 
-# 8. Simulate 'duration_days'
-# Based on sample data, durations range roughly between 2,500 to 10,500 days
-# Use a skewed distribution, e.g., log-normal
-# Parameters can be adjusted to fit desired range
+# 8. Simulate 'end_date' first, then calculate 'duration_days'
 
[…]

### 2024-12-03 paper/paper.qmd
editing

@@
   - Robert Ford
-thanks: "Code and data are available at: https://github.com/Ford-Robert"
+thanks: "Code and data are available at: https://github.com/Ford-Robert .
+Special thanks to Professor Rohan Alexander for motivating this project, providing key support and insight, and demostrating the crucial skills required in modern statistical analysis.
+Check out his excellent book: https://tellingstorieswithdata.com/"
 date: today
@@
 
-In recent years, Canada's federal spending landscape has undergone significant transformations, particularly with a marked increase in military expenditures.
-The government has announced substantial new military spending packages, [REF]
+In recent years, Canada's federal spending landscape has undergone significant transformations, particularly with a marked increase in Military spending.
+The government has announced substantial new military spending packages, [REF].
+This paper will use data collected by the Investigative Journalism Foundation (IJF) detailing Canada's Federal Procurement Awards [REF] to analyse government procurements over the last four years.
 
 Federal procurements involve the acquisition of goods and services by government departments, Buyers, from companies, Suppliers.
-These procurements are essential for a wide range of government functions from; acquiring defense equipment, building infrastructure, procuring healthcare supplies, technological services and engineering consultants for example.
-This paper will analyse federal procurement data to dissect government procurements over the last four years.
+These procurements are essential for a wide range of government functions from; acquiring defense equipment, building infrastructure, procuring healthcare supplies, technological services and engineering consultants to name a few.
 
@@
 
-I collected and processed detailed data on over 301,000 contracts awarded since January 2020.
+I collected and processed data on well over 300,000 contracts awarded since January 2020.
 The dataset includes variables such as contract amounts, durations, award dates, Buyers, and Suppliers.
@@
 
-Some significant findings presented in this paper include: military procurement spending has surged in the past few years.
+Some significant findings presented in this paper include: military procurement spending has surged in the last two years.
 Where the Department of National Defence now dominates federal procurement spending, with expenditures exceeding \$56 billion since 2020.
-This increase in spending is in part thanks to large contracts, including a \$11.2 billion contract to SkyAlyne for fighter pilot training—the largest in our dataset—and multi-billion-dollar contracts for fighter jets and naval vessels.
-
+This increase in spending is in part thanks to large contracts, including a \$11.2 billion contract to SkyAlyne for fighter pilot training (the largest contract in our dataset) and multi-billion-dollar contracts for fighter jets and naval vessels.
 While  other government departments have experienced a notable decrease in procurement spending.
-For instance, the Public Health Agency of Canada's spending, which was substantial in 2020 due to the COVID-19 pandemic, has declined significantly in subsequent years.
+For instance, the Public Health Agency of Canada's spending, which was substantial in 2020 due to the COVID-19 pandemic, has declined significantly almost nothing in subsequent years.
 
@@
 
-The paper is structured as follows: [Section 2] details the data collection and processing methodologies, including the sources of procurement data and the steps taken to clean and standardize the dataset.
-Then I present a series of graphs to demonstrate government procurements from various angles.
-[Section 3] describes the linear regression models used to assess trends in contract amounts over time and contract duration.
-[Section 4] presents the results of the analysis, highlighting key trends in military and non-military procurement spending.
-[Section 5] discusses the limitations of the study, including data constraints and methodological considerations, and suggests areas for future research.
-Finally, Section 6 concludes with a summary of the findings and their implications for policymakers, industry stakeholders, and future studies.
-
-By providing a comprehensive analysis of federal procurement spending, this paper aims to fill the existing gap in understanding how government consumption of private sector goods and services is changing over time.
+The paper is structured as follows: @sec-data details the data collection and processing methodologies, including the sources of procurement data and the steps taken to clean and standardize the dataset.
+Then I present a series of graphs to demonstrate government procurements from various angles to provide a clear picture of spending over the last four years.
+@sec-model describes the linear regression models used to assess trends in contract amounts over time and contract duration.
+@sec-results presents the results of the analysis, highlighting key trends in military and non-military procurement spending.
+@sec-discussion discusses the limitations of the study, including data constraints and methodological considerations, and suggests areas for future research.
+Finally, @sec-discussion concludes with a summary of the findings and their implications for policymakers, industry stakeholders, and future studies.
+Additional information about how the data was extracted using web scraping, how the data was cleaned, and how suppliers names were processed can be found in @sec-appendix.
+Model diagnostics can be found at @sec-diagnostics.
+
+By providing an analysis of federal procurement spending, this paper aims to inform readers on how government consumption of private sector goods and services is changing over time.
 The insights gained from this study can inform policy decisions, contribute to academic discourse, and enhance transparency regarding government spending practices.
 
-TODO Cite R and all other packages
-
-TODO Estimand?
-
 \newpage
@@
 
-The data was collected using R and other packages [REF].
-A web scraping script was used to collect all Federal Award Data from the IJF [REF].
-The variables this paper focuses collected from the IJF are; Contract, Buyer, Supplier, Amount, Award Date, Start Date, and End Date.
+Data was analyzed through the R programming software [@citeR], and `RSelenium` [TODO] with `rvest` [TODO] were used to support a web scraping script to download all data.
+Additional packages such as `tidyverse` [@tidyverse], `knitr` [@knitr], `actuar` [TODO], `readr` [TODO], `stringr` [TODO], `broom` [TODO] and `dplyr` [@dplyr] were used to help clean, simulate, analyze, and test the data.
+The presentation of this data using tables and graphs was made possible by r-packages; `ggplot2` [TODO], `lubridate` [TODO], `kableExtra` [TODO], and `zoo` [TODO].
+
+A web scraping script was used to collect all Federal Award Data from the IJF [REF]. The variables this paper focuses are; Contract, Buyer, Supplier, Amount, Award Date, Start Date, and End Date.
 
@@
 
-The head of the IFJ Database:
+The @fig-head1 details the first few rows of the IFJ Database:
 ```{r fig-head1}
@@
 
-Additional Derived Variables:
+Additional Derived Variables, displayed in @fig-head2:
 
@@
 There were a number of contracts which started and ended on the same day, to simplify later calculation I simply added a day to their End Date, figuring that same day contracts are equivalent to one day contracts.
-Further data cleaning details are available in Appendix TODO.
-After cleaning I was left with 301,094 contracts spanning TODO to TODO.
+Further data cleaning details are available in @sec-appendix.
+After cleaning I was left with 320,175 contracts spanning January 2020 to November 2024.
 
@@
 The data is recorded and published through various government agencies, who use different accounting standards and procedures.
-The three sources IFJ used to collect the federal procurement data are; Canada Buys, Buy and Sell, and Proactive Disclosures (which makes up the vast majority of the procurement data).
-However, because the data used in this paper is pulled from the IFJ’s Federal Awards page this data comes exclusively from the Proactive Disclosures.
+The three sources IJF used to collect the federal procurement data are; Canada Buys, Buy and Sell, and Proactive Disclosures (which makes up the vast majority of the procurement data).
+However, because the data used in this paper is pulled from the IJF’s Federal Awards page this data comes exclusively from the Proactive Disclosures.
 The Treasury Board of Canada Secretariat is responsible for publishing these Proactive Disclosures under the Access to Information Act [REF].
@@
 
-This is a break down of total amount awarded by the top 5 biggest spending Buyers.
+@fig-pie-buyers is a break down of total amount awarded by the top 5 biggest spending Buyers.
 The Other category consists the of the total awards given by the remaining 88 smallest Government Buyers.
-It is clear that National Defence dwarfs the procurement budget, spending over \$56 Billion on procurements since 2020.
+It is clear that National Defence dwarfs the procurement budget, spending over \$56 Billion on procurements since 2020, making up nearly half of all government procurement.
 
@@
 
-FIG show how National Defence spending has taken off in recent years, where is 2021 it had not even crested \$5 Billion, but in the last 3 years spending has exploded to \$15 Billion a year.
+@fig-mil-spending shows how National Defence spending has taken off in recent years, where is 2021 it had not even crested \$5 Billion, but in the last 3 years spending has exploded to \$15 Billion a year.
+From 2021 to 2022, military procurements increased by more than 300%.
+Though it is important to note that these amounts reflect the amount awarded in any given year, while many awarded contracts last for years.
+Especially large contracts which can last decades.
+Meaning the payouts for these lengthy contracts could be amortized over the length of the contract.
 
@@
   geom_bar(stat = "identity", fill = def_c) +
-  labs(x = "Year", y = "Total Spending (Billions)") +
+  labs(x = "Year", y = "Total Awarded (Billions)") +
   theme_minimal() +
@@
 
-Public Services and Procurements Canada is the clearinghouse for the entire federal government.
+Public Services and Procurements Canada (PSPC) is the clearinghouse for the entire federal government.
 This means they are responsible for managing the procurement process for buyers across the government [REF].
-It is important to note that the PSPC sometimes does not publish small contracts, less than \$40,000.
-PSPC spending over time shows while military spending increases the rest of the government is taking a hit.
+It is important to note that the PSPC sometimes does not publish contracts less than \$40,000.
+@fig-public-spending shows PSPC spending over time, indicating that general government procurements are taking a hit while military spending balloons.
 ```{r fig-public-spending}
@@
   geom_bar(stat = "identity", fill = public_c) +
-  labs(x = "Year", y = "Total Spending (Billions)") +
+  labs(x = "Year", y = "Total Awarded (Billions)") +
   theme_minimal() +
@@
 
-This FIG shows the top five Buyers spending per year since 2020.
+@fig-top-buyers shows the top five Buyers spending per year since 2020.
 In the first 2 years military spending held steady and did not dominate the budget by any means.
 But since then Defence spending exploded not just nominally but compared to the rest of the budget as well.
-Notably the Public Health Agency of Canada’s spending while strong in 2020, decreased to nothing over the following four years.
-The spending in 2020 is linked to purchases of medical equipment, so it is possible they have not needed to invest in new equipment since then.
-But this illustrates a boarder trend among all departments except National Defence, where their spending has steadily declined in the last 4 years.
+Notably the Public Health Agency of Canada's spending while strong in 2020, decreased to nothing over the following four years.
+The spending in 2020 is linked to purchases of medical equipment, likely for combating the pandemic.
+Maybe this over investment has meant that that the Public Health Agency has not needed to invest in equipment in the subsequent years.
+Though investments across all government agencies, excluding the military, are falling and are lowest in 2024.
+Note that the 2024 data only goes to November, so more spending may happen in December.
 
@@
 
-To find out the firms that are benefiting the most through this procurement process we can look to FIG.
+To find out the firms that are benefiting the most through this procurement process we can look to @fig-top-supps.
 Unsurprisingly, the biggest suppliers are primarily are awarded contracts by the military.
-Interestingly the top supplier, Skyalyne, was only awarded one huge 11.2 Billion contract for fighter pilot training services.
+Interestingly the top supplier, SkyAlyne, was only awarded one huge \$11.2 Billion contract for fighter pilot training services.
 This is the largest contract in the dataset.
-Then considering the largest contracts awarded in the last 3 years to; the F-35 program (Fighter Planes, 4.5 Billion), General Atomics (Airplanes, 1.7 Billion), CAE (Aviation Engineering and Training, 4.4 Billion), Airbus (Airplanes, 3.7 Billion), for a grand total of 25.5 Billion.
-These large single contracts devoted to advanced fighter jets and training pilots to fly them is part of the reason that military spending has exploded in the last 3 years.
-The military also paid Vancouver Shipyards 3.4 Billion in 2024 for Ships and Boat parts.
+Then considering the largest contracts awarded since 2022 to; the F-35 program (Fighter Planes, \$4.5 Billion), General Atomics (Airplanes, \$1.7 Billion), CAE (Aviation Engineering and Training, \$4.4 Billion), Airbus (Airplanes, \$3.7 Billion), and Textron Aviation (Aircraft parts, \$3.1 Billion ) for a grand total of \$28.6 Billion.
+These few large contracts devoted to advanced fighter jets, training pilots, and parts are a significant portion of surge in military spending.
+Remember military procurement spending since 2022 has been about \$15 Billion dollar a year up from \$5 Billion in 2021, as per @fig-mil-spending.
+That equates to roughly \$30 Billion in extra spending (\$10 Billion extra over the \$5 Billion year for 3 years), \$28.6 Billion of which went to aviation projects.The military also paid Vancouver Shipyards \$3.4 Billion in 2024 for Ships and Boat parts.
 
@@
 
-Dettwiler is an engineering firm that provides the Canadian Space Agency with engineering and scientific consulting. Surprisingly 86.3% of the Space Agency’s procurement budget goes to Dettwiler.
-
-## Distribution
+Dettwiler is an engineering firm that provides the Canadian Space Agency with engineering and scientific consulting. Surprisingly 86.3% of the Space Agency's procurement budget goes to Dettwiler.
+
+## Contract Distribution
 
@@
 
-Though most contracts are small contracts (less than \$100,000), the sum of these 246,000 contracts is only worth a measly 7.34 Billion. The Skyalyne contract itself outweighs all the small contracts.
+Though most contracts are small contracts (less than \$100,000), the sum of these 246,000 contracts is only worth a measly 7.34 Billion. The SkyAlyne contract itself outweighs all the small contracts.
 
@@
 
-# Model
+# Model {#sec-model}
 
@@
 
-Tidyverse: For data manipulation and visualization [REF]
-
-Dplyr: [REF]
-
-Car: [REF]
-
-Caret:[REF]
+- Tidyverse: For data manipulation and visualization [REF] 
+- Dplyr: For data manipulation and transformation [REF] 
+- Car: For advanced regression modeling tools and diagnostics [REF] 
+- Caret: For classification and regression training, including model tuning and validation [REF]
 
@@
 
-TODO
+$$
+\text{Amount}_i = \beta_0 + \beta_1 \times \text{Base Line Days}_i + \beta_2 \times \text{Duration Days}_i + \epsilon_i
+$$
 
@@
 
-Amount:
-
-Baseline Days:
-
-Duration Days:
+- Amount: The Amount awarded to a contract
+- Baseline Days: The number of days since 1st of January 2020
+- Duration Days: The number of days that the contract spans
+
+The point of Baseline Days is to measure when the contracts take place, so when Baseline days is zero then it is one of the earliest contracts, as it gets larger that describes more recent contracts.
 
@@
 
-# Results
+# Results {#sec-results}
 
@@
 
-# Discussion
-
-## First discussion point {#sec-first-point}
+# Discussion {#sec-discussion}
+
+## What Was Done
 
@@
 Contracts that started and ended on the same day were adjusted by extending the end date by one day to simplify duration calculations.
-After these refinements, the dataset comprised 301,094 contracts spanning from [TODO: start date] to [TODO: end date].
 
@@
 
-## Second discussion point
+## State of Canadian Procurements
 
@@
 
-## Third discussion point
-
 ## Weaknesses
@@
 
-# Appendix {-}
-
-# Web Scraping
+# Data Details {#sec-appendix}
+
+## Web Scraping
 
@@
 
-# Additional data details
+## Data Cleaning
 
@@
 
-# Proccessing Supplier
+## Proccessing Supplier
 
 TODO processed_supplier
-Talk about what I did thourougly, see data_cleaning script
+Talk about what I did thoroughly, see data_cleaning script
 
@@
 
-# Model details {#sec-model-details}
-
-## Diagnostics
+# Model Diagnostics {#sec-diagnostics}
 

### 2024-12-03 scripts/05-exploratory_data_analysis.R
looking for longest contracts

@@ -78,6 +78,14 @@ total_spend <- df %>%
   summarize(total_amount = sum(amount, na.rm = TRUE)) %>%
   pull(total_amount)
 
+longest_contract <- df %>%
+  arrange(desc(duration_days)) %>%
+  slice(20)
+
+# Display the contract with the longest duration
+View(longest_contract)
+
+
 print(supp_total / dept_total)
 
 print(dept_total)

### 2024-12-03 other/EDA/EDA.qmd
added monthly spend graph

@@
 
-View(df)
+#View(df)
 
@@
 ```{r fig-template}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-cap: template
+#| fig-align: center
+
+df <- df %>%
+  mutate(Month = month(award_date, label = TRUE, abbr = FALSE))
+
+# Aggregate total amount awarded per month
+monthly_trends <- df %>%
+  group_by(Month) %>%
+  summarize(Total_Amount = sum(amount, na.rm = TRUE)) %>%
+  ungroup()
+
+# Order the months correctly
+monthly_trends$Month <- factor(monthly_trends$Month, levels = month.name)
+
+# Plot the bar chart
+ggplot(monthly_trends, aes(x = Month, y = Total_Amount / 1e9)) +
+  geom_bar(stat = "identity", fill = "skyblue") +
+  labs(
+    title = "Total Amount Awarded by Month",
+    x = "Month",
+    y = "Total Amount Awarded (Billions)"
+  ) +
+  theme_minimal() +
+  scale_y_continuous(labels = comma)
+
+```
+
+```{r fig-template101}
 #| message: false

### 2024-12-04 paper/paper.qmd
added ijf refrences + did TODOs

@@
 The government has announced substantial new military spending packages, [REF].
-This paper will use data collected by the Investigative Journalism Foundation (IJF) detailing Canada's Federal Procurement Awards [REF] to analyse government procurements over the last four years.
+This paper will use data collected by the Investigative Journalism Foundation [@ijf] detailing Canada's Federal Procurement Awards [@ijfprocure] to analyse government procurements over the last four years.
 
@@
 
-Data was analyzed through the R programming software [@citeR], and `RSelenium` [TODO] with `rvest` [TODO] were used to support a web scraping script to download all data.
-Additional packages such as `tidyverse` [@tidyverse], `knitr` [@knitr], `actuar` [TODO], `readr` [TODO], `stringr` [TODO], `broom` [TODO] and `dplyr` [@dplyr] were used to help clean, simulate, analyze, and test the data.
-The presentation of this data using tables and graphs was made possible by r-packages; `ggplot2` [TODO], `lubridate` [TODO], `kableExtra` [TODO], and `zoo` [TODO].
-
-A web scraping script was used to collect all Federal Award Data from the IJF [REF]. The variables this paper focuses are; Contract, Buyer, Supplier, Amount, Award Date, Start Date, and End Date.
+Data was analyzed through the R programming software [@citeR], and `RSelenium` [@rselenium] with `rvest` [@rvest] were used to support a web scraping script to download all data.
+Additional packages such as `tidyverse` [@tidyverse], `knitr` [@knitr], `actuar` [@actuar], `readr` [@readr], `stringr` [@stringr], `broom` [@broom] and `dplyr` [@dplyr] were used to help clean, simulate, analyze, and test the data.
+The presentation of this data using tables and graphs was made possible by r-packages; `ggplot2` [@ggplot2], `lubridate` [@lubridate], `kableExtra` [@kable], and `zoo` [@zoo].
+All colours are taken from `RColorBrewer` [@colorbrewer]
+
+A web scraping script was used to collect all Federal Award Data from the IJF [@ijfprocure].
+The variables this paper focuses are; Contract, Buyer, Supplier, Amount, Award Date, Start Date, and End Date.
 
@@
 Processed Supplier is required as many companies have slightly different spellings, such as [“General Dynamics”, “GENERAL DYNAMICS”, “General Dynamics Mission Systems”, “GENERAL DYNAMICS MISSION SYSTEMS INC”].
-See Appendix TODO for how I processed and regularized the suppliers.
+See @sec-appendix for more details on how I processed and regularized the suppliers.
 
@@
 
-# TODO: Assign specific colors to each buyer group
 # Example color assignments (replace with desired colors)
@@
 
-#TODO Make right color
-
 df_filtered <- df %>%
@@
 #| fig-pos: "H"
-
-#TODO Make right color
 
@@
 
-All model construction and diagnostics were conduction using R [REF], and the following packages:
-
-- Tidyverse: For data manipulation and visualization [REF] 
-- Dplyr: For data manipulation and transformation [REF] 
-- Car: For advanced regression modeling tools and diagnostics [REF] 
-- Caret: For classification and regression training, including model tuning and validation [REF]
+All model construction and diagnostics were conduction using R [@citeR], and the following packages:
+
+- Tidyverse: For data manipulation and visualization [@tidyverse] 
+- Dplyr: For data manipulation and transformation [@dplyr] 
+- Car: For advanced regression modeling tools and diagnostics [@car] 
+- Caret: For classification and regression training, including model tuning and validation [@caret]
 
@@
 
-The model diagnostics highlight the limitations of these two linear models, see Appendix TODO for relevant diagnostic plots.
+The model diagnostics highlight the limitations of these two linear models, see @diagnostics for relevant diagnostic plots.
 The Residuals Vs.
@@
 In this case the model suggests that the cost of military contracts is increasing by \$852.50 every day.
-This is trend is due to the huge increases in military procurement we observed in FIG.
+This is trend is due to the huge increases in military procurement we observed in @fig-mil-spending.
 
@@
 Where military contracts cost \$38,146 more per day of the length of the contract.
-This is again due largely by the 11 Billon Skyalyne contract, and the other large long contracts for training and fighter jets.
+This is again due largely by the \$11 Billion SkyAlyne contract, and the other large long contracts for training and fighter jets.
 
@@
 There may be a rule that if a contract is below \$20,000 then it requires less scrutiny to be passed, in effect encouraging department to award more smaller contracts.
-Though this does not seem to be out of control as only TODO% of all spending in last 4 years is concentrated in contracts less than \$100,000.
 
@@
 
-TODO Removal of negative values, adding 1 day to contract lengths of 0, see cleaning script
+I began by converting Amount into a numeric type and removing all the \$ signs.
+Then I converted all the date columns, Award Date, Start Date, End Date into date types.
+There was one date whose start year was was in 2204, this was clearly a typo so I set it to 2024.
+There were a significant number of contracts whose Start Date was the same as their End Date.
+This complicated some Per Day calculations I did, so I simply added one day to the End Dates of those contracts.
+I believe this hardly impacted the dataset because a same day contract and a one day contract are roughly the same in my view.
+I then filtered out any contracts that were awarded or started before 1st of January 2020.
+There were an number of contracts which had ridiculous start dates (i.e.
+2156) so I removed any contracts beyond 31st of December 2099, which may still be to late.
+Finally I removed any duplicate rows, I believe that the IJF database has a significant number of duplicate rows.
+It is also possible that rows were duplicated in the web scraping process as I noticed some strange behavior on IJF's website.
+If you go to the last page take note of the top companies there, and then the hit Previous page button, then the Next page button you will notice that its a difference set of companies.
+In any case I removed any duplicate rows.
+Finally I processed the supplier names.
 
@@
 
-TODO processed_supplier
-Talk about what I did thoroughly, see data_cleaning script
-
-Talk about next steps.
-11th top supplier, and how there are still tons of useless words in there.
-Even how to approach I used may never work.
-Because at some point we will ban too many word and some suppliers with super generic names will be eliminated entirely.
-
-Propose alternate methods
+To begin I made every suppliers name lowercase.
+Then I divided each word in a suppliers name and stored them in a list in the column processed_supplier.
+Then I made a list of 'banned' words, which are word that I will remove from each suppliers name.
+These banned words are thing like: incorporated, llc, the, service, basically any useless non-identifying words.
+But sometimes a company name is two banned words so they are left with and empty list in processed_supplier.
+Most notably this happened with Vancouver Shipyards, to fix this whenever the program found a list with vancouver and shipyards it would concatenate these words into one string.
+Then I loop through the the lists in processed_suppliers and removed any banned words.
+Finally I created a dictionary that would go through each processed_suppliers and pick out the individual words and set that as a key.
+Then I would loop through an if a key was present in a processed_suppliers list it would add the Amount of that contract to the value which would be a running sum for the suppliers with that key.
+
+I would then print out the top ten keys, at first many of these top ten were not descriptive and could not be related to a specific company.
+Then I would add this key to the banned words list and re-run the program.
+Eventually I had banned enough non-descriptive words that the top ten keys were aggregating only one supplier each and could be related to a specific company.
+
+Though this is the draw back of this method.
+If I wanted to get the top 15 or top 100 companies it would be very difficult to root out all of the confounding words.
+I also had to be vigilant when added a word to the banned words list because I could rule out an entire company, like what happened with Vancouver Shipyards.
+Then I had to create a work around that allowed be to remove the words "Vancouver" and "Shipyards", while being able to keep "Vancouver Shipyards".
+Essentially the problem with this method is that it is matching companies up if the share a single word in common, and that if fine for most companies (after getting rid of the banned words), but some company names are made up of generic words like Vancouver Shipyards.
+
+This problem would only get worse the further down the list you go and at some point it may be impossible to continue with this method.
+To make matters worse when adding new data you may already be banning all the words in their names and would have no idea.
+Because at least when you are looking to add a banned word you can check to see if that word is the last word for a lot of companies, but you won't get that opportunity for novel companies.
+In short I am sure the top 10 companies that I am displaying truly are the top 10, however using this method I cannot say what the top 11 or more companies are and if new data is introduced then all bets are off.
+
+To create a more rigorous supplier name processing method I would completely delete this method and start again.
+Some good candidates that I have looked into for better supplier name parsing would be: Fuzzy Matching and String Distance Algorithms, Natural Language Processing (Tokenization, Named Entity Recognition (NER)).
+It would be nice to apply these algorithms to the contract descriptions as well, in that case you could gain further insights into what the government is buying.
 
@@
 # References
-
-TODO Ref:
-
-RSelenuim
-VGAM
-stringr
-dplyr
-tidyverse
-zoo
-kableExtra
-knitr
-scales
-lubridate
-ggplot2
-rvest
-readr
-testthat
-colorbrewerV2

### 2024-12-04 scripts/03-data_cleaning.R
cleaned up

@@ -71,10 +71,10 @@ cleaned <- cleaned[cleaned$duration_days > 0, ]
 
 cleaned <- cleaned[cleaned$amount > 0, ]
 
-# Filter out contracts with 'start_date' or 'award_date' before 2015
+# Filter out contracts with 'start_date' or 'award_date' before 2020
 cleaned <- cleaned[
-  cleaned$start_date >= as.Date("2015-01-01") &
-    cleaned$award_date >= as.Date("2015-01-01"), ]
+  cleaned$start_date >= as.Date("2020-01-01") &
+    cleaned$award_date >= as.Date("2020-01-01"), ]
 
 # Filter out contracts with 'end_date' beyond 2099
 cleaned <- cleaned[cleaned$end_date <= as.Date("2099-12-31"), ]
@@ -253,9 +253,6 @@ cleaned <- cleaned %>%
   mutate(
     baseline_days = as.numeric(award_date - as.Date("2020-01-01"))
   )
-
-cleaned <- cleaned %>%
-  filter(format(award_date, "%Y") != "2019")
 
 
 View(cleaned)

### 2024-12-04 paper/paper.qmd
adde abstract and refrences

@@
 ---
-title: "Procurements of the Canadian Government, Runaway Military Contracts"
+title: "Procurements of the Canadian Government, Ballooning Military Awards for Fighter Jets"
 author: 
@@
 date-format: long
-abstract: "TODO First sentence. TODO Second sentence. TODO Third sentence. TODO Fourth sentence."
+abstract: "This paper gathers data on Canadain Federal Procurement Awards, uses a series of visualizations to analyse the data, builds two linear models to compare general government procurements and military procurements.
+Over the last three years the Canadian military has tripled their spending in an effort to aquire next generation fighter jets and train pilots, while every other government agency has atrophied.
+This spending concerns every Canadian citizen and often goes undercovered compared to analysis of the governments general budget."
 format: pdf
@@
 Data was analyzed through the R programming software [@citeR], and `RSelenium` [@rselenium] with `rvest` [@rvest] were used to support a web scraping script to download all data.
-Additional packages such as `tidyverse` [@tidyverse], `knitr` [@knitr], `actuar` [@actuar], `readr` [@readr], `stringr` [@stringr], `broom` [@broom] and `dplyr` [@dplyr] were used to help clean, simulate, analyze, and test the data.
+Additional packages such as `tidyverse` [@tidyverse], `knitr` [@knitr], `actuar` [@actuar], `arrow` [@arrow], `readr` [@readr], `stringr` [@stringr], `broom` [@broom] and `dplyr` [@dplyr] were used to help clean, simulate, analyze, and test the data.
 The presentation of this data using tables and graphs was made possible by r-packages; `ggplot2` [@ggplot2], `lubridate` [@lubridate], `kableExtra` [@kable], and `zoo` [@zoo].
@@
 
+# Contract Surveys {#sec-surveys}
+
 # Model Diagnostics {#sec-diagnostics}

### 2024-12-04 scripts/03-data_cleaning.R
added parquet!

@@ -8,6 +8,7 @@
 # - The `tidyverse` package must be installed and loaded
 # - The `dplyr` package must be installed and loaded
 # - The `stringr` package must be installed and loaded
+# - The `arrow` package must be installed and loaded
 # Other information? Make sure you are in the `canadian_gov_procurement` rproj
 
 
@@ -15,6 +16,7 @@
 library(tidyverse)
 library(dplyr)
 library(stringr)
+library(arrow)
 
 #### Clean data ####
 combined <- read_csv("data/raw_data/combined_data.csv")
@@ -259,3 +261,4 @@ View(cleaned)
 
 #### Save data ####
 write_csv(cleaned, "data/analysis_data/cleaned_data.csv")
+write_parquet(cleaned, "data/analysis_data/analysis_dataset.parquet")

### 2024-12-04 paper/paper.qmd
PURGED banned rubrik words + final read through editing

@@
 date-format: long
-abstract: "This paper gathers data on Canadain Federal Procurement Awards, uses a series of visualizations to analyse the data, builds two linear models to compare general government procurements and military procurements.
+abstract: "This paper gathers data on Canadain Federal Procurement Awards, and uses a series of visualizations to analyse the data.
+I then build two linear models to compare differences between general government procurements and military procurements.
 Over the last three years the Canadian military has tripled their spending in an effort to aquire next generation fighter jets and train pilots, while every other government agency has atrophied.
@@
 In recent years, Canada's federal spending landscape has undergone significant transformations, particularly with a marked increase in Military spending.
-The government has announced substantial new military spending packages, [REF].
+The government has announced substantial new military spending packages, as it aims to ramp up spending to 2% of GDP to meet NATO requirements [@nato].
 This paper will use data collected by the Investigative Journalism Foundation [@ijf] detailing Canada's Federal Procurement Awards [@ijfprocure] to analyse government procurements over the last four years.
 
-Federal procurements involve the acquisition of goods and services by government departments, Buyers, from companies, Suppliers.
+Federal procurements involve the acquisition of goods and services by government departments (Buyers) from companies (Suppliers).
 These procurements are essential for a wide range of government functions from; acquiring defense equipment, building infrastructure, procuring healthcare supplies, technological services and engineering consultants to name a few.
 
-The primary questions this paper aims to address are: What is the government buying? Who are the primary suppliers? How are spending patterns shifting across different departments and over time?
+The primary questions this paper aims to address are: What is the government buying? Who are the primary suppliers? How are spending patterns shifting across different departments over time?
 
@@
 
-Some significant findings presented in this paper include: military procurement spending has surged in the last two years.
+A significant finding presented in this paper is military procurement spending has surged in the last two years.
 Where the Department of National Defence now dominates federal procurement spending, with expenditures exceeding \$56 billion since 2020.
-This increase in spending is in part thanks to large contracts, including a \$11.2 billion contract to SkyAlyne for fighter pilot training (the largest contract in our dataset) and multi-billion-dollar contracts for fighter jets and naval vessels.
+This increase in spending is in part thanks to large contracts, including a \$11.2 billion contract to SkyAlyne for fighter pilot training (the largest contract in this dataset) and multi-billion-dollar contracts for fighter jets and naval vessels.
 While other government departments have experienced a notable decrease in procurement spending.
-For instance, the Public Health Agency of Canada's spending, which was substantial in 2020 due to the COVID-19 pandemic, has declined significantly almost nothing in subsequent years.
-
-These findings are important for several reasons.
-They provide critical insights into how government spending priorities are shifting, which has implications for economic policy, industry stakeholders, and public transparency.
+For instance, the Public Health Agency of Canada's spending, which was substantial in 2020 due to the COVID-19 pandemic, has declined significantly to almost nothing in recent years.
+
+These findings are important as they provide an understanding into how government spending priorities are shifting, which has implications for economic policy, industry stakeholders, and public transparency.
 Understanding the concentration of spending in large military contracts can inform discussions on national defense strategies, budget allocations, and the impact on domestic industries.
@@
 @sec-discussion discusses the limitations of the study, including data constraints and methodological considerations, and suggests areas for future research.
-Finally, @sec-discussion concludes with a summary of the findings and their implications for policymakers, industry stakeholders, and future studies.
-Additional information about how the data was extracted using web scraping, how the data was cleaned, and how suppliers names were processed can be found in @sec-appendix.
+Additional information about how the data was extracted using web scraping (@sec-appendix-web-scraping), how the data was cleaned (@sec-appendix-data-cleaning), and how suppliers names were processed can be found in @sec-appendix-supplier.
 Model diagnostics can be found at @sec-diagnostics.
@@
 By providing an analysis of federal procurement spending, this paper aims to inform readers on how government consumption of private sector goods and services is changing over time.
-The insights gained from this study can inform policy decisions, contribute to academic discourse, and enhance transparency regarding government spending practices.
+The information gained from this study can inform policy decisions, contribute to academic discourse, and enhance transparency regarding government spending practices.
 
@@
 Processed Supplier is required as many companies have slightly different spellings, such as [“General Dynamics”, “GENERAL DYNAMICS”, “General Dynamics Mission Systems”, “GENERAL DYNAMICS MISSION SYSTEMS INC”].
-See @sec-appendix for more details on how I processed and regularized the suppliers.
+See @sec-appendix-supplier for more details on how I processed and regularized the suppliers.
 
@@
 There were a number of contracts which started and ended on the same day, to simplify later calculation I simply added a day to their End Date, figuring that same day contracts are equivalent to one day contracts.
-Further data cleaning details are available in @sec-appendix.
+Further data cleaning details are available in @sec-appendix-data-cleaning.
 After cleaning I was left with 320,175 contracts spanning January 2020 to November 2024.
@@
 The three sources IJF used to collect the federal procurement data are; Canada Buys, Buy and Sell, and Proactive Disclosures (which makes up the vast majority of the procurement data).
-However, because the data used in this paper is pulled from the IJF’s Federal Awards page this data comes exclusively from the Proactive Disclosures.
-The Treasury Board of Canada Secretariat is responsible for publishing these Proactive Disclosures under the Access to Information Act [REF].
+However, because the data used in this paper is pulled from the IJF’s Federal Awards [@ijfprocure] page, this data comes exclusively from Proactive Disclosures.
+The Treasury Board of Canada Secretariat is responsible for publishing these Proactive Disclosures under the Access to Information Act.
+To gather more data on procurements in Canada I propose that the government requires Sellers to fill out a survey, and to publish this survey to the public.
+This survey would be designed to gain information into: the purpose and scope of a contract, relationship with Buyers, procurement process feedback, challenges on seller side, and future opportunities.
+See [@sec-survey] for my design of what this survey should look like.
 
 ## Analysis
-
-@fig-pie-buyers is a break down of total amount awarded by the top 5 biggest spending Buyers.
-The Other category consists the of the total awards given by the remaining 88 smallest Government Buyers.
-It is clear that National Defence dwarfs the procurement budget, spending over \$56 Billion on procurements since 2020, making up nearly half of all government procurement.
 
@@
 ```
-
-@fig-mil-spending shows how National Defence spending has taken off in recent years, where is 2021 it had not even crested \$5 Billion, but in the last 3 years spending has exploded to \$15 Billion a year.
-From 2021 to 2022, military procurements increased by more than 300%.
-Though it is important to note that these amounts reflect the amount awarded in any given year, while many awarded contracts last for years.
-Especially large contracts which can last decades.
-Meaning the payouts for these lengthy contracts could be amortized over the length of the contract.
+@fig-pie-buyers is a break down of the total amount awarded by the top 5 biggest spending Buyers.
+The Other category consists the of the total awards given by the remaining 88 smallest Government Buyers.
+It is clear that National Defence dwarfs the procurement budget, spending over \$56 Billion on procurements since 2020, making up nearly half of all government procurement.
 
@@
 ```
-
-Public Services and Procurements Canada (PSPC) is the clearinghouse for the entire federal government.
-This means they are responsible for managing the procurement process for buyers across the government [REF].
-It is important to note that the PSPC sometimes does not publish contracts less than \$40,000.
-@fig-public-spending shows PSPC spending over time, indicating that general government procurements are taking a hit while military spending balloons.
+@fig-mil-spending shows how National Defence spending has taken off in recent years, where is 2021 it had not even crested \$5 Billion, but in the last 3 years spending has exploded to \$15 Billion a year.
+From 2021 to 2022, military procurements have increased by more than 300%.
+Though it is important to note that these amounts reflect the amount awarded in any given year, while many awarded contracts can last for years.
+Especially large contracts which can last decades.
+Meaning the payouts for these lengthy contracts could be amortized over the length of the contract.
+
 ```{r fig-public-spending}
@@
 ```
-
-@fig-top-buyers shows the top five Buyers spending per year since 2020.
-In the first 2 years military spending held steady and did not dominate the budget by any means.
-But since then Defence spending exploded not just nominally but compared to the rest of the budget as well.
-Notably the Public Health Agency of Canada's spending while strong in 2020, decreased to nothing over the following four years.
-The spending in 2020 is linked to purchases of medical equipment, likely for combating the pandemic.
-Maybe this over investment has meant that that the Public Health Agency has not needed to invest in equipment in the subsequent years.
-Though investments across all government agencies, excluding the military, are falling and are lowest in 2024.
-Note that the 2024 data only goes to November, so more spending may happen in December.
+Public Services and Procurements Canada (PSPC) is the clearinghouse for the entire federal government.
+This means they are responsible for managing the procurement process for buyers across the government.
+It is important to note that the PSPC sometimes does not publish contracts less than \$40,000.
+@fig-public-spending shows PSPC spending over time, indicating that general government procurements have taken a hit in recent years.
 
@@
 ```
+@fig-top-buyers shows the top five Buyers spending per year since 2020.
+In the first 2 years military spending held steady and did not dominate the budget.
+But since then Defence spending exploded not just nominally but compared to the rest of the budget as well.
+Notably the Public Health Agency of Canada's spending while strong in 2020, decreased to nothing over the following four years.
+The spending in 2020 is linked to purchases of medical equipment, likely for combating the pandemic.
+Maybe the over investment that year has meant that that the Public Health Agency has not needed to invest in equipment in the subsequent years.
+Though investments across all government agencies, excluding the military, are falling and are at their lowest in 2024.
+Note that the 2024 data only goes to November, so more spending may happen in December.
 
 To find out the firms that are benefiting the most through this procurement process we can look to @fig-top-supps.
-Unsurprisingly, the biggest suppliers are primarily are awarded contracts by the military.
+Unsurprisingly, the biggest suppliers are primarily those awarded contracts by the military.
 Interestingly the top supplier, SkyAlyne, was only awarded one huge \$11.2 Billion contract for fighter pilot training services.
 This is the largest contract in the dataset.
-Then considering the largest contracts awarded since 2022 to; the F-35 program (Fighter Planes, \$4.5 Billion), General Atomics (Airplanes, \$1.7 Billion), CAE (Aviation Engineering and Training, \$4.4 Billion), Airbus (Airplanes, \$3.7 Billion), and Textron Aviation (Aircraft parts, \$3.1 Billion ) for a grand total of \$28.6 Billion.
-These few large contracts devoted to advanced fighter jets, training pilots, and parts are a significant portion of surge in military spending.
+Now consider some of the largestest contracts awarded since 2022 to; the F-35 program (Fighter Planes, \$4.5 Billion), General Atomics (Airplanes, \$1.7 Billion), CAE (Aviation Engineering and Training, \$4.4 Billion), Airbus (Airplanes, \$3.7 Billion), and Textron Aviation (Aircraft parts, \$3.1 Billion ) for a grand total of \$28.6 Billion.
+These large contracts devoted to advanced fighter jets, training pilots, and aircraft parts are a significant portion of the surge in military spending.
 Remember military procurement spending since 2022 has been about \$15 Billion dollar a year up from \$5 Billion in 2021, as per @fig-mil-spending.
@@
 
-Dettwiler is an engineering firm that provides the Canadian Space Agency with engineering and scientific consulting. Surprisingly 86.3% of the Space Agency's procurement budget goes to Dettwiler.
+Dettwiler is an engineering firm that provides the Canadian Space Agency with engineering and scientific consulting.
+Surprisingly 86.3% of the Space Agency's procurement budget goes to Dettwiler alone, about \$2.2 Billion.
 
@@
 
-Though most contracts are small contracts (less than \$100,000), the sum of these 246,000 contracts is only worth a measly 7.34 Billion. The SkyAlyne contract itself outweighs all the small contracts.
+Though, most contracts are small contracts (less than \$100,000), the sum of these 246,000 contracts is only worth a measly \$7.34 Billion.
+The SkyAlyne contract itself handily outweighs all the small contracts.
 
@@
 This paper will use a simple linear regression to better understand the how contract award amounts are affected over time compared to their lengths.
-To do this the response variable will be Amount, and the two dependent variables will be the number of day since January 1st 2020 (start of the data set) and the length of the contract.
-The coefficients of these variables will indicate the relative importance of when the contract was issues compared to the length of contracts on determining the size of contracts.
+To do this the response variable will be Amount, and the two dependent variables will be the number of days since January 1st 2020 (start of the data set) and the length of the contract.
+The coefficients of these variables will indicate the relative importance between when the contract was issues compared to the length of contracts.
 I will create two models, one to measure the coefficients for all the contracts excluding military contracts and a model for solely military contracts.
@@
 - Dplyr: For data manipulation and transformation [@dplyr] 
-- Car: For advanced regression modeling tools and diagnostics [@car] 
+- Car: For regression modeling tools and diagnostics [@car] 
 - Caret: For classification and regression training, including model tuning and validation [@caret]
@@
 
+Where i is either General Government Contracts (all but National Defence) or National Defence
+
 Variables:
@@
 
-The point of Baseline Days is to measure when the contracts take place, so when Baseline days is zero then it is one of the earliest contracts, as it gets larger that describes more recent contracts.
+The point of Baseline Days is to measure when the contracts take place, so when Baseline days is zero then it is one of the earliest contracts, as it gets larger it describes more recent contracts.
 
@@
 
-Additionally, the presence of outliers or extreme values, particularly in financial data like contract amounts, can disproportionately influence the model estimates and lead to misleading conclusions.
+Additionally, the presence of outliers or extreme values can disproportionately influence the model estimates and lead to misleading conclusions.
 
@@
 This may be difficult as dividing the data in to groups based on contract size would not be sufficient as there is still a wide variety of contracts that are of similar size.
-Possibly using the Contract descriptions to divide the data, but this would require more advanced Natural Language processing techniques and in many cases the descriptions are hardly descriptive.
+Possibly using the Contract descriptions to divide the data, but this would require Natural Language Processing techniques, and in many cases the descriptions are hardly descriptive.
 
 Moreover, the model assumes that the effect of each predictor is constant over time.
-If the influence of contract duration or time on contract amounts changes due to evolving market conditions or government strategies, the model's fixed coefficients would not reflect these dynamic relationships.
+If the influence of contract duration or time on contract amounts changes due to market conditions or government strategies, the model's fixed coefficients would not reflect these dynamic relationships.
 
@@
 
-The model diagnostics highlight the limitations of these two linear models, see @diagnostics for relevant diagnostic plots.
+The model diagnostics highlight the limitations of these two linear models, see @sec-diagnostics for relevant diagnostic plots.
 The Residuals Vs.
@@
 The QQ-Normal plots indicate that the majority of the residuals follow a normal distribution.
-However, there is a significant tail off in the end likely due to a large number of outliers which conclusively violates the normality assumption.
-The Scale-Location plot for the General Model indicates a funnel shape near the origin, indicates the assumption of homoscedasticity is violated.
-While the National Defence Model has a clump there is no clear pattern so the assumption of homoscedasticity maybe intact.
-For all predictors across both models their VIF values are very close to 1, indicating no multicollinearity.
-Both Cook's distance plots show there is a significant number of outliers in both models.
+However, there is a significant tail off in the end likely due to a large number of outliers which conclusively violate the normality assumption.
+The Scale-Location plot for the General Model indicates a funnel shape near the origin, indicating the assumption of homoscedasticity is violated.
+While the National Defence Model has a clump there is no clear pattern so the assumption of homoscedasticity may be intact.
+For all predictors across both models their VIF values are very close to 1, indicating no multicollinearity is present.
+Both Cook's distance plots show there is a significant number of outliers.
 
@@
 
-The model general contracts indicates that the contract amount tends to decrease over time.
-Because Baseline days is slightly negative.
+The General Government Model Baseline Days coefficient indicates that the contract amount tends to decrease over time.
+Because Baseline Days is slightly negative.
 This value suggests that every day that passes on average the value of non-military contracts is going down by \$61.62 a day.
 This decline could reflect shifts in government spending priorities away from non-military departments.
-Additionally, as expect Duration Days is positive of the contract amount .
-This means that longer contracts are more expensive for the government.
-
-In contrast, the model for military contracts reveals that the contract amount increases over time.
+Additionally, as expected Duration Days is positive which means that longer contracts are more expensive for the government.
+
+In contrast, the model for Military contracts indicates that the contract amount increases over time.
 The positive coefficient for Baseline days suggests an upward trend in military contract values since 2020.
@@
 Where military contracts cost \$38,146 more per day of the length of the contract.
-This is again due largely by the \$11 Billion SkyAlyne contract, and the other large long contracts for training and fighter jets.
+This is again due largely due to the \$11 Billion SkyAlyne contract, and the other large long contracts for training and fighter jets.
 
@@
 
-This study conducted an in-depth analysis of Canadian federal procurement data, with a focus on military contracts.
+This study conducted an analysis of Canadian federal procurement data, with a focus on military contracts.
 The data was collected and processed to ensure accuracy and relevance.
-Using web scraping techniques, we gathered information from the Institute for Government's (IJF) Federal Award Data.
+Using web scraping techniques, we gathered information from the IJF's Federal Award Data (@ijfprocure).
 The primary variables included contract descriptions, buyer and supplier names, contract amounts, and key dates such as award, start, and end dates.
@@
 
-Data cleaning was a critical step.
-Contracts with negative or zero amounts were excluded due to their ambiguous nature.
-Contracts that started and ended on the same day were adjusted by extending the end date by one day to simplify duration calculations.
-
 For the modeling phase, we employed a simple linear regression to explore how contract amounts are influenced by time and contract duration.
 Two separate models were constructed: one encompassing all contracts excluding military contracts, and another focusing solely on military contracts.
-This bifurcation allowed us to compare trends between military and non-military spending.
+This split allowed us to compare trends between military and non-military spending.
 
@@
 
-The analysis revealed significant trends in Canada's federal procurement spending:
-
-Surge in Military Spending: Military spending has dramatically increased over the past couple of years.
-National Defence emerged as the dominant spender, allocating over \$56 billion since 2020.
+Military spending has dramatically increased over the past couple of years.
+National Defence is the dominant spender, allocating over \$56 billion since 2020.
 While military spending was under \$5 billion in 2021, it escalated to approximately \$15 billion annually in the following years.
 This spike is attributed to substantial contracts for advanced military equipment and services, such as the \$11.2 billion contract awarded to SkyAlyne for fighter pilot training—the largest in the dataset.
-Other notable contracts include multi-billion-dollar deals with Lockheed Martin, General Atomics, CAE, and Airbus for fighter planes and aviation services.
-
-Decline in Other Departments' Spending: In contrast, other government departments have experienced significant reductions in procurement spending.
-For instance, the Public Health Agency of Canada's spending, which was robust in 2020 due to pandemic-related expenditures, diminished to negligible levels in the subsequent years.
-
-Prevalence of Small Contracts: The majority of government contracts are relatively small.
+Other notable contracts include multi-billion-dollar deals with the F-35 Joint Program, General Atomics, CAE, and Airbus for fighter planes and aviation services.
+
+In contrast, other government departments have experienced significant reductions in procurement spending.
+For instance, the Public Health Agency of Canada's spending, which was robust in 2020 due to pandemic-related expenditures, diminished to negligible levels in recent years.
+
+The majority of government contracts are relatively small.
 Approximately 81.75% of all contracts are valued under \$100,000, with the most common contract size ranging between \$10,000 and \$20,000.
 This pattern indicates a governmental preference for smaller contracts, possibly for regulatory reasons.
-There may be a rule that if a contract is below \$20,000 then it requires less scrutiny to be passed, in effect encouraging department to award more smaller contracts.
+Though the sum of these contracts in minuscule compared to the rest of government procurements, \$7.83 Billion to \$112.2 Billion.
 
@@
 
-While the study provides valuable insights, several limitations must be acknowledged:
-
-Supplier Name Analysis: The methodology for standardizing supplier names, requires more rigorous techniques.
-The method used ensured that the top 10 suppliers are accurately aggregated.
-However, some contracts that these suppliers may have been missed.
+The methodology for standardizing supplier names, requires more rigorous techniques.
+The method used in this paper ensured that the top 10 suppliers are accurately aggregated.
 But beyond the top 10, the names of suppliers not well parsed out, with problems like variations of “incorporated” and that sort of thing, dominating the rest of the suppliers list.
-
-Model Limitations: The linear regression model demonstrated weaknesses, primarily due to its simplicity and the limitations of the dataset.
-The assumption of a linear relationship between the predictors (Baseline Days and Duration Days) and contract amounts may not capture the complexities of procurement spending, which can be influenced by many things such as; policy changes, economic conditions, and sector-specific dynamics.
-
-Methodological Constraints: The model did not account for external variables such as inflation, exchange rates, or changes in government procurement policies, which could significantly influence contract amounts.
+This method is also not robust to new additions to the supplier list.
+
+The linear regression model demonstrated weaknesses, primarily due to its simplicity and the limitations of the dataset.
+The assumption of a linear relationship between the predictors and contract amounts may not capture the complexities of procurement spending, which can be influenced by many things such as; policy changes, economic conditions, and sector-specific dynamics.
+
+The model did not account for external variables such as inflation, exchange rates, or changes in government procurement policies, which could significantly influence contract amounts.
 Additionally, treating all contracts as a homogeneous group overlooks the diversity in contract types and scopes.
@@
 
-To address these weaknesses and build upon the findings, the following steps are recommended:
-
-Enhance Data Analysis Techniques: 
-Supplier Name Processing: Implement advanced text processing techniques, such as fuzzy matching, to accurately standardize supplier names and improve data reliability. ADD MORE
-
-Contract Description Analysis: Utilize text mining and content analysis methods on contract descriptions to uncover patterns in procurement objectives and priorities. ADD MORE
-
-Expand Data Collection:
-Comprehensive Spending Data: Include other forms of government spending, such as in-house expenditures on salaries and operational costs, as this would contextualize the procurement budget in the larger scheme of federal spending overall.
+To address these weaknesses and build upon this paper I want to implement more complicated text processing techniques, such as fuzzy matching, to accurately standardize supplier names and improve data reliability.
+Utilize text mining and content analysis methods on contract descriptions to uncover patterns in procurement objectives and priorities.
+
+Include other forms of government spending, such as in-house expenditures on salaries and operational costs, as this would contextualize the procurement budget in the larger scheme of federal spending overall.
 How much of the total federal budget goes into procurements?
 Are procurements more cost effective than in house spending?
-
-Investigate Specific Trends:
-Public Health Agency Spending: Examine the factors contributing to the decline in the Public Health Agency of Canada's procurement spending.
-
-Proliferation of Small Contracts: Analyze the rationale behind the high frequency of small contracts, exploring whether this is a strategic decision to support smaller businesses or a reflection of procurement policy regulations.
-
-Refine Modeling Approaches:
-Advanced Statistical Models: Explore more complex modeling techniques, such as nonlinear regression, time series analysis, or machine learning models, to capture the intricate relationships between variables.
-
-Incorporate Additional Variables: Include economic indicators, policy changes, and other relevant factors to improve the explanatory power of the models.
-
-By undertaking these steps, future research can provide a more comprehensive and nuanced understanding of Canadian federal procurement spending.
+Examine the factors contributing to the decline in the Public Health Agency of Canada's procurement spending.
+
+Explore more complex modeling techniques, such as nonlinear regression, time series analysis, or machine learning models, to capture the intricate relationships between variables.
+Include economic indicators, policy changes, and other relevant factors to improve the explanatory power of the models.
+
+By undertaking these steps, future research can provide a more complete understanding of Canadian federal procurement spending.
 This further analysis could inform policy decisions, promote transparency, and ensure that government resources are allocated effectively and efficiently.
@@
 
-## Web Scraping
+## Web Scraping {#sec-appendix-web-scraping}
 
@@
 
-## Data Cleaning
-
-I began by converting Amount into a numeric type and removing all the \$ signs.
-Then I converted all the date columns, Award Date, Start Date, End Date into date types.
-There was one date whose start year was was in 2204, this was clearly a typo so I set it to 2024.
+## Data Cleaning {#sec-appendix-data-cleaning}
+
+I began by converting Amount into a `numeric type` and removing all the \$ signs.
+Then I converted all the date columns, Award Date, Start Date, End Date into `date type`.
+There was one contract whose start year was was in 2204, this was clearly a typo so I set it to 2024.
 There were a significant number of contracts whose Start Date was the same as their End Date.
@@
 There were an number of contracts which had ridiculous start dates (i.e.
-2156) so I removed any contracts beyond 31st of December 2099, which may still be to late.
+2156) so I removed any contracts beyond 31st of December 2099, which may still be too late.
 Finally I removed any duplicate rows, I believe that the IJF database has a significant number of duplicate rows.
-It is also possible that rows were duplicated in the web scraping process as I noticed some strange behavior on IJF's website.
+Or it is also possible that rows were duplicated in the web scraping process as I noticed some strange behavior on the IJF website.
 If you go to the last page take note of the top companies there, and then the hit Previous page button, then the Next page button you will notice that its a difference set of companies.
@@
 
-## Proccessing Supplier
+## Proccessing Supplier {#sec-appendix-supplier}
 
 To begin I made every suppliers name lowercase.
-Then I divided each word in a suppliers name and stored them in a list in the column processed_supplier.
-Then I made a list of 'banned' words, which are word that I will remove from each suppliers name.
-These banned words are thing like: incorporated, llc, the, service, basically any useless non-identifying words.
+Then I divided each word in a suppliers name and stored each word in a list in the column processed_supplier.
+Then I made a list of 'banned' words, which are words that I will remove from each suppliers name.
+These banned words are thing like: "incorporated", "llc", "the", "service", basically any useless non-identifying words.
 But sometimes a company name is two banned words so they are left with and empty list in processed_supplier.
-Most notably this happened with Vancouver Shipyards, to fix this whenever the program found a list with vancouver and shipyards it would concatenate these words into one string.
+Most notably this happened with Vancouver Shipyards, to fix this whenever the program found a list with "vancouver" and "shipyards" it would concatenate these words into one string.
 Then I loop through the the lists in processed_suppliers and removed any banned words.
 Finally I created a dictionary that would go through each processed_suppliers and pick out the individual words and set that as a key.
-Then I would loop through an if a key was present in a processed_suppliers list it would add the Amount of that contract to the value which would be a running sum for the suppliers with that key.
+Then I would loop through and if a Key was present in a processed_suppliers list it would add the Amount of that contract to the Value which would be a running sum for the suppliers with that key.
 
@@
 Then I would add this key to the banned words list and re-run the program.
-Eventually I had banned enough non-descriptive words that the top ten keys were aggregating only one supplier each and could be related to a specific company.
+Eventually I had banned enough non-descriptive words that the top ten keys were aggregating with only one supplier each and could be related to a specific company.
 
@@
 If I wanted to get the top 15 or top 100 companies it would be very difficult to root out all of the confounding words.
-I also had to be vigilant when added a word to the banned words list because I could rule out an entire company, like what happened with Vancouver Shipyards.
-Then I had to create a work around that allowed be to remove the words "Vancouver" and "Shipyards", while being able to keep "Vancouver Shipyards".
+I also had to be vigilant when added a word to the banned words list, because I could rule out an entire company, like what happened with Vancouver Shipyards.
+Then I had to create a work around that allowed me to remove the words "Vancouver" and "Shipyards", while being able to keep "Vancouver Shipyards".
 Essentially the problem with this method is that it is matching companies up if the share a single word in common, and that if fine for most companies (after getting rid of the banned words), but some company names are made up of generic words like Vancouver Shipyards.
@@
 This problem would only get worse the further down the list you go and at some point it may be impossible to continue with this method.
-To make matters worse when adding new data you may already be banning all the words in their names and would have no idea.
-Because at least when you are looking to add a banned word you can check to see if that word is the last word for a lot of companies, but you won't get that opportunity for novel companies.
-In short I am sure the top 10 companies that I am displaying truly are the top 10, however using this method I cannot say what the top 11 or more companies are and if new data is introduced then all bets are off.
+To make matters worse when adding new data you may have already be banning all the words in their names and would have no idea.
+Because at least when you are looking to add a banned word you can check to see if that word is the last word for a lot of companies, but you won't get that opportunity for newly added companies.
+In short I am sure that the top 10 companies that I am displaying truly are the top 10.
+However, using this method I cannot say what the top 11 or more companies are, and if new data is introduced then all bets are off.
 
@@
 Some good candidates that I have looked into for better supplier name parsing would be: Fuzzy Matching and String Distance Algorithms, Natural Language Processing (Tokenization, Named Entity Recognition (NER)).
-It would be nice to apply these algorithms to the contract descriptions as well, in that case you could gain further insights into what the government is buying.
-
-# Contract Surveys {#sec-surveys}
+It would be nice to apply these algorithms to the contract descriptions as well, in that case you could gain a bit more knowledge into what the government is buying.
+
+# Contract Surveys {#sec-survey}
+
+A survey targeting sellers would serve multiple purposes.
+It would aim to gather detailed information into the nature of the contracts, understand how these contracts assist governmental functions, and identify areas for improvement in the procurement process.
+The survey would be designed to capture both quantitative and qualitative data.
+
+- The survey would begin by asking sellers to elaborate on the specific goods or services provided under the contract. Questions would include:
+
+Contract Description: "Please provide a detailed description of the goods or services supplied."
+Objectives Alignment: "How do your offerings align with the strategic objectives of the government department you are contracted with?"
+
+Needs Addressed: "What specific governmental needs or challenges does your product/service address?"
+These questions help in understanding not just what is being supplied, but also the relevance and importance of these goods or services to government operations.
+
+- To assess how the contract is helping the Canadian government, the survey would include questions aimed at measuring outcomes and benefits:
+
+Performance Metrics: "What key performance indicators (KPIs) are used to measure the success of your contract deliverables?"
+Quantifiable Benefits: "Can you quantify any cost savings, efficiency gains, or service improvements resulting from your contract?"
+Case Studies: "Please provide examples or case studies that highlight the positive impact of your work on government operations."
+By gathering this information, the government can evaluate the tangible benefits derived from the contracts.
+
+- To measure Collaboration and Relationship with Government Entities, the survey would ask:
+
+Communication Effectiveness: "How would you rate the communication channels with the government during the contract period?"
+Support and Resources: "Did you receive adequate support and resources from the government to fulfill your contractual obligations?"
+Feedback Mechanisms: "Were there established mechanisms for providing and receiving feedback throughout the contract duration?"
+These questions aim to identify strengths and weaknesses in collaboration, which can inform future improvements.
+
+- Procurement Process Feedback
+
+To refine the procurement process, the survey would solicit sellers' experiences and suggestions:
+
+Clarity of Guidelines: "Were the procurement guidelines and requirements clear and understandable?"
+Bidding Process: "Did you find the bidding process fair and transparent?"
+Administrative Efficiency: "How efficient was the contract administration from award to execution?"
+Improvement Suggestions: "What recommendations do you have to improve the procurement process?"
+Feedback in this area is vital for enhancing fairness, transparency, and efficiency in future procurements.
+
+- Challenges and Obstacles
+
+Identifying potential barriers faced by sellers helps in creating solutions:
+
+Obstacles Encountered: "Did you encounter any challenges or obstacles during the contract? Please specify."
+Resolution Strategies: "How were these challenges addressed, and what support did you receive in resolving them?"
+Systemic Issues: "Are there systemic issues in the procurement process that need attention?"
+Understanding challenges at the ground level enables the government to address systemic issues that may affect multiple suppliers.
+
+- Innovation and Future Opportunities
+
+Encouraging innovation and forward-thinking, the survey would include:
+
+Innovation Implementation: "Have you introduced any innovative solutions or technologies through your contract?"
+Future Collaborations: "In what ways can you contribute to future government projects or initiatives?"
+Market Trends: "Are there emerging trends or developments in your industry that the government should be aware of?"
+This section helps the government stay ahead by leveraging supplier expertise.
+
+- Survey Design Considerations
+
+Confidentiality Assurance: Ensuring respondents that their feedback will be kept confidential encourages honesty.
+Accessibility: Providing the survey in both English and French, and ensuring it's accessible on various devices, increases participation rates.
+Question Clarity: Questions should be clear, concise, and free of jargon to prevent misunderstandings.
+Balanced Question Types: A mix of open-ended questions for detailed feedback and closed-ended questions for quantitative analysis is ideal.
+Pilot Testing: Before full deployment, pilot testing the survey can help identify and rectify any issues.
+Conclusion
+
+By encompassing these elements, the idealized survey would not only collect data on how contracts are benefiting the Canadian government but also strengthen the overall procurement framework.
+It would provide a platform for sellers to share their experiences, challenges, and suggestions, fostering a collaborative environment that promotes efficiency, transparency, and mutual benefit.
 
